### 2015-08-02 paper/spares_interaction.tex
adding paper

\documentclass[twoside,11pt]{article}

\usepackage{array}
\usepackage{clrscode3e}
\usepackage{amsmath}

\setlength{\parindent}{0em}
\setlength{\parskip}{1em}

\begin{document}

\title{Efficient Calculation of Interaction Features on Sparse Matrices}
\author{Andrew Nystrom}
\date{}

\maketitle

\begin{abstract}%   <- trailing '%' for backward compatibility of .sty file
FILL THIS IN
\end{abstract}

\section{Introduction}

Introduction
Interaction features are a way of capturing correlations between features in a machine 
learning setting. They consist of products of combinations of features. This work 
describes a method for efficiently calculating second degree interaction features on a 
sparse matrix, and could be generalized to higher orders.

Consider the following matrix:

\[ \left( \begin{array}{cccc}
3 & 0 & 0 & 3 \\
0 & 4 & 2 & 0 \\
1 & 2 & 0 & 3 \end{array} \right)\]

If each row is an instance vector, then the interaction feature matrix is

\[ \left( \begin{array}{cccccc}
0 & 0 & 9 & 0 & 0 & 0 \\
0 & 0 & 0 & 8 & 0 & 0 \\
2 & 0 & 3 & 0 & 6 & 0 \end{array} \right)\]

Note that there are $\binom{D}{2} = \frac{D^2-D}{2}$ columns in the interaction matrix, which is D choose 2, 
since we generate products of all combinations of 2 features in the original matrix. 
Which column corresponds with which product pair is not important so long as it’s 
consistent. In this example, each column in the interaction matrix corresponded to the 
following product pairs: $(1, 2),  (1, 3),  (1, 4),  (2, 3), ( 2, 4), (3, 4)$.

Notice that the interaction matrix contains many zero entries. This is of course because 
the original matrix contained zero entries, so the interaction features, which are 
products of pairs of features, contain many zeros. This means that the only products that 
need to actually be calculated or those for which both features in the combination are 
nonzero. If the original matrix is sparse and represented in a sparse matrix format 
(e.g. compressed sparse row), a list of nonzero column indices are stored for each row 
and are easily retrievable in $O(1)$ time. Interaction features can be generated from this 
list via the following method.

\section{Approach}
Let the list of nonzero columns for a given row be $N_{zc}$. The nonzero second degree 
interaction features are simply the products of all combinations of two elements whose 
columns are in $N_{zc}$. However, to do this efficiently, a consistent mapping from the column 
index pairs of $N_{zc}$ into the columns of the interaction matrix is needed. The mapping is 
from the space (a, b) where a,b are in $1,2,..., D$ onto the space $1,2,..., \frac{D^2-D}{2}$. This 
is isomorphic to mapping the coordinates of the upper triangle of a matrix onto a flat 
list. The following is a proof by construction for such a mapping.

INSERT JOHN'S PROOF HERE

With this mapping, an algorithm for generating second degree interaction features on a 
matrix $A$ can be formulated as follows:

\begin{codebox}
\Procname{$\proc{Sparse Interaction}(A)$}
\li $\func{map}(b, a) = \frac{2Da-a^2+2b-3a-2}{2}$
\li $B$ $\gets$ Compressed Sparse Row Matrix of size $N \times \frac{D^2-D}{2}$
\li \For $\id{row} in A$
\li     $N_{zc} \gets nonzero columns of row$
\li     \For each combination $a, b$ of elements of $N_{zc}$
\li         $k \gets \func{map}(b, a)$
\li         $i \gets index of \id{row}$
\li         $B[i, k] \gets \id{row}[a] * \id{row}[b]$
        \End
    \End
\end{codebox}

\section{Complexity Analysis}
Assume that A is a matrix with sparsity $0 < d < 1$, $N$ rows, and $D$ columns. Finding 
interaction features with the proposed algorithm has time and space complexity 
$\func{O}(d N D^2)$, 
whereas a naive approach of using non-sparse matrices and multiplying all column 
combinations has time and space complexity $\func{O}(N D^2)$. The algorithm is therefore an 
improvement by a factor of the density factor of $A$.

This can represent a large gain in speed and time. For example, the 20 Newsgroups dataset 
has density $d$ of 0.12 when its unigrams are represented in a vector space model. This 
means the proposed approach would take less than $\frac{1}{8}$ time and memory.

The real benefit of this method is revealed when the average complexity is analysed. The 
number of interactions calculated for a given row are $\binom{|N_{zc}|}{2}$. If the matrix has 
density $d$, then on average, $N_{zc} = D d$, so the number of interaction features 
calculated in total is

\begin{align*}
N \binom{d D}{2} &= \frac{N (Dd)!}{2!(Dd-2)!}\\
    \\
    &= \frac{(D^2d^2-Dd)}{2}
\end{align*}

This means that the average complexity decreases quadratically with the density.

\section{Future Work}
The approach for generating second degree interaction features required a mapping from 
combinations of two to the space $1,2,...,\frac{D^2-D}{2}$, which is isomorphic to a mapping from 
the indices of an upper triangular matrix to the indices of a flat list of the same size. 
To generate third degree interaction features, a mapping from combinations of three 
$(a,b,c)$ to the space $1,2,...\frac{D^3-3D^2+2D}{6}$ (which is $\binom{D}{3}$), or the upper $3$-simplex of a tensor to a flat 
list of the same size $\frac{D^3-3D^2+2D}{6}$ would be required. In general, for interaction 
features of degree k, the upper $k$-simplex of a $k$-dimensional tensor must be mapped to the 
space $1,2,...\frac{D!}{k!(D-k)!}$. A similar approach for finding these mappings could be taken 
as the one used here for $k=2$.

Motivation for deriving mapping functions for higher orders
of interaction features is that the average complexity of generating degree $k$ interaction
features is $N \binom{Dd}{k}$, which decreases polynomially with respect to k compared to
generating the features naively.

\vskip 0.2in
\bibliography{sample}

\end{document}

### 2015-08-07 paper/spares_interaction.tex
fixed indentation in algorithm
single sided
.
clarifying what interaction features are

@@
-\documentclass[twoside,11pt]{article}
+\documentclass[11pt]{article}
 
@@
 Interaction features are a way of capturing correlations between features in a machine 
-learning setting. They consist of products of combinations of features. This work 
-describes a method for efficiently calculating second degree interaction features on a 
-sparse matrix, and could be generalized to higher orders.
+learning setting. A feature vector $\vec{x}$ of dimensionality $D$ has second degree interaction features 
+$\{x_a \cdot x_b : a, b \in \{0,1,..., D-1\} \land a < b\}$, 
+so a $D$ dimensional vector has $\binom{D}{2} = \frac{D^2-D}{2}$ second degree interaction features. A naive
+approach to calculating these features is to simply iterate through the combinations of the column indices.
+For a sparse vector, many of the resulting interaction features would be zero, and could therefore be ignored.
+This work describes a method to efficiently calculate second degree interaction features for a sparse matrix 
+that has time and space complexities that decrease quadratically with the density of the input matrix with respect to the naive approach.
 
-Consider the following matrix:
+%Consider the following matrix:
 
-\[ \left( \begin{array}{cccc}
-3 & 0 & 0 & 3 \\
-0 & 4 & 2 & 0 \\
-1 & 2 & 0 & 3 \end{array} \right)\]
+%\[ \left( \begin{array}{cccc}
+%3 & 0 & 0 & 3 \\
+%0 & 4 & 2 & 0 \\
+%1 & 2 & 0 & 3 \end{array} \right)\]
 
-If each row is an instance vector, then the interaction feature matrix is
+%If each row is an instance vector, then the interaction feature matrix is
 
-\[ \left( \begin{array}{cccccc}
-0 & 0 & 9 & 0 & 0 & 0 \\
-0 & 0 & 0 & 8 & 0 & 0 \\
-2 & 0 & 3 & 0 & 6 & 0 \end{array} \right)\]
+%\[ \left( \begin{array}{cccccc}
+%0 & 0 & 9 & 0 & 0 & 0 \\
+%0 & 0 & 0 & 8 & 0 & 0 \\
+%2 & 0 & 3 & 0 & 6 & 0 \end{array} \right)\]
 
-Note that there are $\binom{D}{2} = \frac{D^2-D}{2}$ columns in the interaction matrix, which is D choose 2, 
-since we generate products of all combinations of 2 features in the original matrix. 
-Which column corresponds with which product pair is not important so long as it’s 
-consistent. In this example, each column in the interaction matrix corresponded to the 
-following product pairs: $(1, 2),  (1, 3),  (1, 4),  (2, 3), ( 2, 4), (3, 4)$.
+%Note that there are $\binom{D}{2} = \frac{D^2-D}{2}$ columns in the interaction matrix, which is D choose 2, 
+%since we generate products of all combinations of 2 features in the original matrix. 
+%Which column corresponds with which product pair is not important so long as it’s 
+%consistent. In this example, each column in the interaction matrix corresponded to the 
+%following product pairs: $(1, 2),  (1, 3),  (1, 4),  (2, 3), ( 2, 4), (3, 4)$.
 
-Notice that the interaction matrix contains many zero entries. This is of course because 
-the original matrix contained zero entries, so the interaction features, which are 
-products of pairs of features, contain many zeros. This means that the only products that 
-need to actually be calculated or those for which both features in the combination are 
-nonzero. If the original matrix is sparse and represented in a sparse matrix format 
-(e.g. compressed sparse row), a list of nonzero column indices are stored for each row 
-and are easily retrievable in $O(1)$ time. Interaction features can be generated from this 
-list via the following method.
+%Notice that the interaction matrix contains many zero entries. This is of course because 
+%the original matrix contained zero entries, so the interaction features, which are 
+%products of pairs of features, contain many zeros. This means that the only products that 
+%need to actually be calculated or those for which both features in the combination are 
+%nonzero. If the original matrix is sparse and represented in a sparse matrix format 
+%(e.g. compressed sparse row), a list of nonzero column indices are stored for each row 
+%and are easily retrievable in $O(1)$ time. Interaction features can be generated from this 
+%list via the following method.
 
 \section{Approach}
-Let the list of nonzero columns for a given row be $N_{zc}$. The nonzero second degree 
+Let the list of nonzero columns for a given row $\vec{x}$ be denoted by $N_{zc}$. The nonzero second degree 
 interaction features are simply the products of all combinations of two elements whose 
-columns are in $N_{zc}$. However, to do this efficiently, a consistent mapping from the column 
+columns are in $N_{zc}$. However, to properly place an interaction feature into the correct column, a mapping from the column 
 index pairs of $N_{zc}$ into the columns of the interaction matrix is needed. The mapping is 
@@
 \Procname{$\proc{Sparse Interaction}(A)$}
-\li $\func{map}(b, a) = \frac{2Da-a^2+2b-3a-2}{2}$
-\li $B$ $\gets$ Compressed Sparse Row Matrix of size $N \times \frac{D^2-D}{2}$
-\li \For $\id{row} in A$
-\li     $N_{zc} \gets nonzero columns of row$
-\li     \For each combination $a, b$ of elements of $N_{zc}$
-\li         $k \gets \func{map}(b, a)$
-\li         $i \gets index of \id{row}$
-\li         $B[i, k] \gets \id{row}[a] * \id{row}[b]$
+    \zi $\func{map}(b, a) = \frac{2Da-a^2+2b-3a-2}{2}$
+    \zi $B$ $\gets$ Compressed Sparse Row Matrix of size $N \times \frac{D^2-D}{2}$
+    \zi \For $\id{row}$ in $A$ \Do
+    \zi     $N_{zc} \gets$ nonzero columns of $row$
+    \zi     \For each combination $a, b$ of elements of $N_{zc}$ \Do
+    \zi         $k \gets \func{map}(b, a)$
+    \zi         $i \gets$ index of $\id{row}$
+    \zi         $B[i, k] \gets \id{row}[a] \cdot \id{row}[b]$
             \End
@@
 \begin{align*}
-N \binom{d D}{2} &= \frac{N (Dd)!}{2!(Dd-2)!}\\
+N \binom{d D}{2} &= N \frac{(Dd)!}{2!(Dd-2)!}\\
     \\
-    &= \frac{(D^2d^2-Dd)}{2}
+    &= N \frac{(D^2d^2-Dd)}{2}
 \end{align*}

### 2015-08-24 paper/spares_interaction.tex
fixed algorithm and added package files

@@
 \usepackage{amsmath}
+\usepackage{kbordermatrix}
 
@@
 learning setting. A feature vector $\vec{x}$ of dimensionality $D$ has second degree interaction features 
-$\{x_a \cdot x_b : a, b \in \{0,1,..., D-1\} \land a < b\}$, 
+$\{x_i \cdot x_j : i, j \in \{0,1,..., D-1\} \land i < j\}$, 
 so a $D$ dimensional vector has $\binom{D}{2} = \frac{D^2-D}{2}$ second degree interaction features. A naive
@@
 that has time and space complexities that decrease quadratically with the density of the input matrix with respect to the naive approach.
-
-%Consider the following matrix:
-
-%\[ \left( \begin{array}{cccc}
-%3 & 0 & 0 & 3 \\
-%0 & 4 & 2 & 0 \\
-%1 & 2 & 0 & 3 \end{array} \right)\]
-
-%If each row is an instance vector, then the interaction feature matrix is
-
-%\[ \left( \begin{array}{cccccc}
-%0 & 0 & 9 & 0 & 0 & 0 \\
-%0 & 0 & 0 & 8 & 0 & 0 \\
-%2 & 0 & 3 & 0 & 6 & 0 \end{array} \right)\]
-
-%Note that there are $\binom{D}{2} = \frac{D^2-D}{2}$ columns in the interaction matrix, which is D choose 2, 
-%since we generate products of all combinations of 2 features in the original matrix. 
-%Which column corresponds with which product pair is not important so long as it’s 
-%consistent. In this example, each column in the interaction matrix corresponded to the 
-%following product pairs: $(1, 2),  (1, 3),  (1, 4),  (2, 3), ( 2, 4), (3, 4)$.
-
-%Notice that the interaction matrix contains many zero entries. This is of course because 
-%the original matrix contained zero entries, so the interaction features, which are 
-%products of pairs of features, contain many zeros. This means that the only products that 
-%need to actually be calculated or those for which both features in the combination are 
-%nonzero. If the original matrix is sparse and represented in a sparse matrix format 
-%(e.g. compressed sparse row), a list of nonzero column indices are stored for each row 
-%and are easily retrievable in $O(1)$ time. Interaction features can be generated from this 
-%list via the following method.
 
@@
 \Procname{$\proc{Sparse Interaction}(A)$}
-    \zi $\func{map}(b, a) = \frac{2Da-a^2+2b-3a-2}{2}$
+    \zi $\func{map}(a, b) = \frac{2Da-a^2+2b-3a-2}{2}$
+    \zi $N \gets$ row count of $A$
+    \zi $D \gets$ column count of $A$
     \zi $B$ $\gets$ Compressed Sparse Row Matrix of size $N \times \frac{D^2-D}{2}$
@@
     \zi     $N_{zc} \gets$ nonzero columns of $row$
-    \zi     \For each combination $a, b$ of elements of $N_{zc}$ \Do
-    \zi         $k \gets \func{map}(b, a)$
-    \zi         $i \gets$ index of $\id{row}$
-    \zi         $B[i, k] \gets \id{row}[a] \cdot \id{row}[b]$
+    \zi     \For $i \gets 0 \To |N_{zc}|-2$ \Do
+    \zi         \For $j \gets i+1 \To |N_{zc}|-1$ \Do
+    \zi             $k \gets \func{map}(i, j)$
+    \zi             $r \gets$ index of $\id{row}$
+    \zi             $B[r, k] \gets \id{row}[i] \cdot \id{row}[j]$
+                \End
             \End

### 2015-08-26 sparse_interaction/sparse_interaction.pyx
can now do polynomial features

@@ -10,34 +10,45 @@ from scipy.sparse import hstack, coo_matrix
 from itertools import combinations
 
 from libc.math cimport pow
-from cpython.mem cimport PyMem_Malloc, PyMem_Realloc, PyMem_Free
+from libcpp cimport bool
 
 ctypedef unsigned long long INDEXDTYPE
 
-class SparseInteractionFeatures(TransformerMixin, BaseEstimator):
+class SparsePolynomialFeatures(TransformerMixin, BaseEstimator):
     """
     Compute interaction features on a sparse (CSR) matrix while keeping its format sparse.
     """
+    
+    def __init__(self, interaction_only=False):
+        self.interaction_only = interaction_only
     
     def fit(self, X, Y=None, **fit_params):
         return self
     
     @cython.cdivision(True)
-    def transform(self, X, **transform_params):
+    def transform(self, X, int interaction_only=False, **transform_params):
                                                         
-        cdef INDEXDTYPE a, b, x, y, row_num, interaction_col, n_cols_X, interaction_dim, nnz, num_interactions = 0
-        
-        #If the dimensionality of the input is less than 363 we can use 32 bits
+        cdef INDEXDTYPE a, b, x, y, row_num, interaction_col, n_cols_X, output_dim, nnz, num_interactions = 0, min_col_count
+        cdef int combo_j_offset, combo_i_offset
                  
         n_cols_X = X.shape[1]
-        interaction_dim = <INDEXDTYPE>((n_cols_X**2-n_cols_X) / 2) #The size of the combination output space
-        
-        #http://math.stackexchange.com/questions/646117/how-to-find-a-function-mapping-matrix-indices
-        #Thanks, Professor John Hughes of Brown University!
+        if interaction_only == 1:
+            combo_j_offset = 1
+            combo_i_offset = -1
+            output_dim = <INDEXDTYPE>((n_cols_X**2-n_cols_X) / 2) #upper triangular
+            min_col_count = 2
+            for nnz in np.bincount(X.nonzero()[0]):
+                num_interactions += <INDEXDTYPE>((pow(nnz,2)-nnz)/2)
+        else:
+            combo_j_offset = 0
+            combo_i_offset = 0
+            min_col_count = 1
+            output_dim = <INDEXDTYPE>((n_cols_X**2+n_cols_X) / 2) #diagonal counted
+            for nnz in np.bincount(X.nonzero()[0]):
+                num_interactions += <INDEXDTYPE>((pow(nnz,2)+nnz)/2)
         
[…]

### 2015-09-19 paper/paper2.tex
redoing the paper

\documentclass[11pt]{article}

\usepackage{array}
\usepackage{clrscode3e}
\usepackage{amsmath}
\usepackage{kbordermatrix}

\setlength{\parindent}{0em}
\setlength{\parskip}{1em}

\begin{document}

\title{Efficient Calculation of Polynomial Features on Sparse Matrices}
\author{Andrew Nystrom}
\date{}

\maketitle

\begin{abstract}%   <- trailing '%' for backward compatibility of .sty file
FILL THIS IN
\end{abstract}

\section{Introduction}
When performing a modeling task, one is often faced with a non-linearly separable problem,
or is attempting to predict a target that is a non-linear combination of the variables at one's disposal.
There are several general classes of methods for dealing with these situations: kernel methods, 
high variance models, and input transformations.

Of the possible types of input transformations one
might apply, a widely used method is the polynomial transformation, which is defined as 
all products of all combinations with replacement of $k$ components of a $D$ dimensional row vector $\vec{x}$ (assuming row-wise instance representation), where $k$ is the degree of the polynomial.
In the case of $k=2$, the following set of polynomial features are augmented to each $\vec{x}_i$:

\begin{equation}
\{x_a \cdot x_b : a, b \in \{0,1,..., D-1\} \land a \le b\}
\end{equation}

which results in $\binom{D+1}{D-1}$ additional features, and so the generation of 
these features has a time complexity of $O(D^2)$. In the general case of degree $k$, there 
are $\binom{D+k-1}{D-1}$ polynomial features and the time complexity is $O(D^k)$.

If the vectors of a feature matrix are sparse, many of these products of features will be
zero, and are therefore unnecessary to calculate if the matrix is stored in a sparse matrix 
data structure. If the density (the fraction of nonzero elements) of the matrix is $d$,
the number of nonzero polynomial features is

\begin{equation}
\binom{dD+2-1}{dD-1} = \frac{d^2D^2+dD}{2}
\end{equation}

and if only they were calculated, the time complexity of generating polynomial features
would be $O(d^2D^2)$, quadratically lower than calculating them via brute force.

The question then is how to devise an algorithm to calculate polynomial features for a vector $\vec{x}$ so that
only the nonzero elements of $\vec{x}$ are considered. Two main components are needed: (1) the ability to
quickly access the nonzero elements of $\vec{x}$ and (2) The ability to know which polynomial feature column
the product of features with indices $a$ and $b$ should be stored in.

Motivation for the need for (2) is not obvious. To make this need more clear, consider the
algorithm for the brute force calculation of polynomial features:

\begin{codebox}
\Procname{$\proc{Dense Polynomial Features}(\vec{x})$}
    \zi $D \gets |\vec{x}|$
    \zi $b$ $\gets$ vector of size $\frac{D^2+D}{2}$
    \zi $k \gets 0$
    \zi \For $\i \gets 0 \To D$ \Do
    \zi     \For $\j \gets i \To D$ \Do
    \zi         $\vec{b}[k] \gets \vec{x}[i] \cdot \vec{x}[j]$
    \zi         $k \gets k + 1$
            \End
        \End
    \End
    return $\vec{x} | \vec{b}$
\end{codebox}

Notice that the way the column the polynomial feature between features $i$ and $j$ can be determined
by a simple counter $k$ that is incremented each pass through the inner loop. This cannot be
done if we only calculate products between nonzero elements of $\vec{x}$, because all that
will be known is the nonzero element indices - many iterations will be skipped (hence the improved computational complexity).

What is therefore needed is a mapping between column index pairs of $\vec{x}$ and the space ${0, 1, ..., \binom{D+1}{D-1}-1}$
(the output size of second degree polynomial features when $D = |\vec{x}|$. The ordering of the mapping is irrelevant so long as
it its input $ I = {(a, b) : a, b \in \{0,1,..., D-1\} \land a \le b}$ is bijective with its output, ${0, 1, ..., \binom{D+1}{D-1}-1}$.

To construct this mapping, consider a matrix of size $D \times D$ where cell $(a,b)$ represents the
polynomial feature between indices $a$ and $b$ for $a,b \in I$ for vector $\vec{x}$.

\renewcommand{\kbldelim}{(}% Left delimiter
\renewcommand{\kbrdelim}{)}% Right delimiter
\[
  \kbordermatrix{
            & x_0 & x_1   & \cdot   & \cdot   &  \cdot  &  x_{D-1} \\
    x_0     & 0   & 1     & \cdot   &  \cdot  &  \cdot  &  D-1     \\
    x_1     & -   & D     & \cdot   &  \cdot  &  \cdot  &  2D-2    \\
     \cdot     & \cdot  & \cdot    &  \cdot  &      &      &  \cdot       \\
     \cdot     & \cdot  &       &  \cdot  &  \cdot  &      &  \cdot       \\
     \cdot     & \cdot  &       &      &  \cdot  &  \cdot  &  \cdot       \\
    x_{D-1} & -   & \cdot    &  \cdot  & \cdot   & -    &  \frac{D(D+1)}{2}-1\\
  }
\]

This is a rather natural sequencial ordering on both the input and output spaces. As can 
be seen, the required mapping can e viewed as mapping the upper triangular matrix (including 
the diagonal) to $0, 1, ..., \frac{D(D+1)}{2}-1$. The construction of this mapping will now
be given.

\vskip 0.2in
\bibliography{sample}

\end{document}

### 2015-09-21 paper/paper2.tex
need john's proof

@@
 \usepackage{kbordermatrix}
+\usepackage{mathtools}
+\usepackage{cite}
+\usepackage{hyperref}
+\usepackage{rotating}
+\usepackage{svg}
+
+\newcommand{\defeq}{\vcentcolon=}
+\newcommand{\eqdef}{=\vcentcolon}
 
@@
 
-\title{Efficient Calculation of Polynomial Features on Sparse Matrices}
-\author{Andrew Nystrom}
+\title{Efficient Calculation of Polynomial and Interaction Features on Sparse Matrices}
+\author{
+  Andrew Nystrom\\
+  \texttt{awnystrom@gmail.com}
+  \and
+  John Hughes\\
+  \texttt{jfh@cs.brown.edu}
+}
 \date{}
@@
 \begin{abstract}%   <- trailing '%' for backward compatibility of .sty file
-FILL THIS IN
+We give an algorithm to efficiently calculate second degree polynomial or interaction 
+that has complexity quadratically (with respect to the density of the matrix) better than 
+the naive approach and allows for the matrix to remain in sparse form. We apply the method
+to several real datasets and give both analytical and empirical comparisons with the
+naive approach.
 \end{abstract}
@@
 
-\begin{equation}
+\begin{equation*}
 \{x_a \cdot x_b : a, b \in \{0,1,..., D-1\} \land a \le b\}
-\end{equation}
+\end{equation*}
 
-which results in $\binom{D+1}{D-1}$ additional features, and so the generation of 
-these features has a time complexity of $O(D^2)$. In the general case of degree $k$, there 
-are $\binom{D+k-1}{D-1}$ polynomial features and the time complexity is $O(D^k)$.
+which results in $\binom{D+1}{D-1} = \frac{D^2+D}{2}$ additional features, and so the generation of 
+these features for a matrix of size $N \times D$ has a time complexity of $O(ND^2)$. In the general case of degree $k$, there 
+are $\binom{D+k-1}{D-1}$ polynomial features and the time complexity is $O(ND^k)$.
 
+To help visualize what second degree polynomial features are capable of in a machine learning paradigm,
+we trained a linear classifier (logistic regression) on four different datasets and show their decision
+boundaries (figure \eqref{fig:boundaries} on page \pageref{fig:boundaries}) with and without polynomial features, along with their accuracy on a holdout set. We compare
+these to an RBF SVM via scikit-learn\footnote{\url{http://scikit-learn.org/stable/modules/generated/sklearn.svm.SVC.html}} 
+and gradient boosted decision trees via the xgboost package\footnote{\url{https://github.com/dmlc/xgboost/}}.
+
+\begin{sidewaysfigure}
+    \centering
+    \includegraphics[scale=0.5]{classifier_boundaries.png}
+    \caption{Decision boundaries for four classifier types on four different datasets. The darker the color, the more certain the decision.
+             The plain dataset is shown in the first column. Each other column is a type of classifier applied to that dataset. The classifier
+             types are logistic regression, logistic regression with second degree polynomial features, RBF SVM, and gradient boosting with decision trees.
+             The accuracy of each model is shown in the bottom right corner of that model's subplot.}
+    \label{fig:boundaries}
+\end{sidewaysfigure}
+
+With polynomial features added, a simple linear classifier is capable of drawing decision boundaries
+that take on the form of arbitrary parabolas, hyperbolas, and elipses. Beyond classification, polynomial regression \cite{theoryoflinearmodels} is a staple regression method.
+While polynomial features are a well known and used concept in machine learning (e.g.
+\cite{pavlidis2002learning, konidaris2009skill, wiesler2009investigations}) and work has been done to improve their utilization (e.g.
+\cite{pevckov2008minimal, huang2010predicting}), this is the first work that we know of that exploits matrix sparcity
+to improve the speed with which they are calculated and allows them to be calculated for a sparse matrix.
+
+\section{Sparse Polynomial Features} \label{sec:sparse}
 If the vectors of a feature matrix are sparse, many of these products of features will be
 zero, and are therefore unnecessary to calculate if the matrix is stored in a sparse matrix 
-data structure. If the density (the fraction of nonzero elements) of the matrix is $d$,
+data structure. If the density (the fraction of nonzero elements) of the matrix is $d$ where $0 \le d \le 1$,
 the number of nonzero polynomial features is
 
-\begin{equation}
+\begin{equation*}
 \binom{dD+2-1}{dD-1} = \frac{d^2D^2+dD}{2}
-\end{equation}
+\end{equation*}
 
 and if only they were calculated, the time complexity of generating polynomial features
-would be $O(d^2D^2)$, quadratically lower than calculating them via brute force.
+for a matrix of size $N \times D$ is $O(Nd^2D^2)$, quadratically lower than calculating them via brute force.
 
 The question then is how to devise an algorithm to calculate polynomial features for a vector $\vec{x}$ so that
-only the nonzero elements of $\vec{x}$ are considered. Two main components are needed: (1) the ability to
-quickly access the nonzero elements of $\vec{x}$ and (2) The ability to know which polynomial feature column
+only the nonzero elements of $\vec{x}$ are considered during the calculation process. Two main components are needed: (1) the ability to
+quickly access only the nonzero elements of $\vec{x}$ and (2) The ability to know which polynomial feature column
 the product of features with indices $a$ and $b$ should be stored in.
 
-Motivation for the need for (2) is not obvious. To make this need more clear, consider the
+The first of these necessary components is gotten simply by using the appropriate data structure
+to store the sparse matrix (e.g. storing it in compressed sparse row format). The second component 
+is not only less readily available, but its necessity is less obvious. To make this need more clear, consider the
 algorithm for the brute force calculation of polynomial features:
@@
 \begin{codebox}
-\Procname{$\proc{Dense Polynomial Features}(\vec{x})$}
-    \zi $D \gets |\vec{x}|$
-    \zi $b$ $\gets$ vector of size $\frac{D^2+D}{2}$
+\Procname{$\proc{Dense Polynomial Features}(A)$}
+    \zi $N \gets$ row count of $A$
+    \zi $D \gets$ column count of $A$
+    \zi $B$ $\gets$ Matrix of size $N \times \frac{D^2+D}{2}$
+    \zi \For $\id{row}$ in $A$ \Do
     \zi     $k \gets 0$
@@
     \zi         \For $\j \gets i \To D$ \Do
-    \zi         $\vec{b}[k] \gets \vec{x}[i] \cdot \vec{x}[j]$
+    \zi             $B[i, k] \gets \id{row}[i] \cdot \id{row}[j]$
     \zi             $k \gets k + 1$
@@
        	\End
-    return $\vec{x} | \vec{b}$
 \end{codebox}
 
-Notice that the way the column the polynomial feature between features $i$ and $j$ can be determined
+Notice that, for an individual row, the way the column the polynomial feature between features $i$ and $j$ can be determined
 by a simple counter $k$ that is incremented each pass through the inner loop. This cannot be
@@
 
-What is therefore needed is a mapping between column index pairs of $\vec{x}$ and the space ${0, 1, ..., \binom{D+1}{D-1}-1}$
-(the output size of second degree polynomial features when $D = |\vec{x}|$. The ordering of the mapping is irrelevant so long as
-it its input $ I = {(a, b) : a, b \in \{0,1,..., D-1\} \land a \le b}$ is bijective with its output, ${0, 1, ..., \binom{D+1}{D-1}-1}$.
+What is therefore needed is a mapping between column index pairs of $\vec{x}$ and the space ${0, 1, ..., \frac{D^2+D}{2}-1}$
+(the output size of second degree polynomial features when $D = |\vec{x}|$). The ordering of the mapping is irrelevant so long as its input
+
+\begin{equation*} \label{eq:input}
+I \defeq \{(a, b) : a, b \in \{0,1,..., D-1\} \land a \le b\}
+\end{equation*}
+
+is bijective with its output
+
+\begin{equation*} \label{eq:output}
+O \defeq \{0, 1, ..., \frac{D^2+D}{2}-1\}
+\end{equation*}
 
@@
      \cdot   & \cdot  &       &         &  \cdot  &  \cdot  &  \cdot       \\
-    x_{D-1} & -   & \cdot    &  \cdot  & \cdot   & -    &  \frac{D(D+1)}{2}-1\\
+    x_{D-1}  & -      & \cdot &  \cdot  & \cdot   & -    &  \frac{D^2+D}{2}-1\\
   }
@@
 
-This is a rather natural sequencial ordering on both the input and output spaces. As can 
-be seen, the required mapping can e viewed as mapping the upper triangular matrix (including 
-the diagonal) to $0, 1, ..., \frac{D(D+1)}{2}-1$. The construction of this mapping will now
+Again, any bijective mapping $I \mapsto O$ would suffice, but this particular mapping could
+be considered one of the more natural options as it maps $I$ onto $O$ where both $I$ and $O$ are sorted in ascending order.
+As can be seen, the required mapping can be viewed as mapping the upper triangular matrix (including 
+the diagonal), whose indicies are $I$, onto $O$. The construction of this mapping will now
 be given.
 
+\newpage
+INSERT JOHN'S PROOF HERE
+
+The mapping can be modified slightly to yield interaction features instead of polynomial features
+by noting that the section of the $D \times D$ mapping matrix that will yield interaction
+features differs only in that its diagonal is excluded. The mapping is therefore gotten by
+taking the polinomial mapping and subtracting $i+1$.
+
+\newpage
+
+\section{The Algorithm}
+In section X we showed why a mapping from the space X
+onto the space X is necessary for calculating polynomial features for a sparse matrix, then
+derived such a mapping.
+
+We now have the necessary components for an algorithm to efficiently calculate polynomial features for
+a sparse matrix. The matrix needs to be stored in a form that allows for its nonzero column indices, e.g. in sparse row form,
+and we will use the mapping derived in X to determine which column in the output space
+the polynomial feature between two input columns is mapping onto. Combining these ideas yields the following algorithm:
+
+\begin{codebox}
+\Procname{$\proc{Sparse Polynomial Features}(A)$}
+    \zi $\func{PolyMap}(a, b) = \frac{2Da-a^2+2b-3a-2}{2}+a+1$
+    \zi $N \gets$ row count of $A$
+    \zi $D \gets$ column count of $A$
+    \zi $B$ $\gets$ Compressed Sparse Row Matrix of size $N \times \frac{D^2+D}{2}$
+    \zi \For $\id{row}$ in $A$ \Do
+    \zi     $N_{zc} \gets$ nonzero columns of $row$
+    \zi     \For $i \gets 0 \To |N_{zc}|$ \Do
+    \zi         \For $j \gets i \To |N_{zc}|$ \Do
+    \zi             $k \gets \func{PolyMap}(i, j)$
+    \zi             $r \gets$ index of $\id{row}$
+    \zi             $B[r, k] \gets \id{row}[i] \cdot \id{row}[j]$
+                \End
+            \End
+       	\End
+\end{codebox}
+
+Using the mapping for interaction features and changing the bounds in the loop conditions, 
+the algorithm can be modified to give second order interaction features:
+
+\begin{codebox}
+\Procname{$\proc{Sparse Interaction Features}(A)$}
+    \zi $\func{InterMap}(a, b) = \frac{2Da-a^2+2b-3a-2}{2}$
+    \zi $N \gets$ row count of $A$
+    \zi $D \gets$ column count of $A$
+    \zi $B$ $\gets$ Compressed Sparse Row Matrix of size $N \times \frac{D^2+D}{2}$
+    \zi \For $\id{row}$ in $A$ \Do
+    \zi     $N_{zc} \gets$ nonzero columns of $row$
+    \zi     \For $i \gets 0 \To |N_{zc}|-1$ \Do
+    \zi         \For $j \gets i+1 \To |N_{zc}|$ \Do
+    \zi             $k \gets \func{InterMap}(i, j)$
+    \zi             $r \gets$ index of $\id{row}$
+    \zi             $B[r, k] \gets \id{row}[i] \cdot \id{row}[j]$
+                \End
+            \End
+       	\End
+\end{codebox}
+
+where $N_{zc}$ are the columns of nonzero elements of row $\id{row}$.
+
+\section{Analysis}
+
+\subsection{Analytical}
+We assume the input matrix $A$ is in sparse row form. The steps prior to the first loop are
+constant operations. The outer loop will be executed $N$ times (once for each row). For each row, 
+the innermost loop will be executed and the inner loop will be $\frac{|N_{zc}|^2+|N_{zc}|}{2}$ times.
+If the density is uniformly distributed across all rows, $E[|N_{zc}|] = dD$, but this cannot be assumed. In the worst case,
+then, the innermost loop will each be executed $\frac{D^2+D}{2}$ times, but on average, $\frac{d^2D^2+dD}{2}$ times.
+The operations in the innermost loop are constant. The time complexity is therefore $O(ND^2)$ and $\Theta(Nd^2D^2)$.
+
+It's often the case in a machine learning setting that the input matrix has near uniform density across its rows.
+For example, if the matrix is sparse due to one-of-m encoding (also known as one-hot vectors), each row vector will
+have the same number of nonzero components.
+
+\subsection{Empirical}
+To assess the performance of the algorithm on data of varied density, we compare its speed
+to the PolynomialFeatures class of the popular machine learning scikit-learn \cite{scikit-learn}.
+Uniformly random matrices of size $1000 \times D$ were generated for $d=0.2$, $0.4$, $0.6$, and $0.8$, and 
+$D=1$, $101$, $201$, $301$, $401$, and $501$. Since the scikit-learn version of the algorithm
+does not exploit sparsity, we assume the method's time is invariate with respect the density of the matrix, but
+we gave it the lowest density of each dimensionality to be fair. The results are shown in figure \eqref{fig:benchmark}.
+
+\begin{figure}
+\includegraphics[scale=0.3]{benchmark_results.png}
+\caption{This plot shows how long the sparse polynomial method took to run on matrices with 1000 rows of various dimensionalities and densities and compared to the scikit-learn PolynomialFeatures class on various dimensionalities.}
+\label{fig:benchmark}
+\centering
+\end{figure}
+
+As can be seen, the sparse method starts out slower than scikit-learn for each density. This
+is either due to the overhead required by the sparse method (e.g. utilizing the mapping function),
+or is due to implementation differences. The sparse method eventually becomes faster than scikit-learn
+for each density. Note the excellent performance of $d=0.2$. This level of density is not uncommon when
+working with sparse matrices.
+
+The above benchmark was done on synthetic data. To more realistically determine the performance of
+the algorithm, we applied it to two real world datasets: 20newsgroups\footnote{\url{http://scikit-learn.org/stable/datasets/twenty_newsgroups.html}} and connect-4, which was obtained
+from the UCI Machine Learning Repository \cite{Lichman:2013}\footnote{\url{https://archive.ics.uci.edu/ml/datasets/Connect-4}}.
+We compare the time and space required by the sparse method and scikit-learn in table \eqref{table:results}.
+
+\begin{center}
+    \begin{tabular}{c | c | c} \label{table:results}
+ %       \hline
+        dataset             & 20newsgroups  & connect-4 \\ \hline
+        instances           & 11,314        & 67,557    \\ \hline
+        features            & 130,107       & 126       \\ \hline
+        polynomial features & 8,463,980,778 & 8,127     \\ \hline
+        density             & 0.12          & 0.11      \\ \hline
+        dense space (MB)    & Memory Error  & 4191      \\ \hline
+        sparse space (MB)   & 5333          & 735       \\ \hline
+        dense time (s)      & Memory Error  & 26        \\ \hline
+        sparse time (s)     & 109           & 44        \\ \hline
+%        \hline
+    \end{tabular}
+\end{center}
+
+Notice that scikit-learn was unable to calculate polynomial features for the 20newsgroups 
+matrix; a memory error was encountered. The sparse method, however, succeeded in 109 seconds.
+The resulting matrix had nearly 8.5 billion features. While most machine learning libraries would
+struggle with such a high dimensionality, some, such as the system described in \cite{agarwal2014reliable} (which is 
+available as part of the Vowpal Wabbit package\footnote{\url{https://github.com/JohnLangford/vowpal_wabbit/wiki}}), would likely be capable
+of learning a model from such data.
+
+The connect-4 dataset had its polynomial features computed faster via the dense method than the sparse. This
+is likely because the dimensionality is sufficiently low for the overhead of the sparse method
+to not have overtaken the dense method in terms of speed. However, the sparse method took less than
+a fifth of the memory and took less than twice the speed, so using the sparse method during parameter 
+optimization in a multi-core setting would drastically increase parameter search throughput.
+
+\section{Summary \& Future Work}
+In this work we have given a method of calculating second degree polynomial features on sparse matrices
+with time complexity that quadratically decreases with respect to the density of the matrix.
+
+This work also gives a general pattern for calculating polynomial features of arbitrary degrees:
+instead of finding a mapping from the 2-tuples of indices onto the space $\{0, 1, ..., \frac{D^2+D}{2}-1\}$,
+construct a mapping from k-tuples of indices onto the space $\{0, 1, ..., \binom{D+k-1}{D-1}-1\}$, where $k$ is the degree of the polynomial.
+In terms of the upper triangular matrix analogy, this corresponds with mapping the upper k-simplex of a $(D, D, ..., D)$-tensor onto
+the space $\{0, 1, ..., \binom{D+k-1}{D-1}-1\}$. For the average case of $dD$ nonzero elements per row, 
+there will be $\binom{dD+k-1}{dD-1} = $ polynomial features, so the average time complexity for an $N \times D$ matrix
+is $\Theta(Nd^kD^k)$, which means that its time complexity polynomially decreases with respect to the density
+compared to the dense algorithm (which has $O(ND^k)$ and $\Theta(ND^3)$).
+
 \vskip 0.2in
-\bibliography{sample}
-
+\bibliography{paper2}{}
+\bibliographystyle{plain}
 \end{document}

### 2015-09-21 paper/paper2.tex
need john's proof

@@
 We give an algorithm to efficiently calculate second degree polynomial or interaction 
-that has complexity quadratically (with respect to the density of the matrix) better than 
+that has complexity quadratically with respect to the density of the matrix, and can therefore be much faster than 
 the naive approach and allows for the matrix to remain in sparse form. We apply the method

### 2015-09-21 paper/paper2.tex
need john's proof

@@
 
-which results in $\binom{D+1}{D-1} = \frac{D^2+D}{2}$ additional features, and so the generation of 
+which results in $\binom{D+1}{D-1} = \frac{D^2+D}{2}$ additional features, so the generation of 
 these features for a matrix of size $N \times D$ has a time complexity of $O(ND^2)$. In the general case of degree $k$, there 
@@
 
-The first of these necessary components is gotten simply by using the appropriate data structure
-to store the sparse matrix (e.g. storing it in compressed sparse row format). The second component 
+The first of these necessary components is obtained simply by using the appropriate data structure
+to store the sparse matrix (e.g. storing it in compressed sparse row form). The second component 
 is not only less readily available, but its necessity is less obvious. To make this need more clear, consider the
@@
 is $\Theta(Nd^kD^k)$, which means that its time complexity polynomially decreases with respect to the density
-compared to the dense algorithm (which has $O(ND^k)$ and $\Theta(ND^3)$).
+compared to the dense algorithm (which has $O(ND^k)$ and $\Theta(ND^k)$).
 

### 2015-09-21 paper/various_spares_interaction.tex
need john's proof

--- 2015-08-07 paper/spares_interaction.tex
@@
 that has time and space complexities that decrease quadratically with the density of the input matrix with respect to the naive approach.
+
+$\{x_i \cdot x_j : i, j \in \{0,1,..., D-1\} \land i \le j\}$  \\
+\\
+$\{(i,j) : i, j \in \{0,1,..., D-1\} \land i \le j\} \to \{0, 1, ..., \frac{D^2+D}{2}-1\}$ \\
+
+$i$
+$j$
+
+$\vec{x}$
+
+\begin{codebox}
+\Procname{$\proc{Dense Polynomials}(A)$}
+    \zi $N \gets$ row count of $A$
+    \zi $D \gets$ column count of $A$
+    \zi $B$ $\gets$ Matrix of size $N \times \frac{D^2+D}{2}$
+    \zi \For $\id{row}$ in $A$ \Do
+    \zi     $k \gets 0$
+    \zi     \For $\i \gets 0 \To D$ \Do
+    \zi         \For $\j \gets i \To D$ \Do
+    \zi             $B[i, k] \gets \id{row}[i] \cdot \id{row}[j]$
+    \zi             $k \gets k + 1$
+                \End
+            \End
+       	\End
+\end{codebox}
+
+$[a, b, c, d, e]$
+
+$[ab, ac, ad, ae, bc, bd, be, cd, ce, de]$
+
+\renewcommand{\kbldelim}{(}% Left delimiter
+\renewcommand{\kbrdelim}{)}% Right delimiter
+\[
+  \kbordermatrix{
+        & x_0 & x_1 & x_2 & x_3 & x_4 \\
+    x_0 & 0 & 1 & 2 & 3 & 4\\
+    x_1 & - & 5 & 6 & 7 & 8\\
+    x_2 & - & - & 9 & 10 & 11\\
+    x_3 & - & - & - & 12 & 13\\
+    x_4 & - & - & - & - & 14\\
+  }
+\]
+
+$\text{polynomial-index}(i, j | D) = \frac{2Di-i^2+2j-3i-2}{2}+i+1$
 
@@
 %1 & 2 & 0 & 3 \end{array} \right)\]
+\[ \left( \begin{array}{ccccc}
+A & b & c & d \end{array} \right)\]
+
+%\[ \left( \begin{array}{cccccccccc}
+%1 & 2 & 0 & 3 & 0 \end{array} \right)\]
 
@@
 \begin{codebox}
-\Procname{$\proc{Sparse Interaction}(A)$}
-    \zi $\func{map}(b, a) = \frac{2Da-a^2+2b-3a-2}{2}$
-    \zi $B$ $\gets$ Compressed Sparse Row Matrix of size $N \times \frac{D^2-D}{2}$
+\Procname{$\proc{Sparse Polynomials}(A)$}
+    \zi $\func{map}(a, b) = \frac{2Da-a^2+2b-3a-2}{2}+a+1$
+    \zi $N \gets$ row count of $A$
+    \zi $D \gets$ column count of $A$
+    \zi $B$ $\gets$ Compressed Sparse Row Matrix of size $N \times \frac{D^2+D}{2}$
     \zi \For $\id{row}$ in $A$ \Do
     \zi     $N_{zc} \gets$ nonzero columns of $row$
-    \zi     \For each combination $a, b$ of elements of $N_{zc}$ \Do
-    \zi         $k \gets \func{map}(b, a)$
-    \zi         $i \gets$ index of $\id{row}$
-    \zi         $B[i, k] \gets \id{row}[a] \cdot \id{row}[b]$
+    \zi     \For $i \gets 0 \To |N_{zc}|$ \Do
+    \zi         \For $j \gets i \To |N_{zc}|$ \Do
+    \zi             $k \gets \func{map}(i, j)$
+    \zi             $r \gets$ index of $\id{row}$
+    \zi             $B[r, k] \gets \id{row}[i] \cdot \id{row}[j]$
+                \End
             \End

### 2015-09-21 paper/paper2.tex
need john's proof
fixed abstract

@@
 \usepackage{svg}
+\usepackage{pdfpages}
 
@@
 \begin{abstract}%   <- trailing '%' for backward compatibility of .sty file
-We give an algorithm to efficiently calculate second degree polynomial or interaction 
-that has complexity quadratically with respect to the density of the matrix, and can therefore be much faster than 
-the naive approach and allows for the matrix to remain in sparse form. We apply the method
-to several real datasets and give both analytical and empirical comparisons with the
-naive approach.
+We give an algorithm to efficiently calculate second degree polynomial or interaction features for a sparse matrix.
+The algorithm has average time complexity that decreases quadratically with respect to the density of the matrix, which
+is a large improvement over the naive approach. It also allows for the matrix to remain in sparse form, reducing memory demands.
+We apply the method to several real datasets and give both analytical and empirical comparisons with the
+naive approach. The work also gives a generalizable method for creating algorithms for higher orders of polynomials.
 \end{abstract}
@@
 \end{sidewaysfigure}
+
+%Same as above, but in vector graphic form. Use this in the final.
+%\begin{sidewaysfigure}
+%    \centering
+%    \includegraphics[scale=0.65]{classifier_boundaries.pdf}
+%    \caption{Decision boundaries for four classifier types on four different datasets. The darker the color, the more certain the decision.
+%             The plain dataset is shown in the first column. Each other column is a type of classifier applied to that dataset. The classifier
+%             types are logistic regression, logistic regression with second degree polynomial features, RBF SVM, and gradient boosting with decision trees.
+%             The accuracy of each model is shown in the bottom right corner of that model's subplot.}
+%    \label{fig:boundaries}
+%\end{sidewaysfigure}
 

### 2015-09-21 paper/paper2.tex
fixed the size of the output matrix in interaction algo

@@
 
-Using the mapping for interaction features and changing the bounds in the loop conditions, 
+Using the mapping for interaction features and changing the bounds in the loop conditions and the size of the output matrix, 
 the algorithm can be modified to give second order interaction features:
@@
     \zi $D \gets$ column count of $A$
-    \zi $B$ $\gets$ Compressed Sparse Row Matrix of size $N \times \frac{D^2+D}{2}$
+    \zi $B$ $\gets$ Compressed Sparse Row Matrix of size $N \times \frac{D^2-D}{2}$
     \zi \For $\id{row}$ in $A$ \Do

### 2015-09-21 paper/paper2.tex
double dolumn
added pagenum to fig reference
back to one col

@@
-\documentclass[11pt]{article}
+\documentclass[11pt,onecolumn]{article}
 
@@
 \usepackage{rotating}
-\usepackage{svg}
 \usepackage{pdfpages}
@@
 \section{Introduction}
+
 When performing a modeling task, one is often faced with a non-linearly separable problem,
@@
 
-\begin{sidewaysfigure}
-    \centering
-    \includegraphics[scale=0.5]{classifier_boundaries.png}
-    \caption{Decision boundaries for four classifier types on four different datasets. The darker the color, the more certain the decision.
-             The plain dataset is shown in the first column. Each other column is a type of classifier applied to that dataset. The classifier
-             types are logistic regression, logistic regression with second degree polynomial features, RBF SVM, and gradient boosting with decision trees.
-             The accuracy of each model is shown in the bottom right corner of that model's subplot.}
-    \label{fig:boundaries}
-\end{sidewaysfigure}
-
 %Same as above, but in vector graphic form. Use this in the final.
@@
 %\end{sidewaysfigure}
+
+\begin{sidewaysfigure*}[ht!]
+    \centering
+    \includegraphics[scale=0.5]{classifier_boundaries.png}
+    \caption{Decision boundaries for four classifier types on four different datasets. The darker the color, the more certain the decision.
+             The plain dataset is shown in the first column. Each other column is a type of classifier applied to that dataset. The classifier
+             types are logistic regression, logistic regression with second degree polynomial features, RBF SVM, and gradient boosting with decision trees.
+             The accuracy of each model is shown in the bottom right corner of that model's subplot.}
+    \label{fig:boundaries}
+\end{sidewaysfigure*}
 
@@
 
-\newpage
 INSERT JOHN'S PROOF HERE
@@
 taking the polinomial mapping and subtracting $i+1$.
-
-\newpage
 
@@
 does not exploit sparsity, we assume the method's time is invariate with respect the density of the matrix, but
-we gave it the lowest density of each dimensionality to be fair. The results are shown in figure \eqref{fig:benchmark}.
+we gave it the lowest density of each dimensionality to be fair. The results are shown in figure \eqref{fig:benchmark} on page \pageref{fig:benchmark}.
 
-\begin{figure}
+\begin{figure*}[ht!]
     \includegraphics[scale=0.3]{benchmark_results.png}
@@
     \centering
-\end{figure}
+\end{figure*}
 
@@
         dataset             & 20newsgroups  & connect-4 \\ \hline
-        instances           & 11,314        & 67,557    \\ \hline
-        features            & 130,107       & 126       \\ \hline
-        polynomial features & 8,463,980,778 & 8,127     \\ \hline
+        instances           & 11314        & 67557    \\ \hline
+        features            & 130107       & 126       \\ \hline
+        polynomial features & 8463980778 & 8,127     \\ \hline
         density             & 0.12          & 0.11      \\ \hline

### 2015-09-22 paper/paper2.tex
made more clear

@@
 
-Notice that, for an individual row, the way the column the polynomial feature between features $i$ and $j$ can be determined
+Notice that, for an individual row, the location of a polynomial feature ($\id{row}i \cdot \id{row}j$) can be determined
 by a simple counter $k$ that is incremented each pass through the inner loop. This cannot be

### 2015-09-22 paper/paper2.tex
made more clear

@@
 
-Notice that, for an individual row, the location of a polynomial feature ($\id{row}i \cdot \id{row}j$) can be determined
+Notice that, for an individual row, the location of a polynomial feature ($\id{row}[i] \cdot \id{row}[j]$) can be determined
 by a simple counter $k$ that is incremented each pass through the inner loop. This cannot be

### 2015-09-24 paper/paper2.tex
made more clear
table caption
.

@@
 
-\begin{sidewaysfigure*}[ht!]
+\begin{sidewaysfigure*}[ht!, numbered]
     \centering
@@
 
-It's often the case in a machine learning setting that the input matrix has near uniform density across its rows.
+In machine learning, an input matrix often has near uniform density across its rows.
 For example, if the matrix is sparse due to one-of-m encoding (also known as one-hot vectors), each row vector will
@@
 
-\begin{figure*}[ht!]
+\begin{figure*}[ht!, numbered]
     \includegraphics[scale=0.3]{benchmark_results.png}
@@
 
-\begin{center}
-    \begin{tabular}{c | c | c} \label{table:results}
+\begin{table}
+    \centering
+    \begin{tabular}[numbered]{c | c | c} \label{table:results}
  %       \hline
@@
     \end{tabular}
-\end{center}
+    \caption{Time and space comparison of second degree polynomial feature calculation between the sparse method and scikit-learn.}
+\end{table}
 

### 2015-09-30 compare.py
scripts for benchmarks and demos

@@ -1,7 +1,7 @@
 import sparse_interaction
 from sklearn.datasets import fetch_20newsgroups_vectorized
 from sklearn.cross_validation import StratifiedKFold
-from sklearn.svm import LinearSVC
+from sklearn.linear_model import SGDClassifier
 from sklearn.metrics import f1_score
 
 dat = fetch_20newsgroups_vectorized()
@@ -9,12 +9,14 @@ X = dat.data
 Y = dat.target
 cv = StratifiedKFold(Y)
 
+X = X[:, :20000]
+
 si = sparse_interaction.SparseInteractionFeatures()
 X_i = si.transform(X)
 
 scores, scores_i = [], []
 
-clf = LinearSVC()
+clf = SGDClassifier(penalty='l1', n_iter=10)
 
 for train, test in cv:
     clf.fit(X[train], Y[train])

### 2015-09-30 random_matrix_compare.py

@@ -1,0 +1,158 @@
+from time import time
+from scipy.sparse import rand
+from sparse_interaction import SparsePolynomialFeatures
+import matplotlib.pyplot as plt
+from sklearn.preprocessing import PolynomialFeatures
+from code import interact
+from numpy import arange, median
+from memory_profiler import profile
+import StringIO
+
+N = 1000
+iters=1
+
+sp = SparsePolynomialFeatures()
+poly = PolynomialFeatures(include_bias=False)
+
+dD_to_times = {}
+other_d_to_times={}
+
+min_D = 1
+D_step = 100
+max_D = 500+D_step
+
+
+min_d = 0.2
+d_step = .2
+max_d = 1.
+
+def speed_test():
+    for i in xrange(iters):
+        for d in arange(min_d, max_d, d_step):
+            for D in arange(min_D, max_D, D_step):
+                X = rand(N, D, d).tocsr()
+                X_d = X.toarray()
+        
+                a = time()
+                A = sp.transform(X)
+                t = time()-a
+                if d not in dD_to_times:
+                    dD_to_times[d] = {}
+            
+                dD_to_times[d][D] = dD_to_times[d].get(D, 0) + t
+                print i, d, D, t
+            
+                if d == max_d-d_step:
+                    a = time()
+                    B = poly.fit_transform(X_d)
+                    t = time()-a
+                    other_d_to_times[D] = other_d_to_times.get(D, 0) + t
+
+
+    for d in sorted(dD_to_times.keys()):
+        for D in dD_to_times[d].keys():
+            dD_to_times[d][D] /= iters
+        pairs = dD_to_times[d].items()
+        pairs.sort()
+        x, y = zip(*pairs)
+        plt.plot(x, y, label='d=%s'%(d,))
+    
+    for k, val in other_d_to_times.iteritems():
+        other_d_to_times[k] /= iters
+
+    pairs = other_d_to_times.items()
+    pairs.sort()
+    x, y = zip(*pairs)
+    plt.plot(x, y, label='scikit-learn', color='red', marker='*')
+
+    plt.title('Times of Polynomial Feature Generation on Random Matrices With %s Rows' % (N,))
+    plt.xlabel('Dimensionality')
+    plt.ylabel('Time to Calculate (seconds)')
+    plt.legend(loc='best')
+    plt.xlim([0, max_D-D_step])
+    plt.grid(True)
[…]

### 2015-09-30 regression.py

@@ -1,0 +1,34 @@
+from sklearn.linear_model import LinearRegression
+from sklearn.preprocessing import PolynomialFeatures
+import matplotlib.pyplot as plt
+import numpy as np
+
+X = np.arange(0, 1*np.pi, 0.01)
+X = X.reshape(X.shape[0], 1)
+X_poly = PolynomialFeatures(degree=2, include_bias=False).fit_transform(X)
+Y = np.sin(X).flatten()
+noise = np.random.normal(0, 0.1, len(Y))
+print Y.shape, noise.shape
+Y = Y+noise
+
+clf1 = LinearRegression()
+clf1.fit(X, Y)
+pred1 = clf1.predict(X)
+
+clf2 = LinearRegression()
+clf2.fit(X_poly, Y)
+pred2 = clf2.predict(X_poly)
+
+fig, ax = plt.subplots(1)
+ax.scatter(X, Y, alpha=0.3)
+ax.plot(X, pred1, label="Linear Prediction")
+ax.plot(X, pred2, label="Polynomial Prediction")
+ax.legend()
+ax.set_xlabel(r'$x$')
+ax.set_ylabel(r'$y$')
+textstr = r'$y = sin(x) + \mathit{N}(0, 0.1)$'
+props = dict(boxstyle='round', facecolor='wheat', alpha=0.5)
+ax.text(0.05, 0.95, textstr, transform=ax.transAxes, fontsize=14,
+        verticalalignment='top', bbox=props)
+ax.set_title('Goodness of fit of polynomial features to a trigonomic function')
+plt.show()

### 2015-09-30 time_space_compare.py

@@ -1,0 +1,27 @@
+from memory_profiler import profile
+from sparse_interaction import SparsePolynomialFeatures
+from sklearn.preprocessing import PolynomialFeatures
+from cPickle import load
+from scipy.sparse import csr_matrix
+from sklearn.datasets import fetch_20newsgroups_vectorized
+from time import time
+
+#@profile
+def sparse(X):
+    si = SparsePolynomialFeatures()
+    X_i = si.transform(X)
+    
+#@profile
+def dense(X):
+    poly = PolynomialFeatures(include_bias=False)
+    X_i = poly.fit_transform(X)
+
+if __name__ == '__main__':
+    #X_filename = 'connect_4/X'
+    #X = load(open(X_filename))
+    #X_sparse = csr_matrix(X)
+    X_sparse = fetch_20newsgroups_vectorized().data
+    X = X_sparse.toarray()
+
+    a=time(); dense(X); print time()-a
+    #a=time(); sparse(X_sparse); print time()-a

### 2015-10-02 paper/paper2.tex
fixed row index in algo
properly slicing into nz cols
two cols and all figs eps

@@
-\documentclass[11pt,onecolumn]{article}
+\documentclass[11pt,twocolumn]{article}
 
@@
 
-%Same as above, but in vector graphic form. Use this in the final.
-%\begin{sidewaysfigure}
-%    \centering
-%    \includegraphics[scale=0.65]{classifier_boundaries.pdf}
-%    \caption{Decision boundaries for four classifier types on four different datasets. The darker the color, the more certain the decision.
-%             The plain dataset is shown in the first column. Each other column is a type of classifier applied to that dataset. The classifier
-%             types are logistic regression, logistic regression with second degree polynomial features, RBF SVM, and gradient boosting with decision trees.
-%             The accuracy of each model is shown in the bottom right corner of that model's subplot.}
-%    \label{fig:boundaries}
-%\end{sidewaysfigure}
+To demonstrate the benefit of polynomial features in regression, we two ordinary least squares models to $y=sin(x) + \mathcal{N}(0, 0.1), x \in [0, \pi]$.
+The first model (the blue curve) was given only $x$ as its input.
+The second model (the green curve) was given $x$ and second degree polynomial features derived from $x$ (namely $x^2$).
+The curves the models learned can be seen in figure \ref{fig:regression}.
 
-\begin{sidewaysfigure*}[ht!, numbered]
+\begin{sidewaysfigure}
     \centering
-    \includegraphics[scale=0.5]{classifier_boundaries.png}
-    \caption{Decision boundaries for four classifier types on four different datasets. The darker the color, the more certain the decision.
+    \includegraphics[scale=0.3]{boundaries.eps}
+
+     \caption{\footnotesize{Decision boundaries for four classifier types on four different datasets. The darker the color, the more certain the decision.
              The plain dataset is shown in the first column. Each other column is a type of classifier applied to that dataset. The classifier
              types are logistic regression, logistic regression with second degree polynomial features, RBF SVM, and gradient boosting with decision trees.
-             The accuracy of each model is shown in the bottom right corner of that model's subplot.}
+             The accuracy of each model is shown in the bottom right corner of that model's subplot.}}
     \label{fig:boundaries}
-\end{sidewaysfigure*}
+\end{sidewaysfigure}
+
+\begin{figure}
+    \centering
+    \includegraphics[scale=0.4]{regression.eps}
+     \caption{\footnotesize{Two linear regression models fit to $y=sin(x) + \mathcal{N}(0, 0.1), x \in [0, \pi]$.
+The blue line's model was given only $x$, while the green line's model was given $x$ and second degree polynomial features.}}
+    \label{fig:regression}
+\end{figure}
+
+%\newpage
 
@@
 \begin{codebox}
+\footnotesize
 \Procname{$\proc{Dense Polynomial Features}(A)$}
@@
     \zi         \For $\j \gets i \To D$ \Do
-    \zi             $B[i, k] \gets \id{row}[i] \cdot \id{row}[j]$
+    \zi             $r \gets$ index of $\id{row}$
+    \zi             $B[r, k] \gets \id{row}[i] \cdot \id{row}[j]$
     \zi             $k \gets k + 1$
@@
 \begin{codebox}
+\footnotesize
 \Procname{$\proc{Sparse Polynomial Features}(A)$}
@@
     \zi $D \gets$ column count of $A$
-    \zi $B$ $\gets$ Compressed Sparse Row Matrix of size $N \times \frac{D^2+D}{2}$
+    \zi $B$ $\gets$ empty sparse $N \times \frac{D^2+D}{2}$ matrix
     \zi \For $\id{row}$ in $A$ \Do
@@
     \zi         \For $j \gets i \To |N_{zc}|$ \Do
-    \zi             $k \gets \func{PolyMap}(i, j)$
+    \zi             $k \gets \func{PolyMap}(N_{zc}[i], N_{zc}[j])$
     \zi             $r \gets$ index of $\id{row}$
-    \zi             $B[r, k] \gets \id{row}[i] \cdot \id{row}[j]$
+    \zi             $B[r, k] \gets \id{row}[N_{zc}[i]] \cdot \id{row}[N_{zc}[j]]$
                 \End
@@
 \begin{codebox}
+\footnotesize
 \Procname{$\proc{Sparse Interaction Features}(A)$}
@@
     \zi $D \gets$ column count of $A$
-    \zi $B$ $\gets$ Compressed Sparse Row Matrix of size $N \times \frac{D^2-D}{2}$
+    \zi $B$ $\gets$ empty sparse $N \times \frac{D^2-D}{2}$ matrix
     \zi \For $\id{row}$ in $A$ \Do
@@
     \zi         \For $j \gets i+1 \To |N_{zc}|$ \Do
-    \zi             $k \gets \func{InterMap}(i, j)$
+    \zi             $k \gets \func{InterMap}(N_{zc}[i], N_{zc}[j])$
     \zi             $r \gets$ index of $\id{row}$
-    \zi             $B[r, k] \gets \id{row}[i] \cdot \id{row}[j]$
+    \zi             $B[r, k] \gets \id{row}[N_{zc}[i]] \cdot \id{row}[N_{zc}[j]]$
                 \End
@@
 to the PolynomialFeatures class of the popular machine learning scikit-learn \cite{scikit-learn}.
-Uniformly random matrices of size $1000 \times D$ were generated for $d=0.2$, $0.4$, $0.6$, and $0.8$, and 
-$D=1$, $101$, $201$, $301$, $401$, and $501$. Since the scikit-learn version of the algorithm
+Uniformly random matrices of size $100 \times D$ were generated for $d=0.1$, $0.2$, $0.3$, $0.4$, and $0.5$, and 
+$D=1$, $101$, $201$, $301$, $401$, and $501$. Each point was resampled five times and averaged to decrease variance. Since the scikit-learn version of the algorithm
 does not exploit sparsity, we assume the method's time is invariate with respect the density of the matrix, but
@@
 
-\begin{figure*}[ht!, numbered]
-    \includegraphics[scale=0.3]{benchmark_results.png}
+\begin{figure}[ht, numbered]
+    \includegraphics[scale=0.4]{benchmark.eps}
     \caption{This plot shows how long the sparse polynomial method took to run on matrices with 1000 rows of various dimensionalities and densities and compared to the scikit-learn PolynomialFeatures class on various dimensionalities.}
@@
     \centering
-\end{figure*}
+\end{figure}
 
@@
     \centering
+    \small    
     \begin{tabular}[numbered]{c | c | c} \label{table:results}
@@
         dataset             & 20newsgroups  & connect-4 \\ \hline
-        instances           & 11314        & 67557    \\ \hline
+        instances           & 11,314        & 67,557    \\ \hline
         features            & 130107       & 126       \\ \hline
-        polynomial features & 8463980778 & 8,127     \\ \hline
+        polynomial features & 8,463,980,778 & 8,127     \\ \hline
         density             & 0.12          & 0.11      \\ \hline

### 2015-10-05 paper/paper2.tex
reworded intro

@@
 
-To help visualize what second degree polynomial features are capable of in a machine learning paradigm,
-we trained a linear classifier (logistic regression) on four different datasets and show their decision
-boundaries (figure \eqref{fig:boundaries} on page \pageref{fig:boundaries}) with and without polynomial features, along with their accuracy on a holdout set. We compare
-these to an RBF SVM via scikit-learn\footnote{\url{http://scikit-learn.org/stable/modules/generated/sklearn.svm.SVC.html}} 
-and gradient boosted decision trees via the xgboost package\footnote{\url{https://github.com/dmlc/xgboost/}}.
+To show the types of nonlinearity polynomial features are capable of imbuing a linear model with,
+we trained logistic regression models on four inseparable datasets and plot their decision boundaries, as seen in figure \eqref{fig:boundaries} on page \pageref{fig:boundaries}. We compare
+these decision boundaries to those learnt by more complex methods, namely a support vector machine with a radial basis function kernel\footnote{\url{http://scikit-learn.org/stable/modules/generated/sklearn.svm.SVC.html}}
+and gradient boosted decision trees\footnote{\url{https://github.com/dmlc/xgboost/}}. As can be seen, polynomial features allow a simple linear classifier is capable of drawing decision boundaries
+that in the form of arbitrary parabolas, hyperbolas, and elipses.
 
-To demonstrate the benefit of polynomial features in regression, we two ordinary least squares models to $y=sin(x) + \mathcal{N}(0, 0.1), x \in [0, \pi]$.
+Beyond classification, polynomial regression \cite{theoryoflinearmodels} is a staple regression method.
+We provide a demonstration of the power polynomial features can lend to a regression problem by fitting two ordinary least squares models to $y=sin(x) + \mathcal{N}(0, 0.1), x \in [0, \pi]$.
 The first model (the blue curve) was given only $x$ as its input.
@@
 The curves the models learned can be seen in figure \ref{fig:regression}.
+
+While polynomial features are a well known and used concept in machine learning (e.g.
+\cite{pavlidis2002learning, konidaris2009skill, wiesler2009investigations}) and work has been done to improve their utilization (e.g.
+\cite{pevckov2008minimal, huang2010predicting}), this is the first work that we know of that exploits matrix sparcity
+to improve the speed with which they are calculated and allows them to be calculated for a sparse matrix.
 
@@
 %\newpage
-
-With polynomial features added, a simple linear classifier is capable of drawing decision boundaries
-that take on the form of arbitrary parabolas, hyperbolas, and elipses. Beyond classification, polynomial regression \cite{theoryoflinearmodels} is a staple regression method.
-While polynomial features are a well known and used concept in machine learning (e.g.
-\cite{pavlidis2002learning, konidaris2009skill, wiesler2009investigations}) and work has been done to improve their utilization (e.g.
-\cite{pevckov2008minimal, huang2010predicting}), this is the first work that we know of that exploits matrix sparcity
-to improve the speed with which they are calculated and allows them to be calculated for a sparse matrix.
 

### 2015-10-05 paper/paper2.tex
resized boundaries image to not cut off text
cleaned up conclusion

@@
     \centering
-    \includegraphics[scale=0.3]{boundaries.eps}
+    \includegraphics[scale=0.25]{boundaries.eps}
 
@@
 the space $\{0, 1, ..., \binom{D+k-1}{D-1}-1\}$. For the average case of $dD$ nonzero elements per row, 
-there will be $\binom{dD+k-1}{dD-1} = $ polynomial features, so the average time complexity for an $N \times D$ matrix
+there will be $\binom{dD+k-1}{dD-1}$ polynomial features, so the average time complexity for an $N \times D$ matrix
 is $\Theta(Nd^kD^k)$, which means that its time complexity polynomially decreases with respect to the density

### 2015-10-06 paper/paper2.tex
Added description of index-mapping
Added description of index mapping and its derivation, and
generalization to interactions among 3 features.

@@
 
-What is therefore needed is a mapping between column index pairs of $\vec{x}$ and the space ${0, 1, ..., \frac{D^2+D}{2}-1}$
+What is therefore needed is a mapping between row-column index pairs of $\vec{x}$ and the space ${0, 1, ..., \frac{D^2+D}{2}-1}$
 (the output size of second degree polynomial features when $D = |\vec{x}|$). The ordering of the mapping is irrelevant so long as its input
@@
 
-To construct this mapping, consider a matrix of size $D \times D$ where cell $(a,b)$ represents the
-polynomial feature between indices $a$ and $b$ for $a,b \in I$ for vector $\vec{x}$.
+Interaction features are a way of capturing correlations between features in a machine 
+learning setting. A feature vector $\vec{x}$ of dimensionality $D$ has second degree interaction features 
+$\{x_i \cdot x_j : i, j \in \{0,1,..., D-1\} \land i < j\}$, 
+so a $D$ dimensional vector has $\binom{D}{2} = \frac{D^2-D}{2}$ second degree interaction features. A naive
+approach to calculating these features is to simply iterate through the combinations of the column indices.
+For a sparse vector, many of the resulting interaction features would be zero, and could therefore be ignored.
+This work describes a method to efficiently calculate second degree interaction features for a sparse matrix 
+that has time and space complexities that decrease quadratically with the density of the input matrix with respect to the naive approach.
 
-\renewcommand{\kbldelim}{(}% Left delimiter
-\renewcommand{\kbrdelim}{)}% Right delimiter
-\[
-  \kbordermatrix{
-             & x_0    & x_1   & \cdot   & \cdot   &  \cdot  &  x_{D-1} \\
-     x_0     & 0      & 1     & \cdot   &  \cdot  &  \cdot  &  D-1     \\
-     x_1     & -      & D     & \cdot   &  \cdot  &  \cdot  &  2D-2    \\
-     \cdot   & \cdot  & \cdot &  \cdot  &      &      &  \cdot       \\
-     \cdot   & \cdot  &       &  \cdot  &  \cdot  &      &  \cdot       \\
-     \cdot   & \cdot  &       &         &  \cdot  &  \cdot  &  \cdot       \\
-    x_{D-1}  & -      & \cdot &  \cdot  & \cdot   & -    &  \frac{D^2+D}{2}-1\\
-  }
-\]
+\section{Approach}
+Let the list of nonzero columns for a given row $\vec{x}$ be denoted by $N_{zc}$. The nonzero second degree 
+interaction features are simply the products of all combinations of two elements whose 
+columns are in $N_{zc}$. However, to properly place an interaction feature into the correct column, a mapping from the column 
+index pairs of $N_{zc}$ into the columns of the interaction matrix is needed. The mapping is 
+from pairs $(a, b)$,  where $a$ and $b$ are in $1,2,..., D$, and $a < b$, to $1,2,..., \frac{D^2-D}{2}$,  Such a mapping essentially consists of mapping the indices of entries in the upper triangle of a matrix to indices in a flat
+list. We now describe the construction of such a mapping.
 
-Again, any bijective mapping $I \mapsto O$ would suffice, but this particular mapping could
-be considered one of the more natural options as it maps $I$ onto $O$ where both $I$ and $O$ are sorted in ascending order.
-As can be seen, the required mapping can be viewed as mapping the upper triangular matrix (including 
-the diagonal), whose indicies are $I$, onto $O$. The construction of this mapping will now
-be given.
+\subsection{Mapping construction}
+\textbf{TODO: convert all $i$ and $j$ indices into $a$ and $b$.--jfh}
 
-INSERT JOHN'S PROOF HERE
+We seek a map from matrix indices $(i, j)$ (with $i < j$ and $0 \le i < D$) to numbers $f(i, j)$ with $0 \le f(i, j) < \frac{D(D-1)}{2}$, one that follows the pattern indicated by 
+\begin{align}
+\begin{bmatrix}
+x & 0 & 1 & 3 \\
+x & x & 2 & 4 \\
+x & x & x & 5 \\
+x & x & x & x
+\end{bmatrix}
+\label{eq:4x4mat}
+\end{align}
+where the entry at location $(i,j)$ displays the value $f(i, j)$.
 
-The mapping can be modified slightly to yield interaction features instead of polynomial features
-by noting that the section of the $D \times D$ mapping matrix that will yield interaction
-features differs only in that its diagonal is excluded. The mapping is therefore gotten by
-taking the polinomial mapping and subtracting $i+1$.
+To simplify slightly, we introduce a notation for the $n$th triangular number, 
+\begin{align}
+T_2(n) = \frac{n(n+1)}{2}
+\end{align}
+\noindent
+The subscript $2$ is there to indicate that these are triangles in two dimensions; we'll use $T_3(n)$ to indicate the $n$th tetrahedral number, and so on for higher dimensions.
+
+Observe that in Equation~\ref{eq:4x4mat}, each entry in column $j$ lies in the range
+\begin{align}
+T_2(j-1) &\le e < T_2(j).
+\end{align}
+\noindent
+And in fact, the entry in the $i$th row of that column is just $i + T_2(j-1)$. Thus we have
+\begin{align}
+f(i, j) 
+&= i + T_2(j-1)\\
+&= i + \frac{(j-1)j}{2}\\
+&=  \frac{2i + j^2-j}{2}.
+\end{align}
+
+For instance, in column $j = 2$ in our example (the \emph{third} column), the entries range from $1$ to $2$, while $T_2(j-1) = T_2(1) = 1$ and $T_2(j) = T_2(2) = 3$, and the entry in column $j = 2$, row $i = 1$ is
+$i + T_2(j-1) = 1 + 1 = 2$.
+
+\subsubsection{Other indices}
+With one-based indexing in both the domain and codomain, the formula above becomes
+\begin{align}
+f_1(i, j) &= 1+ f(i-1, j-1) \\
+&= 1+ f(i-1, j-1) \\
+& = \frac{2 + 2(i-1) + (j-1)^2-(j-1)}{2}\\
+& = \frac{2i + j^2 - 3j + 2}{2}
+\end{align}
+\subsubsection{Higher dimensions}
+To handle three-way interactions, we need to map triples of indices in a 3-index array to a flat list, and similarly for higher-order interactions.
+
+For three indices, $i,j,k$, with $i < j < k$ and $0 \le i,j,k < D$, we have a similar recurrence. Calling the mapping $g$, we have 
+\begin{align}
+g(i,j,k) = i + T_2(j-1) + T_3(k-1);
+\end{align}
+if we define $T_1(i) = i$, then this has the very regular form
+\begin{align}
+g(i,j,k) = 1 + T_1(i-1) + T_2(j-1) + T_3(k-1);
+\end{align}
+and from this, the generalization to higher dimensions is straightforward. The formulas for ``higher triangular numbers'', i.e., those defined by
+\begin{align}
+T_k(n) &= \sum_{i=1}^n T_{k-1}(n)
+\end{align}
+for $k > 1$ can be determined inductively. For $k = 3$, the result is 
+\begin{align}
+T_3(n) &= \sum_{i=1}^n T_{2}(n)\\
+&= \frac{n^3 + 3n^2 + 2n}{6},
+\end{align}
+so that the formula for 3-way interactions, with zero-based indexing, becomes 
+\begin{align}
+g(i, j, k) &= 1 + (i-1) + \frac{(j-1)j}{2} + \\
+& \frac{(k-1)^3 + 3(k-1)^2 + 2(k-1)}{6}. 
+\end{align}
+\subsection{Inversion}
+We not only want to be able to compute the ``flat index'' $f(i,j)$ corresponding to a particular index pair $(i, j)$, we'd like, given a value $N = f(i, j)$, to be able to determine $i$ and $j$.
+We know, given $N$, that $f(0, j) \le N \le f(j,j-1)$, i.e., that
+\begin{align}
+\frac{j^2 - 3j + 2}{2} &\le N &\le \frac{2(j-1) + j^2 - 3j + 2}{2}\\
+{j^2 - 3j + 2} &\le 2N &\le {j^2 - j}.
+\end{align}
+If we find the (real number) value $j_{*}$ for which $j_{*}^2 - j_{*} = 2N$, then we know that $j = \lfloor j_{*}\rfloor$. Finding $j_{*}$ amounts to solving a quadratic, whence
+\begin{align}
+j_{*} = \frac{1 \pm \sqrt{1 + 8N}}{2}.
+\end{align}
+Clearly the larger value, 
+\begin{align}
+j_{*} = \frac{1 + \sqrt{1 + 8N}}{2},
+\end{align}
+is the one we seek; we set $j = \lfloor j_{*}\rfloor$, and compute $i = N - \frac{j^2 - j}{2}$.
+
+In practice, repeatedly computing square roots may be annoying; instead, when we compute $N = f(i,j)$, we can record the values of $i$ and $j$ along with the interaction features, making inversion a simple matter of looking up these values.
+
+%To construct this mapping, consider a matrix of size $D \times D$ where %cell $(a,b)$ represents the
+%polynomial feature between indices $a$ and $b$ for $a,b \in I$ for vector $\vec{x}$.
+
+%\renewcommand{\kbldelim}{(}% Left delimiter
+%\renewcommand{\kbrdelim}{)}% Right delimiter
+%\[
+%  \kbordermatrix{
+%             & x_0    & x_1   & \cdot   & \cdot   &  \cdot  &  x_{D-1} \\
+%     x_0     & 0      & 1     & \cdot   &  \cdot  &  \cdot  &  D-1     \\
+%     x_1     & -      & D     & \cdot   &  \cdot  &  \cdot  &  2D-2    \\
+%     \cdot   & \cdot  & \cdot &  \cdot  &      &      &  \cdot       \\
+%     \cdot   & \cdot  &       &  \cdot  &  \cdot  &      &  \cdot       %\\
+%     \cdot   & \cdot  &       &         &  \cdot  &  \cdot  &  \cdot       %\\
+%    x_{D-1}  & -      & \cdot &  \cdot  & \cdot   & -    &  \frac{D^2+D}%{2}-1\\
+  %}
+%\]
+
+%Again, any bijective mapping $I \mapsto O$ would suffice, but this %particular mapping could
+%be considered one of the more natural options as it maps $I$ onto $O$ %where both $I$ and $O$ are sorted in ascending order.
+%As can be seen, the required mapping can be viewed as mapping the upper %triangular matrix (including 
+%the diagonal), whose indicies are $I$, onto $O$. The construction of this %mapping will now
+%be given.
+
+%The mapping can be modified slightly to yield interaction features %instead of polynomial features
+%by noting that the section of the $D \times D$ mapping matrix that will %yield interaction
+%features differs only in that its diagonal is excluded. The mapping is %therefore gotten by
+%taking the polinomial mapping and subtracting $i+1$.
 

### 2015-10-07 paper/spares_interaction.tex
fixed an index in the algo

@@
 index pairs of $N_{zc}$ into the columns of the interaction matrix is needed. The mapping is 
-from the space (a, b) where a,b are in $1,2,..., D$ onto the space $1,2,..., \frac{D^2-D}{2}$. This 
-is isomorphic to mapping the coordinates of the upper triangle of a matrix onto a flat 
-list. The following is a proof by construction for such a mapping.
+from pairs $(a, b)$,  where $a$ and $b$ are in $1,2,..., D$, and $a < b$, to $1,2,..., \frac{D^2-D}{2}$,  Such a mapping essentially consists of mapping the indices of entries in the upper triangle of a matrix to indices in a flat
+list. We now describe the construction of such a mapping.
+
+\subsection{Mapping construction}
+We seek a map from matrix indices $(i, j)$ (with $i < j$ and $0 \le i < D$) to numbers $f(i, j)$ with $0 \le f(i, j) < \frac{D(D-1)}{2}$, one that follows the pattern indicated by 
+\begin{align}
+\begin{bmatrix}
+x & 0 & 1 & 2 \\
+x & x & 3 & 4 \\
+x & x & x & 5 \\
+x & x & x & x
+\end{bmatrix}
+\end{align}
+It's considerably easier, however, to consider the same indices, but subtracted from $6$ (or more generally, from $\frac{D(D-1)}{2}$; that gives the pattern
+\begin{align}
+\label{eq:4x4mat}
+\begin{bmatrix}
+x & 6 & 5 & 4 \\
+x & x & 3 & 2 \\
+x & x & x & 1 \\
+x & x & x & x
+\end{bmatrix}
+\end{align}
+We'll call the function defined by this example $(i, j) \mapsto g(i, j)$, and then observe that 
+\begin{align}
+f(i, j) = \frac{D(D-1)}{2} - g(i, j)
+\end{align}
+
+To simplify slightly, we introduce a notation for the $n$th triangular number, 
+\begin{align}
+\begin{bmatrix}
+T_2(n) = \frac{n(n+1)}{2}
+\end{bmatrix}
+\end{align}
+\noindent
+The subscript $2$ is there to indicate that these are triangles in two dimensions; we'll use $T_3(n)$ to indicate the $n$th tetrahedral number, and so on for higher dimensions.
+
+The codomain of $g$ is now numbers from $1$ to $T_2(D-1)$, inclusive.
+
+Observe that in Equation~\ref{eq:4x4mat}, each entry in row $i$ lies in the range
+\begin{align}
+T_2(D-i-1) &< e \le T_2(D-i).
+\end{align}
+\noindent
+For instance, in row $2$ in our example, where $D = 4$, the entries range from $2$ to $3$, while $T_2(D-i-1) = T_2(1) = 1$ and $T_2(D-i) = T_2(2) = 3$.
+(Note that row indices start at zero.) Unfortunately, the numbers increase from right to left.
+The entry in column $j$ is just $T_2(D-i-1) + D-j$, which adds one for the rightmost column (because $D - (D-1) = 1$).
+Thus, the formula for $g$ is simply
+\begin{align}
+g(i, j) 
+&= T_2(D-i-1) + (D-j) \\
+&= \frac{(D-i-1)(D-i)}{2} + D-j \\
+&= \frac{(D^2 - (2i)D - D - i^2 - i) + 2D - 2j}{2} \\
+&= \frac{D^2 - (2i)D + D - i^2 - i - 2j}{2}
+\end{align}
+and hence 
+\begin{align}
+f(i, j) 
+&= \frac{D(D-1)}{2} - g(i,j) \\
+&= \frac{D^2-D}{2} - \frac{D^2 - (2i)D + D - i^2 - i - 2j}{2}\\
+&= \frac{D^2-D - D^2 + (2i)D - D + i^2 + i + 2j}{2}\\
+&= \frac{(2i)D - 2D + i^2 + i + 2j}{2}
+\end{align}
+[WRONG]
+Correct result:
+\begin{align}
+R = \frac{ 2in - i^2  + 2j - 3i - 2}{2}.
+\end{align}
+\subsubsection{Other indices}
+With one-based indexing, the formula above becomes
+\begin{align}
+f_1(i, j) &= ....
+\end{align}
+
+  column
 
@@
     \zi     $N_{zc} \gets$ nonzero columns of $row$
-    \zi     \For $i \gets 0 \To |N_{zc}|-2$ \Do
-    \zi         \For $j \gets i+1 \To |N_{zc}|-1$ \Do
+    \zi     \For $i \gets 0 \To |N_{zc}|-1$ \Do
+    \zi         \For $j \gets i+1 \To |N_{zc}|$ \Do
     \zi             $k \gets \func{map}(i, j)$

### 2015-12-03 paper/paper2.tex

@@
 index pairs of $N_{zc}$ into the columns of the interaction matrix is needed. The mapping is 
-from pairs $(a, b)$,  where $a$ and $b$ are in $1,2,..., D$, and $a < b$, to $1,2,..., \frac{D^2-D}{2}$,  Such a mapping essentially consists of mapping the indices of entries in the upper triangle of a matrix to indices in a flat
+from pairs $(i, j)$,  where $i$ and $j$ are in $0, 1, \ldots , D-1$, and $i < j$, to $0, 1,2,..., \frac{D^2-D}{2} - 1$,  Such a mapping essentially consists of mapping the indices of entries in the upper triangle of a matrix to indices in a flat
 list. We now describe the construction of such a mapping.
@@
 \subsection{Mapping construction}
-\textbf{TODO: convert all $i$ and $j$ indices into $a$ and $b$.--jfh}
 
@@
 \end{align}
-where the entry at location $(i,j)$ displays the value $f(i, j)$.
+where the entry in row $i$, column $j$, displays the value $f(i, j)$.
 
@@
 \noindent
-The subscript $2$ is there to indicate that these are triangles in two dimensions; we'll use $T_3(n)$ to indicate the $n$th tetrahedral number, and so on for higher dimensions.
+The subscript $2$ indicates that these are triangles in two dimensions; we'll use $T_3(n)$ to indicate the $n$th tetrahedral number, and so on for higher dimensions.
 
-Observe that in Equation~\ref{eq:4x4mat}, each entry in column $j$ lies in the range
+Observe that in Equation~\ref{eq:4x4mat}, each entry in column $j$ (for $j > 0$) lies in the range
 \begin{align}
@@
 \end{align}
+
+\subsubsection{Polynomial Features}
+For polynommial features, we seek a map from matrix indices $(i, j)$ (with $i \le j$ and $0 \le i < D$) to numbers $g(i, j)$ with $0 \le f(i, j) < \frac{D(D+1)}{2}$, one that follows the pattern indicated by
+\begin{align}
+\begin{bmatrix}
+ 0 & 1 & 3 & 6 \\
+ x & 2 & 4 & 7\\
+ x & x & 5 & 8 \\
+ x & x & x & 9
+\end{bmatrix}
+\label{eq:4x4mat-poly}
+\end{align}
+i.e., essentially the same task as before, except that the diagonal is included.
+One can regard all but the last column of entries in Equation~\ref{eq:4x4mat-poly} as corresponding to the entries in Equation~\ref{eq:4x4mat}, but shifted to the left.
+Thus the formula for $g(i, j)$ is simply the formula for $f$, shifted by 1, i.e.,
+\begin{align}
+g(i, j) &= f(i, j+1)  \\
+&=  \frac{2i + (j+1)^2-(j+1)}{2}\\
+&=  \frac{2i + j^2 + j + 1)}{2}.
+\end{align}
+Alternatively, we can write this as
+\begin{align}
+g(i, j) &= i + T_2(j),
+\end{align}
+\noindent
+and get the same result.
+
 \subsubsection{Higher dimensions}
@@
 
-For three indices, $i,j,k$, with $i < j < k$ and $0 \le i,j,k < D$, we have a similar recurrence. Calling the mapping $g$, we have 
+For three indices, $i,j,k$, with $i < j < k$ and $0 \le i,j,k < D$, we have a similar recurrence. Calling the mapping $h$, we have 
 \begin{align}
-g(i,j,k) = i + T_2(j-1) + T_3(k-1);
+h(i,j,k) = i + T_2(j-1) + T_3(k-2);
 \end{align}
@@
 \begin{align}
-g(i,j,k) = 1 + T_1(i-1) + T_2(j-1) + T_3(k-1);
+h(i,j,k) =  T_1(i) + T_2(j-1) + T_3(k-2);
 \end{align}
@@
 \begin{align}
-g(i, j, k) &= 1 + (i-1) + \frac{(j-1)j}{2} + \\
-& \frac{(k-1)^3 + 3(k-1)^2 + 2(k-1)}{6}. 
+h(i, j, k) &= 1 + (i-1) + \frac{(j-1)j}{2} + \\
+& \frac{(k-2)^3 + 3(k-2)^2 + 2(k-2)}{6}. 
 \end{align}
+\subsubsection{Higher-dimension polynomial features}
+For the case where we include the diagonal in higher dimensions, we must shift $j$ by $1$, $k$ by $2$, and so on, and the formula becomes
+\begin{align}
+\ell(i,j,k) =  T_1(i) + T_2(j) + T_3(k),
+\end{align}
+with analogous formulas for higher degree polynomial interactions.
+
 \subsection{Inversion}
@@
 \begin{align}
-\frac{j^2 - 3j + 2}{2} &\le N &\le \frac{2(j-1) + j^2 - 3j + 2}{2}\\
-{j^2 - 3j + 2} &\le 2N &\le {j^2 - j}.
+\frac{j^2 - 3j + 2}{2} &\le N \le \frac{2(j-1) + j^2 - 3j + 2}{2}\\
+{j^2 - 3j + 2} &\le 2N \le {j^2 - j}.
 \end{align}
@@
 is the one we seek; we set $j = \lfloor j_{*}\rfloor$, and compute $i = N - \frac{j^2 - j}{2}$. 
-
 In practice, repeatedly computing square roots may be annoying; instead, when we compute $N = f(i,j)$, we can record the values of $i$ and $j$ along with the interaction features, making inversion a simple matter of looking up these values.
 
-%To construct this mapping, consider a matrix of size $D \times D$ where %cell $(a,b)$ represents the
-%polynomial feature between indices $a$ and $b$ for $a,b \in I$ for vector $\vec{x}$.
+Similarly, for polynomial features, to invert 
+\begin{align}
+g(i, j) &=  \frac{2i + j^2 + j + 1)}{2}
+\end{align}
+we know that if $N = g(i,j)$, then $g(0,j) \le N \le g(j,j)$, which tells us that 
+\begin{align}
+2N \le j^2 + 3j + 1, \text{ so}\\
+0 \le j^2 + 3j + (1-N^2)
+\end{align}. 
+Solving for $j$ when this last inequality is an equality, we get 
+\begin{align}
+j_{*} = \frac{-3 \pm \sqrt{ 5 - N^2 }}{2}; 
+\end{align}
+Once again, we set $j = \lfloor j_{*}\rfloor$, and compute $i = N - \frac{j^2 + j + 1}{2}$.
 
-%\renewcommand{\kbldelim}{(}% Left delimiter
-%\renewcommand{\kbrdelim}{)}% Right delimiter
-%\[
-%  \kbordermatrix{
-%             & x_0    & x_1   & \cdot   & \cdot   &  \cdot  &  x_{D-1} \\
-%     x_0     & 0      & 1     & \cdot   &  \cdot  &  \cdot  &  D-1     \\
-%     x_1     & -      & D     & \cdot   &  \cdot  &  \cdot  &  2D-2    \\
-%     \cdot   & \cdot  & \cdot &  \cdot  &      &      &  \cdot       \\
-%     \cdot   & \cdot  &       &  \cdot  &  \cdot  &      &  \cdot       %\\
-%     \cdot   & \cdot  &       &         &  \cdot  &  \cdot  &  \cdot       %\\
-%    x_{D-1}  & -      & \cdot &  \cdot  & \cdot   & -    &  \frac{D^2+D}%{2}-1\\
-  %}
-%\]
-
-%Again, any bijective mapping $I \mapsto O$ would suffice, but this %particular mapping could
-%be considered one of the more natural options as it maps $I$ onto $O$ %where both $I$ and $O$ are sorted in ascending order.
-%As can be seen, the required mapping can be viewed as mapping the upper %triangular matrix (including 
-%the diagonal), whose indicies are $I$, onto $O$. The construction of this %mapping will now
-%be given.
-
-%The mapping can be modified slightly to yield interaction features %instead of polynomial features
-%by noting that the section of the $D \times D$ mapping matrix that will %yield interaction
-%features differs only in that its diagonal is excluded. The mapping is %therefore gotten by
-%taking the polinomial mapping and subtracting $i+1$.
+Inverting the higher-order formulas, which involve cubics, quartics, etc., is infeasible, but storing the inverse map as described above may be practical, especially for sparse feature sets.
 

### 2016-02-08 new_sparse_poly.py
added 2 and 3 deg poly

@@ -46,12 +46,12 @@ class SparsePolynomialFeatures(TransformerMixin, BaseEstimator):
     def __init__(self, degree, interaction_only=False, include_bias=False):
     
         if degree > 3:
-            raise Exception("SparsePolynomialFeatures currently supports degrees < 4.")
+            raise Exception("SparsePolynomialFeatures currently supports degrees of 2 or 3.")
             
         self.degree = degree
         self.interaction_only = interaction_only
         self.include_bias = include_bias
-        self.__poly_functs = [] #[self.__second_deg_poly, self.__third_deg_poly]
+        self.__poly_functs = [self.__second_deg_poly, self.__third_deg_poly]
         self.__inter_functs = [self.__second_deg_inter, self.__third_deg_inter]
     
     def fit(self, X, **fit_params):
@@ -61,7 +61,7 @@ class SparsePolynomialFeatures(TransformerMixin, BaseEstimator):
     
         X = X.tocoo() #Put back into original format. Also put the new matrix into the same format.
     
-        #how big will the output matrix be?
+        #How big will the output matrix be?
         N, D = X.shape
     
         #start degree at 2 since we're going to just copy the first degree features
@@ -78,13 +78,15 @@ class SparsePolynomialFeatures(TransformerMixin, BaseEstimator):
         funct_list = self.__inter_functs if self.interaction_only else self.__poly_functs
         
         for size, stop_ind, degree, funct in zip(sizes, stop_inds, xrange(2, self.degree+1), funct_list):
-            print stop_ind, degree, offset            
             funct(X, D, offset, rows, cols, data)
             offset += size
         
-        print 'all done'
-        interact(local=locals())
         return coo_matrix((data, (rows, cols)), shape=(N, D+sum(sizes))).tocsr()
+        
+    """
+    Each polynomial has its own method. I could have put it all into one, but
+    this would cause lots of function calls, which would add some time overhead.
+    """
     
     def __second_deg_inter(self, X, D, offset, rows, cols, data):
         #cur_ind will be used to fill in the correct array elements for when rows, cols, and data are arrays.
@@ -97,14 +99,11 @@ class SparsePolynomialFeatures(TransformerMixin, BaseEstimator):
                 for j in xrange(i+1, n_zc):
                     col2 = nz_cols[j]
                     output_col = (2*col1 + col2**2 - col2)/2
-                    if output_col == (D**2-D)/2+D:
[…]

### 2016-02-08 new_sparse_poly.py
added 2 and 3 deg poly

@@ -5,42 +5,7 @@ from scipy.misc import comb
 from scipy.sparse import csr_matrix, coo_matrix
 from code import interact
 
-"""
-For John:
-
-The new mappings don't contain D.
-Could you aso do first and second degrees to include the diagonal?
-
-
-"""
-
-"""
-1. Create the feature matrix, include the original matrix in it - the rest will start out all zeros.
-2. Iterate with as many nested loops as needed. This is the part that will have a different method for each degree.
-"""
-
-
-"""
-
-(0, 1) = 0
-(0, 2) = 1
-(1, 2) = 2
-(0, 3) = 3
-
-x 0 1 3
-x x 2 4
-x x x 5
-x x x x
-
-
-we want
-
-0 1 3 6
-x 2 4 7
-x x 5 8
-x x x 9
-
-"""
+#TODO add include_bias support
 
 class SparsePolynomialFeatures(TransformerMixin, BaseEstimator):
     def __init__(self, degree, interaction_only=False, include_bias=False):

### 2016-02-10 paper/workshop/sparse_poly.tex
workshop paper

\documentclass{article} % For LaTeX2e
\usepackage{iclr2016_workshop,times}
\usepackage{hyperref}
\usepackage{url}

\title{Efficient Calculation of Polynomial Features on Sparse Matrices}

\author{Andrew Nystrom \\
Savvysherpa Inc.\\
6200 Shingle Creek Pkwy \\
Suite 400 \\
Minneapolis, MN 55430, USA \\
\texttt{awnystrom@gmail.com} \\
\And
John Hughes \\
Department of Computational Neuroscience \\
University of the Witwatersrand \\
Joburg, South Africa \\
\texttt{\{robot,net\}@wits.ac.za} \\
}

% The \author macro works with any number of authors. There are two commands
% used to separate the names and addresses of multiple authors: \And and \AND.
%
% Using \And between authors leaves it to \LaTeX{} to determine where to break
% the lines. Using \AND forces a linebreak at that point. So, if \LaTeX{}
% puts 3 of 4 authors names on the first line, and the last on the second
% line, try using \AND instead of \And before the third author name.

\newcommand{\fix}{\marginpar{FIX}}
\newcommand{\new}{\marginpar{NEW}}

\begin{document}

\maketitle

\begin{abstract}
We provide an algorithm for polynomial feature expansion that can operate directly on sparse a
matrix of size $NxD$ with $\O(d^kD^kN)$ where $k$ is the polynomial order and $d$ is the density of each row.

\end{abstract}

\section{Introduction}

Polynomial feature expansion is a tool long used in statistics for approximating nonlinear functions (see \cite{gergonne1974application, smith1918standard}).
While their use has long been widespread, the authors are unaware of any improvements made to their calculation efficiency.
In this work we provide an algorithm for calculating polynomial features for a matrix of size matrix of size $NxD$ with $O(d^kD^kN)$ where $k$ is the polynomial order and $d$ is the density of each row.
The density of a matrix is the percent of its nonzero items, so $0 \le d \le 1$. The standard algorithm is $O(D^kN)$, so the added factor of $d^k$ represents a significant reduction in time.
The algorithm does not require the densification of the matrix, e.i. the matrix remains in sparse form, so the space complexity is also $O(d^kD^kN)$ as opposed to $O(D^kN)$.

\section{Algorithm}
The standard method of calculating polynomial features for a vector $\vec{x}$ involves augmenting a feature for the product of each combination of features in $\vec{x}$ of orders 2 to $k$.
This method does not exploit the sparsity of a sparse matrix and will yield a product of zero any time one of the features involved in the product is zero.
In a sparse matrix, such products will be common, and since the default value of a sparse matrix is zero, these products are entirely unnecessary to compute.

The main idea behind our algorithm is to only compute products that do not involve zeros.
In a compressed sparse row matrix, the columns containing nonzero data are the only columns that are stored.
We can therefore iterate over products of combinations of orders 2 to $k$ of these columns for each row to calculate $k$-degree polynomial features.

While the idea is straightforward, there is yet one unaddressed caveat; 
Given a set of nonzero columns whose product was just calculated to generate a polynomial feature, where in the polynomial matrix does the result of the product belong?
To adress this, we derive a bijective mapping from the set of possible column index combinations of order 2 to $k$ onto the column index space of the polynomial feature matrix.
More precisely, this the mapping is of the form $(i_1, i_1, ..., i_k) \arrow {} where i_1 \le i_1 \le ... \le i_k$.

\subsection{Derivation of Mappings}

\section{Analysis}
\subsection{Analytical}
\subsection{Empirical}
\section{Conclusion}

\section{Citations, figures, tables, references}
\label{others}

These instructions apply to everyone, regardless of the formatter being used.

\subsection{Citations within the text}

Citations within the text should be based on the {\tt natbib} package
and include the authors' last names and year (with the ``et~al.'' construct
for more than two authors). When the authors or the publication are
included in the sentence, the citation should not be in parenthesis (as
in ``See \citet{Hinton06} for more information.''). Otherwise, the citation
should be in parenthesis (as in ``Deep learning shows promise to make progress towards AI~\citep{Bengio+chapter2007}.'').

The corresponding references are to be listed in alphabetical order of
authors, in the \textsc{References} section. As to the format of the
references themselves, any style is acceptable as long as it is used
consistently.

\subsection{Footnotes}

Indicate footnotes with a number\footnote{Sample of the first footnote} in the
text. Place the footnotes at the bottom of the page on which they appear.
Precede the footnote with a horizontal rule of 2~inches
(12~picas).\footnote{Sample of the second footnote}

\subsection{Figures}

All artwork must be neat, clean, and legible. Lines should be dark
enough for purposes of reproduction; art work should not be
hand-drawn. The figure number and caption always appear after the
figure. Place one line space before the figure caption, and one line
space after the figure. The figure caption is lower case (except for
first word and proper nouns); figures are numbered consecutively.

Make sure the figure caption does not get separated from the figure.
Leave sufficient space to avoid splitting the figure and figure caption.

You may use color figures.
However, it is best for the
figure captions and the paper body to make sense if the paper is printed
either in black/white or in color.
\begin{figure}[h]
\begin{center}
%\framebox[4.0in]{$\;$}
\fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
\end{center}
\caption{Sample figure caption.}
\end{figure}

\subsection{Tables}

All tables must be centered, neat, clean and legible. Do not use hand-drawn
tables. The table number and title always appear before the table. See
Table~\ref{sample-table}.

Place one line space before the table title, one line space after the table
title, and one line space after the table. The table title must be lower case
(except for first word and proper nouns); tables are numbered consecutively.

\begin{table}[t]
\caption{Sample table title}
\label{sample-table}
\begin{center}
\begin{tabular}{ll}
\multicolumn{1}{c}{\bf PART}  &\multicolumn{1}{c}{\bf DESCRIPTION}
\\ \hline \\
Dendrite         &Input terminal \\
Axon             &Output terminal \\
Soma             &Cell body (contains cell nucleus) \\
\end{tabular}
\end{center}
\end{table}

\section{Final instructions}
Do not change any aspects of the formatting parameters in the style files.
In particular, do not modify the width or length of the rectangle the text
should fit into, and do not change font sizes (except perhaps in the
\textsc{References} section; see below). Please note that pages should be
numbered.

\section{Preparing PostScript or PDF files}

Please prepare PostScript or PDF files with paper size ``US Letter'', and
not, for example, ``A4''. The -t
letter option on dvips will produce US Letter files.

Consider directly generating PDF files using \verb+pdflatex+
(especially if you are a MiKTeX user).
PDF figures must be substituted for EPS figures, however.

Otherwise, please generate your PostScript and PDF files with the following commands:
\begin{verbatim}
dvips mypaper.dvi -t letter -Ppdf -G0 -o mypaper.ps
ps2pdf mypaper.ps mypaper.pdf
\end{verbatim}

\subsection{Margins in LaTeX}

Most of the margin problems come from figures positioned by hand using
\verb+\special+ or other commands. We suggest using the command
\verb+\includegraphics+
from the graphicx package. Always specify the figure width as a multiple of
the line width as in the example below using .eps graphics
\begin{verbatim}
   \usepackage[dvips]{graphicx} ...
   \includegraphics[width=0.8\linewidth]{myfile.eps}
\end{verbatim}
or % Apr 2009 addition
\begin{verbatim}
   \usepackage[pdftex]{graphicx} ...
   \includegraphics[width=0.8\linewidth]{myfile.pdf}
\end{verbatim}
for .pdf graphics.
See section 4.4 in the graphics bundle documentation (\url{http://www.ctan.org/tex-archive/macros/latex/required/graphics/grfguide.ps})

A number of width problems arise when LaTeX cannot properly hyphenate a
line. Please give LaTeX hyphenation hints using the \verb+\-+ command.

\subsubsection*{Acknowledgments}

Use unnumbered third level headings for the acknowledgments. All
acknowledgments, including those to funding agencies, go at the end of the paper.

\bibliography{sparse_poly}
\bibliographystyle{iclr2016_workshop}

\end{document}

### 2016-02-10 paper/workshop/build.sh

@@ -1,0 +1,4 @@
+pdflatex sparse_poly
+bibtex sparse_poly
+pdflatex sparse_poly
+pdflatex sparse_poly

### 2016-02-12 paper/workshop/sparse_poly.tex
init workshop paper

@@
 \usepackage{url}
+
+\usepackage{amsmath}
+\usepackage{mathtools}
 
@@
 We provide an algorithm for polynomial feature expansion that can operate directly on sparse a
-matrix of size $NxD$ with $\O(d^kD^kN)$ where $k$ is the polynomial order and $d$ is the density of each row.
+matrix of size $NxD$ with $O(d^kD^kN)$ where $k$ is the polynomial order and $d$ is the density of each row.
 
@@
 Given a set of nonzero columns whose product was just calculated to generate a polynomial feature, where in the polynomial matrix does the result of the product belong?
-To adress this, we derive a bijective mapping from the set of possible column index combinations of order 2 to $k$ onto the column index space of the polynomial feature matrix.
-More precisely, this the mapping is of the form $(i_1, i_1, ..., i_k) \arrow {} where i_1 \le i_1 \le ... \le i_k$.
+To adress this, we give a bijective mapping from the set of possible column index combinations of order 2 to $k$ onto the column index space of the polynomial feature matrix.
+More precisely, this the mapping is of the form
 
-\subsection{Derivation of Mappings}
+$(i_0, i_1, \dots, i_{k-1}) \mapsto {0,1,\dots, \binom{D}{k}}$ 
+$\forall i_0, i_1, \dots, i_{k-1} \mid 0 \le i_0 \le i_1 \le \dots \le i_{k-1} \le D$
+
+where $i_0,i_1,\dots i_{k-1}$ are column indicies.
+
+\subsection{Construction of Mappings}
+
+We seek a map from matrix indices $(i, j)$ (with $i < j$ and $0 \le i < D$) to numbers $f(i, j)$ with $0 \le f(i, j) < \frac{D(D-1)}{2}$, one that follows the pattern indicated by 
+\begin{align}
+\begin{bmatrix}
+x & 0 & 1 & 3 \\
+x & x & 2 & 4 \\
+x & x & x & 5 \\
+x & x & x & x
+\end{bmatrix}
+\label{eq:4x4mat}
+\end{align}
+where the entry in row $i$, column $j$, displays the value $f(i, j)$.
+
+To simplify slightly, we introduce a notation for the $n$th triangular number, 
+\begin{align}
+T_2(n) = \frac{n(n+1)}{2}
+\end{align}
+\noindent
+The subscript $2$ indicates that these are triangles in two dimensions; we'll use $T_3(n)$ to indicate the $n$th tetrahedral number, and so on for higher dimensions.
+
+Observe that in Equation~\ref{eq:4x4mat}, each entry in column $j$ (for $j > 0$) lies in the range
+\begin{align}
+T_2(j-1) &\le e < T_2(j).
+\end{align}
+\noindent
+And in fact, the entry in the $i$th row of that column is just $i + T_2(j-1)$. Thus we have
+\begin{align}
+f(i, j) 
+&= i + T_2(j-1)\\
+&= i + \frac{(j-1)j}{2}\\
+&=  \frac{2i + j^2-j}{2}.
+\end{align}
+
+For instance, in column $j = 2$ in our example (the \emph{third} column), the entries range from $1$ to $2$, while $T_2(j-1) = T_2(1) = 1$ and $T_2(j) = T_2(2) = 3$, and the entry in column $j = 2$, row $i = 1$ is
+$i + T_2(j-1) = 1 + 1 = 2$.
+
+\subsubsection{Other indices}
+With one-based indexing in both the domain and codomain, the formula above becomes
+\begin{align}
+f_1(i, j) &= 1+ f(i-1, j-1) \\
+&= 1+ f(i-1, j-1) \\
+& = \frac{2 + 2(i-1) + (j-1)^2-(j-1)}{2}\\
+& = \frac{2i + j^2 - 3j + 2}{2}
+\end{align}
+
+\subsubsection{Polynomial Features}
+For polynommial features, we seek a map from matrix indices $(i, j)$ (with $i \le j$ and $0 \le i < D$) to numbers $g(i, j)$ with $0 \le f(i, j) < \frac{D(D+1)}{2}$, one that follows the pattern indicated by
+\begin{align}
+\begin{bmatrix}
+ 0 & 1 & 3 & 6 \\
+ x & 2 & 4 & 7\\
+ x & x & 5 & 8 \\
+ x & x & x & 9
+\end{bmatrix}
+\label{eq:4x4mat-poly}
+\end{align}
+i.e., essentially the same task as before, except that the diagonal is included.
+One can regard all but the last column of entries in Equation~\ref{eq:4x4mat-poly} as corresponding to the entries in Equation~\ref{eq:4x4mat}, but shifted to the left.
+Thus the formula for $g(i, j)$ is simply the formula for $f$, shifted by 1, i.e.,
+\begin{align}
+g(i, j) &= f(i, j+1)  \\
+&=  \frac{2i + (j+1)^2-(j+1)}{2}\\
+&=  \frac{2i + j^2 + j + 1)}{2}.
+\end{align}
+Alternatively, we can write this as
+\begin{align}
+g(i, j) &= i + T_2(j),
+\end{align}
+\noindent
+and get the same result.
+
+\subsubsection{Higher dimensions}
+To handle three-way interactions, we need to map triples of indices in a 3-index array to a flat list, and similarly for higher-order interactions.
+
+For three indices, $i,j,k$, with $i < j < k$ and $0 \le i,j,k < D$, we have a similar recurrence. Calling the mapping $h$, we have 
+\begin{align}
+h(i,j,k) = i + T_2(j-1) + T_3(k-2);
+\end{align}
+if we define $T_1(i) = i$, then this has the very regular form
+\begin{align}
+h(i,j,k) =  T_1(i) + T_2(j-1) + T_3(k-2);
+\end{align}
+and from this, the generalization to higher dimensions is straightforward. The formulas for ``higher triangular numbers'', i.e., those defined by
+\begin{align}
+T_k(n) &= \sum_{i=1}^n T_{k-1}(n)
+\end{align}
+for $k > 1$ can be determined inductively. For $k = 3$, the result is 
+\begin{align}
+T_3(n) &= \sum_{i=1}^n T_{2}(n)\\
+&= \frac{n^3 + 3n^2 + 2n}{6},
+\end{align}
+so that the formula for 3-way interactions, with zero-based indexing, becomes 
+\begin{align}
+h(i, j, k) &= 1 + (i-1) + \frac{(j-1)j}{2} + \\
+& \frac{(k-2)^3 + 3(k-2)^2 + 2(k-2)}{6}. 
+\end{align}
+\subsubsection{Higher-dimension polynomial features}
+For the case where we include the diagonal in higher dimensions, we must shift $j$ by $1$, $k$ by $2$, and so on, and the formula becomes
+\begin{align}
+\ell(i,j,k) =  T_1(i) + T_2(j) + T_3(k),
+\end{align}
+with analogous formulas for higher degree polynomial interactions.
 

### 2016-02-12 paper/workshop/sparse_poly.tex
fixed indices in mapping definition
dimensionality in abstract

@@
 \usepackage{mathtools}
+\usepackage{amssymb}
 
@@
 We provide an algorithm for polynomial feature expansion that can operate directly on sparse a
-matrix of size $NxD$ with $O(d^kD^kN)$ where $k$ is the polynomial order and $d$ is the density of each row.
+matrix of size $N \times D$ with $O(d^kD^kN)$ where $k$ is the polynomial order and $d$ is the density of each row.
 
@@
 
-$(i_0, i_1, \dots, i_{k-1}) \mapsto {0,1,\dots, \binom{D}{k}}$ 
-$\forall i_0, i_1, \dots, i_{k-1} \mid 0 \le i_0 \le i_1 \le \dots \le i_{k-1} \le D$
-
-where $i_0,i_1,\dots i_{k-1}$ are column indicies.
+\begin{equation}
+(x_{i_0}, x_{i_1}, \dots, x_{i_{k-1}}) \rightarrowtail \hspace{-1.9ex} \twoheadrightarrow p_{i_0i_1 \dots i_{k-1}} \in \{0,1,\dots, \binom{D}{k}\} 
+\forall (i_0, i_1, \dots, i_{k-1})
+\end{equation}
+such that $ 0 \le i_0 \le i_1 \le \dots \le i_{k-1} < D$
+where $(x_{i_0}, x_{i_1}, \dots, x_{i_{k-1}})$ are column indicies of a row vector $\vec{x}$ of an $N \times D$ input matrix and $p_{i_0i_1 \dots i_{k-1}}$ is a column index of the polynomial expansion matrix.
 

### 2016-02-13 paper/workshop/sparse_poly.tex
cleaned up intro

@@
 We provide an algorithm for polynomial feature expansion that can operate directly on sparse a
-matrix of size $N \times D$ with $O(d^kD^kN)$ where $k$ is the polynomial order and $d$ is the density of each row.
+matrix. For a vector $\vec{x}$ of a $N \times D$ matrix, the algorithm has time and space complexity $O(d^kD^k)$ where $k$ is the polynomial order and $d$ is the density of $\vec{X}$.
 
@@
 Polynomial feature expansion is a tool long used in statistics for approximating nonlinear functions (see \cite{gergonne1974application, smith1918standard}).
-While their use has long been widespread, the authors are unaware of any improvements made to their calculation efficiency.
-In this work we provide an algorithm for calculating polynomial features for a matrix of size matrix of size $NxD$ with $O(d^kD^kN)$ where $k$ is the polynomial order and $d$ is the density of each row.
-The density of a matrix is the percent of its nonzero items, so $0 \le d \le 1$. The standard algorithm is $O(D^kN)$, so the added factor of $d^k$ represents a significant reduction in time.
+While their use is widespread, the authors of this work are unaware of any improvements made to their calculation efficiency.
+In this work we provide an algorithm for calculating polynomial features for a row vector $\vec{x}$ of a $N \times D$ matrix with time and space complexity $O(d^kD^k)$ where $k$ is the polynomial order and $d$ is the density of $\vec{x}$.
+The density of a vector is the percent of elements that are nonzero, so $0 \le d \le 1$.
+The standard algorithm has time and space complexity $O(D^kN)$, so the added factor of $d^k$ represents a significant reduction in time.
 The algorithm does not require the densification of the matrix, e.i. the matrix remains in sparse form, so the space complexity is also $O(d^kD^kN)$ as opposed to $O(D^kN)$.
@@
 \section{Algorithm}
-The standard method of calculating polynomial features for a vector $\vec{x}$ involves augmenting a feature for the product of each combination of features in $\vec{x}$ of orders 2 to $k$.
+The traditional method of calculating polynomial features for a vector $\vec{x}$ involves augmenting a feature for the product of each combination of features in $\vec{x}$ of orders 2 to $k$.
 This method does not exploit the sparsity of a sparse matrix and will yield a product of zero any time one of the features involved in the product is zero.
-In a sparse matrix, such products will be common, and since the default value of a sparse matrix is zero, these products are entirely unnecessary to compute.
+In a sparse matrix, products resulting in zero will be common.
+Since the default value of a sparse matrix is zero, these products are entirely unnecessary to compute.
 
@@
 
-While the idea is straightforward, there is yet one unaddressed caveat; 
-Given a set of nonzero columns whose product was just calculated to generate a polynomial feature, where in the polynomial matrix does the result of the product belong?
+While the idea is straightforward, there is yet one unaddressed caveat.
+Given a set of nonzero columns whose corresponding components were multiplied to produce a polynomial feature, where in the augmented polynomial vector does the result of the product belong?
 To adress this, we give a bijective mapping from the set of possible column index combinations of order 2 to $k$ onto the column index space of the polynomial feature matrix.
-More precisely, this the mapping is of the form
+More precisely, this the mapping is of the following form:
 

### 2016-02-13 paper/workshop/sparse_poly.tex
cleaned up intro

@@
 such that $ 0 \le i_0 \le i_1 \le \dots \le i_{k-1} < D$
-where $(x_{i_0}, x_{i_1}, \dots, x_{i_{k-1}})$ are column indicies of a row vector $\vec{x}$ of an $N \times D$ input matrix and $p_{i_0i_1 \dots i_{k-1}}$ is a column index of the polynomial expansion matrix.
+where $(x_{i_0}, x_{i_1}, \dots, x_{i_{k-1}})$ are column indicies of a row vector $\vec{x}$ of an $N \times D$ input matrix and $p_{i_0i_1 \dots i_{k-1}}$ is a column index into the polynomial expansion vector for $\vec{x}$.
 

### 2016-02-13 paper/workshop/sparse_poly.tex
fixed johns address
fixed abstract

@@
 John Hughes \\
-Department of Computational Neuroscience \\
-University of the Witwatersrand \\
-Joburg, South Africa \\
-\texttt{\{robot,net\}@wits.ac.za} \\
+Department of Computer Science \\
+Brown University \\
+Providence, RI \\
+\texttt{jfh@cs.brown.edu} \\
 }
@@
 We provide an algorithm for polynomial feature expansion that can operate directly on sparse a
-matrix. For a vector $\vec{x}$ of a $N \times D$ matrix, the algorithm has time and space complexity $O(d^kD^k)$ where $k$ is the polynomial order and $d$ is the density of $\vec{X}$.
+matrix. For a vector $\vec{x}$ of a $N \times D$ matrix, the algorithm has time and space complexity $O(d^kD^k)$ where $k$ is the polynomial order and $d$ is the density of $\vec{x}$.
 

### 2016-02-13 paper/workshop/sparse_poly.tex
removed N from big O since its for a vector
swapped a word

@@
 The standard algorithm has time and space complexity $O(D^kN)$, so the added factor of $d^k$ represents a significant reduction in time.
-The algorithm does not require the densification of the matrix, e.i. the matrix remains in sparse form, so the space complexity is also $O(d^kD^kN)$ as opposed to $O(D^kN)$.
+The algorithm does not require the densification of the matrix, e.i. the matrix remains in sparse form, so the space complexity is also $O(d^kD^k)$ as opposed to $O(D^k)$.
 
@@
 While the idea is straightforward, there is yet one unaddressed caveat.
-Given a set of nonzero columns whose corresponding components were multiplied to produce a polynomial feature, where in the augmented polynomial vector does the result of the product belong?
+Given a set of columns whose corresponding nonzero components were multiplied to produce a polynomial feature, where in the augmented polynomial vector does the result of the product belong?
 To adress this, we give a bijective mapping from the set of possible column index combinations of order 2 to $k$ onto the column index space of the polynomial feature matrix.

### 2016-02-14 paper/workshop/sparse_poly.tex
combos with repetition

@@
 \section{Algorithm}
-The traditional method of calculating polynomial features for a vector $\vec{x}$ involves augmenting a feature for the product of each combination of features in $\vec{x}$ of orders 2 to $k$.
+The traditional method of calculating polynomial features for a vector $\vec{x}$ involves augmenting a feature for the product of each combination with repetition (without repetition for interaction features) of features in $\vec{x}$ of orders 2 to $k$.
 This method does not exploit the sparsity of a sparse matrix and will yield a product of zero any time one of the features involved in the product is zero.
@@
 In a compressed sparse row matrix, the columns containing nonzero data are the only columns that are stored.
-We can therefore iterate over products of combinations of orders 2 to $k$ of these columns for each row to calculate $k$-degree polynomial features.
+We can therefore iterate over products of combinations with repetition of order $k$ of these columns for each row to calculate $k$-degree polynomial features.
 
@@
 Given a set of columns whose corresponding nonzero components were multiplied to produce a polynomial feature, where in the augmented polynomial vector does the result of the product belong?
-To adress this, we give a bijective mapping from the set of possible column index combinations of order 2 to $k$ onto the column index space of the polynomial feature matrix.
+To adress this, we give a bijective mapping from the set of possible column index combinations with repetition of order $k$ onto the column index space of the polynomial feature matrix.
 More precisely, this the mapping is of the following form:
@@
 \begin{equation}
-(x_{i_0}, x_{i_1}, \dots, x_{i_{k-1}}) \rightarrowtail \hspace{-1.9ex} \twoheadrightarrow p_{i_0i_1 \dots i_{k-1}} \in \{0,1,\dots, \binom{D}{k}\} 
+(x_{i_0}, x_{i_1}, \dots, x_{i_{k-1}}) \rightarrowtail \hspace{-1.9ex} \twoheadrightarrow p_{i_0i_1 \dots i_{k-1}} \in \{0,1,\dots, \left(\binom{D}{k}\right)\} 
 \forall (i_0, i_1, \dots, i_{k-1})
@@
 such that $ 0 \le i_0 \le i_1 \le \dots \le i_{k-1} < D$
-where $(x_{i_0}, x_{i_1}, \dots, x_{i_{k-1}})$ are column indicies of a row vector $\vec{x}$ of an $N \times D$ input matrix and $p_{i_0i_1 \dots i_{k-1}}$ is a column index into the polynomial expansion vector for $\vec{x}$.
+where $(x_{i_0}, x_{i_1}, \dots, x_{i_{k-1}})$ are column indicies of a row vector $\vec{x}$ of an $N \times D$ input matrix and $p_{i_0i_1 \dots i_{k-1}}$ is a column index into the polynomial expansion vector for $\vec{x}$
+and $\left(\binom{D}{k}\right)$ is the number of combinations with repetition of size $k$ drawn from $D$ objects.
+In general, and is equal to $\left(\binom{n}{k}\right) = \binom{n+k-1}{n-1} = \binom{n+k-1}{k}$, as given by \cite{stanley1986enumerative}.
 
@@
 \subsection{Analytical}
+
+Calculating $k$-degree polynomial features via our method for a vector of dimensionality $D$ and density $d$ requires $dD$ choose $k$ (with repetition) products.
+The big O of the algorithm is therefore given by
+%\begin{align}
+%O\left(\binom{dD+k-1}{k}\right) & = O\left(\binom{dD+k-1}{dD-1}\right) 
+%& = O\left(\right) = O\left(\frac{(dD+k-1)!}{k!(dD-k)!}\right)
+%& = O\left(\frac{(dD+k-1)(dD+k-2) \dots (dD-k+1)}{k!}\right)
+%& = O\left((dD+k-1)(dD+k-2) \dots (dD-k+1)\right) \mbox{for} k << dD
+%& = O\left(\right)
+%
+%dD = 7
+%k = 3
+
+%(dD+k-1)(dD+k-2) \dots (dD-k+1) = (7+3-1)(7+3-2)(7+3-3) (7+3-4)(7+3-5)... (7-3+1=5)
+%\end{align}
+
 \subsection{Empirical}

### 2016-02-14 paper/workshop/sparse_poly.tex
finished analytical analysis
much less than symbol
removed instructions from conference

@@
 The big O of the algorithm is therefore given by
-%\begin{align}
-%O\left(\binom{dD+k-1}{k}\right) & = O\left(\binom{dD+k-1}{dD-1}\right) 
-%& = O\left(\right) = O\left(\frac{(dD+k-1)!}{k!(dD-k)!}\right)
-%& = O\left(\frac{(dD+k-1)(dD+k-2) \dots (dD-k+1)}{k!}\right)
-%& = O\left((dD+k-1)(dD+k-2) \dots (dD-k+1)\right) \mbox{for} k << dD
-%& = O\left(\right)
-%
-%dD = 7
-%k = 3
-
-%(dD+k-1)(dD+k-2) \dots (dD-k+1) = (7+3-1)(7+3-2)(7+3-3) (7+3-4)(7+3-5)... (7-3+1=5)
-%\end{align}
+\begin{align}
+O\left(\binom{dD+k-1}{k}\right) & = O\left(\frac{(dD+k-1)!}{k!(dD-1)!}\right)\\
+& = O\left(\frac{(dD+k-1)(dD+k-2) \dots (dD)}{k!}\right)\\
+& = O\left((dD+k-1)(dD+k-2) \dots (dD)\right) \mbox{ for } k \ll dD\\
+& = O\left(d^kD^k\right)
+\end{align}
 
@@
 \section{Conclusion}
-
-\section{Citations, figures, tables, references}
-\label{others}
-
-These instructions apply to everyone, regardless of the formatter being used.
-
-\subsection{Citations within the text}
-
-Citations within the text should be based on the {\tt natbib} package
-and include the authors' last names and year (with the ``et~al.'' construct
-for more than two authors). When the authors or the publication are
-included in the sentence, the citation should not be in parenthesis (as
-in ``See \citet{Hinton06} for more information.''). Otherwise, the citation
-should be in parenthesis (as in ``Deep learning shows promise to make progress towards AI~\citep{Bengio+chapter2007}.'').
-
-The corresponding references are to be listed in alphabetical order of
-authors, in the \textsc{References} section. As to the format of the
-references themselves, any style is acceptable as long as it is used
-consistently.
-
-\subsection{Footnotes}
-
-Indicate footnotes with a number\footnote{Sample of the first footnote} in the
-text. Place the footnotes at the bottom of the page on which they appear.
-Precede the footnote with a horizontal rule of 2~inches
-(12~picas).\footnote{Sample of the second footnote}
-
-\subsection{Figures}
-
-All artwork must be neat, clean, and legible. Lines should be dark
-enough for purposes of reproduction; art work should not be
-hand-drawn. The figure number and caption always appear after the
-figure. Place one line space before the figure caption, and one line
-space after the figure. The figure caption is lower case (except for
-first word and proper nouns); figures are numbered consecutively.
-
-Make sure the figure caption does not get separated from the figure.
-Leave sufficient space to avoid splitting the figure and figure caption.
-
-You may use color figures.
-However, it is best for the
-figure captions and the paper body to make sense if the paper is printed
-either in black/white or in color.
-\begin{figure}[h]
-\begin{center}
-%\framebox[4.0in]{$\;$}
-\fbox{\rule[-.5cm]{0cm}{4cm} \rule[-.5cm]{4cm}{0cm}}
-\end{center}
-\caption{Sample figure caption.}
-\end{figure}
-
-\subsection{Tables}
-
-All tables must be centered, neat, clean and legible. Do not use hand-drawn
-tables. The table number and title always appear before the table. See
-Table~\ref{sample-table}.
-
-Place one line space before the table title, one line space after the table
-title, and one line space after the table. The table title must be lower case
-(except for first word and proper nouns); tables are numbered consecutively.
-
-\begin{table}[t]
-\caption{Sample table title}
-\label{sample-table}
-\begin{center}
-\begin{tabular}{ll}
-\multicolumn{1}{c}{\bf PART}  &\multicolumn{1}{c}{\bf DESCRIPTION}
-\\ \hline \\
-Dendrite         &Input terminal \\
-Axon             &Output terminal \\
-Soma             &Cell body (contains cell nucleus) \\
-\end{tabular}
-\end{center}
-\end{table}
-
-\section{Final instructions}
-Do not change any aspects of the formatting parameters in the style files.
-In particular, do not modify the width or length of the rectangle the text
-should fit into, and do not change font sizes (except perhaps in the
-\textsc{References} section; see below). Please note that pages should be
-numbered.
-
-\section{Preparing PostScript or PDF files}
-
-Please prepare PostScript or PDF files with paper size ``US Letter'', and
-not, for example, ``A4''. The -t
-letter option on dvips will produce US Letter files.
-
-Consider directly generating PDF files using \verb+pdflatex+
-(especially if you are a MiKTeX user).
-PDF figures must be substituted for EPS figures, however.
-
-Otherwise, please generate your PostScript and PDF files with the following commands:
-\begin{verbatim}
-dvips mypaper.dvi -t letter -Ppdf -G0 -o mypaper.ps
-ps2pdf mypaper.ps mypaper.pdf
-\end{verbatim}
-
-\subsection{Margins in LaTeX}
-
-Most of the margin problems come from figures positioned by hand using
-\verb+\special+ or other commands. We suggest using the command
-\verb+\includegraphics+
-from the graphicx package. Always specify the figure width as a multiple of
-the line width as in the example below using .eps graphics
-\begin{verbatim}
-   \usepackage[dvips]{graphicx} ...
-   \includegraphics[width=0.8\linewidth]{myfile.eps}
-\end{verbatim}
-or % Apr 2009 addition
-\begin{verbatim}
-   \usepackage[pdftex]{graphicx} ...
-   \includegraphics[width=0.8\linewidth]{myfile.pdf}
-\end{verbatim}
-for .pdf graphics.
-See section 4.4 in the graphics bundle documentation (\url{http://www.ctan.org/tex-archive/macros/latex/required/graphics/grfguide.ps})
-
-A number of width problems arise when LaTeX cannot properly hyphenate a
-line. Please give LaTeX hyphenation hints using the \verb+\-+ command.
-
-\subsubsection*{Acknowledgments}
-
-Use unnumbered third level headings for the acknowledgments. All
-acknowledgments, including those to funding agencies, go at the end of the paper.
 

### 2016-02-16 paper/workshop/sparse_poly.tex
trying to make things more clear

@@
 \begin{abstract}
-We provide an algorithm for polynomial feature expansion that can operate directly on sparse a
-matrix. For a vector $\vec{x}$ of a $N \times D$ matrix, the algorithm has time and space complexity $O(d^kD^k)$ where $k$ is the polynomial order and $d$ is the density of $\vec{x}$.
-
+We provide an algorithm for polynomial feature expansion that can operate directly on sparse a matrix.
+For a vector of dimensionality $D$ and density $d$, the algorithm has time and space complexity $O(d^kD^k)$ where $k$ is the polynomial order.
 \end{abstract}
@@
 While their use is widespread, the authors of this work are unaware of any improvements made to their calculation efficiency.
-In this work we provide an algorithm for calculating polynomial features for a row vector $\vec{x}$ of a $N \times D$ matrix with time and space complexity $O(d^kD^k)$ where $k$ is the polynomial order and $d$ is the density of $\vec{x}$.
-The density of a vector is the percent of elements that are nonzero, so $0 \le d \le 1$.
-The standard algorithm has time and space complexity $O(D^kN)$, so the added factor of $d^k$ represents a significant reduction in time.
-The algorithm does not require the densification of the matrix, e.i. the matrix remains in sparse form, so the space complexity is also $O(d^kD^k)$ as opposed to $O(D^k)$.
+In this work we provide an algorithm for calculating polynomial features for a vector of dimensionality $D$ and density $d$ with time and space complexity $O(d^kD^k)$ where $k$ is the polynomial order.
+The density of a vector is the fraction of elements that are nonzero, so $0 \le d \le 1$.
+The standard algorithm has time and space complexity $O(D^k)$, so the added factor of $d^k$ represents a significant reduction in complexity.
+The algorithm does not require the densification of the vector, e.i. the vector remains in compressed sparse row form, so the space complexity is also $O(d^kD^k)$ as opposed to $O(D^k)$.
 
 \section{Algorithm}
-The traditional method of calculating polynomial features for a vector $\vec{x}$ involves augmenting a feature for the product of each combination with repetition (without repetition for interaction features) of features in $\vec{x}$ of orders 2 to $k$.
-This method does not exploit the sparsity of a sparse matrix and will yield a product of zero any time one of the features involved in the product is zero.
+The traditional method of calculating polynomial features for a vector $\vec{x}$ involves augmenting a feature for the product of each combination of size $k$ with repetition (without repetition for interaction features) of features in $\vec{x}$.
+This method does not exploit data sparsity and will yield a product of zero any time one of the features involved in the product is zero.
 In a sparse matrix, products resulting in zero will be common.
-Since the default value of a sparse matrix is zero, these products are entirely unnecessary to compute.
+%If $\vec{x}$ has density $d$, the fraction of nonzero elements of the $k$-order expansion will be $\frac{\left(\binom{dD+k-1}{k}\right)}{\left(\binom{D+k-1}{k}\right)} = \frac{[(dD+k-1)(dD+k-2) \dots (dD+k)](D-1)!}{(dD-1)!} = $
+If the vectors are stored in a sparse matrix format, the products resulting in zero are unnecessary to compute since zero is the default value in a sparse matrix format.
 
-The main idea behind our algorithm is to only compute products that do not involve zeros.
+The main idea behind our algorithm is to leverage sparsity by only computing products that do not involve zeros.
 In a compressed sparse row matrix, the columns containing nonzero data are the only columns that are stored.
@@
 
-While the idea is straightforward, there is yet one unaddressed caveat.
-Given a set of columns whose corresponding nonzero components were multiplied to produce a polynomial feature, where in the augmented polynomial vector does the result of the product belong?
+While the idea is straightforward, there is yet one unaddressed caveat;
+Given a multiset of column indices whose corresponding nonzero components were multiplied to produce a polynomial feature, where in the augmented polynomial vector does the result of the product belong?
 To adress this, we give a bijective mapping from the set of possible column index combinations with repetition of order $k$ onto the column index space of the polynomial feature matrix.
@@
 \begin{equation}
-(x_{i_0}, x_{i_1}, \dots, x_{i_{k-1}}) \rightarrowtail \hspace{-1.9ex} \twoheadrightarrow p_{i_0i_1 \dots i_{k-1}} \in \{0,1,\dots, \left(\binom{D}{k}\right)\} 
+(i_0, i_1, \dots, i_{k-1}) \rightarrowtail \hspace{-1.9ex} \twoheadrightarrow p_{i_0i_1 \dots i_{k-1}} \in \{0,1,\dots, \left(\binom{D}{k}\right)\} 
 \forall (i_0, i_1, \dots, i_{k-1})
@@
 such that $ 0 \le i_0 \le i_1 \le \dots \le i_{k-1} < D$
-where $(x_{i_0}, x_{i_1}, \dots, x_{i_{k-1}})$ are column indicies of a row vector $\vec{x}$ of an $N \times D$ input matrix and $p_{i_0i_1 \dots i_{k-1}}$ is a column index into the polynomial expansion vector for $\vec{x}$
+where $(i_0, i_1, \dots, i_{k-1})$ are column indicies of a row vector $\vec{x}$ of an $N \times D$ input matrix, $p_{i_0i_1 \dots i_{k-1}}$ is a column index into the polynomial expansion vector for $\vec{x}$ where the product of elements corresponding to indices $i_0, i_1, \dots, i_{k-1}$ will be stored,
 and $\left(\binom{D}{k}\right)$ is the number of combinations with repetition of size $k$ drawn from $D$ objects.
-In general, and is equal to $\left(\binom{n}{k}\right) = \binom{n+k-1}{n-1} = \binom{n+k-1}{k}$, as given by \cite{stanley1986enumerative}.
+In general, $\left(\binom{n}{k}\right) = \binom{n+k-1}{n-1} = \binom{n+k-1}{k}$, as given by \cite{stanley1986enumerative}.
 

### 2016-02-18 paper/workshop/sparse_poly.tex
Small test change to see whether Git process works
I'm seeing whether subsequent changes will lead to problems or not.
Shrinking the text, cleaning up a few typos
Mostly changed from passive to active voice. :)

@@
 \begin{abstract}
-We provide an algorithm for polynomial feature expansion that can operate directly on sparse a matrix.
-For a vector of dimensionality $D$ and density $d$, the algorithm has time and space complexity $O(d^kD^k)$ where $k$ is the polynomial order.
+We provide an algorithm for polynomial feature expansion that operates directly on a sparse matrix.
+For a vector of dimension $D$ and density $d$, the algorithm has time and space complexity $O(d^kD^k)$ where $k$ is the polynomial order.
 \end{abstract}
+
+% this is a test-change for Spike to see whether Git seems to work for him. Ignore.
 
@@
 
-Polynomial feature expansion is a tool long used in statistics for approximating nonlinear functions (see \cite{gergonne1974application, smith1918standard}).
-While their use is widespread, the authors of this work are unaware of any improvements made to their calculation efficiency.
-In this work we provide an algorithm for calculating polynomial features for a vector of dimensionality $D$ and density $d$ with time and space complexity $O(d^kD^k)$ where $k$ is the polynomial order.
-The density of a vector is the fraction of elements that are nonzero, so $0 \le d \le 1$.
-The standard algorithm has time and space complexity $O(D^k)$, so the added factor of $d^k$ represents a significant reduction in complexity.
-The algorithm does not require the densification of the vector, e.i. the vector remains in compressed sparse row form, so the space complexity is also $O(d^kD^k)$ as opposed to $O(D^k)$.
+Polynomial feature expansion has long been used in statistics to approximate nonlinear functions~\cite{gergonne1974application, smith1918standard}.
+Despite this, we are unaware of any efforts to optimize calculating them.
+Here we provide an algorithm for calculating polynomial features for a vector of dimension $D$ and density $d$ with time and space complexity $O(d^kD^k)$ where $k$ is the polynomial order, and $0 \le d \le 1$ is the fraction of elements that are nonzero.
+The standard algorithm has time and space complexity $O(D^k)$, so the added factor of $d^k$ represents a significant complexity reduction.
+The algorithm avoids densification of the vector, i.e. the vector remains in compressed sparse row form, so the space complexity is also $O(d^kD^k)$ as opposed to $O(D^k)$.
 
 \section{Algorithm}
-The traditional method of calculating polynomial features for a vector $\vec{x}$ involves augmenting a feature for the product of each combination of size $k$ with repetition (without repetition for interaction features) of features in $\vec{x}$.
-This method does not exploit data sparsity and will yield a product of zero any time one of the features involved in the product is zero.
-In a sparse matrix, products resulting in zero will be common.
+In the naive computation of polynomial features for a vector $\vec{x}$, we create a new feature for each product (with repetition) of $k$ features in $\vec{x}$ (or without repetition, for ``interaction features'').
+This ignores data sparsity and will yield a product of zero any time one of the features involved in the product is zero.
+In a sparse matrix, such zero-products are common.
 %If $\vec{x}$ has density $d$, the fraction of nonzero elements of the $k$-order expansion will be $\frac{\left(\binom{dD+k-1}{k}\right)}{\left(\binom{D+k-1}{k}\right)} = \frac{[(dD+k-1)(dD+k-2) \dots (dD+k)](D-1)!}{(dD-1)!} = $
-If the vectors are stored in a sparse matrix format, the products resulting in zero are unnecessary to compute since zero is the default value in a sparse matrix format.
+If we store vectors in a sparse matrix format, these zero-products need not be computed or stored.
 
@@
 In a compressed sparse row matrix, the columns containing nonzero data are the only columns that are stored.
-We can therefore iterate over products of combinations with repetition of order $k$ of these columns for each row to calculate $k$-degree polynomial features.
+We can therefore iterate over products of combinations with repetition of order $k$ of \emph{only these columns} for each row to calculate $k$-degree polynomial features.
 
-While the idea is straightforward, there is yet one unaddressed caveat;
+While the idea is straightforward, there is yet an unaddressed challenge:
 Given a multiset of column indices whose corresponding nonzero components were multiplied to produce a polynomial feature, where in the augmented polynomial vector does the result of the product belong?
-To adress this, we give a bijective mapping from the set of possible column index combinations with repetition of order $k$ onto the column index space of the polynomial feature matrix.
-More precisely, this the mapping is of the following form:
+To address this, we give a bijective mapping from the set of possible column index combinations-with-repetition of order $k$ onto the column index space of the polynomial feature matrix.
+Thus the map has the form
 
 \begin{equation}
-(i_0, i_1, \dots, i_{k-1}) \rightarrowtail \hspace{-1.9ex} \twoheadrightarrow p_{i_0i_1 \dots i_{k-1}} \in \{0,1,\dots, \left(\binom{D}{k}\right)\} 
-\forall (i_0, i_1, \dots, i_{k-1})
+(i_0, i_1, \dots, i_{k-1}) \rightarrowtail \hspace{-1.9ex} \twoheadrightarrow p_{i_0i_1 \dots i_{k-1}} \in \{0,1,\dots,\binom{D}{k}\} 
 \end{equation}
 such that $ 0 \le i_0 \le i_1 \le \dots \le i_{k-1} < D$
-where $(i_0, i_1, \dots, i_{k-1})$ are column indicies of a row vector $\vec{x}$ of an $N \times D$ input matrix, $p_{i_0i_1 \dots i_{k-1}}$ is a column index into the polynomial expansion vector for $\vec{x}$ where the product of elements corresponding to indices $i_0, i_1, \dots, i_{k-1}$ will be stored,
-and $\left(\binom{D}{k}\right)$ is the number of combinations with repetition of size $k$ drawn from $D$ objects.
-In general, $\left(\binom{n}{k}\right) = \binom{n+k-1}{n-1} = \binom{n+k-1}{k}$, as given by \cite{stanley1986enumerative}.
+where $(i_0, i_1, \dots, i_{k-1})$ are column indicies of a row vector $\vec{x}$ of an $N \times D$ input matrix, and $p_{i_0i_1 \dots i_{k-1}}$ is a column index into the polynomial expansion vector for $\vec{x}$ where the product of elements corresponding to indices $i_0, i_1, \dots, i_{k-1}$ will be stored.
+%and $\left(\binom{D}{k}\right)$ is the number of combinations with repetition of size $k$ drawn from $D$ objects.
+%In general, $\left(\binom{n}{k}\right) = \binom{n+k-1}{n-1} = \binom{n+k-1}{k}$, as given by \cite{stanley1986enumerative}.
 

### 2016-02-18 paper/workshop/sparse_poly.tex
pruned down to 3 pages

@@
 \And
-John Hughes \\
+John F. Hughes \\
 Department of Computer Science \\
@@
 
-We seek a map from matrix indices $(i, j)$ (with $i < j$ and $0 \le i < D$) to numbers $f(i, j)$ with $0 \le f(i, j) < \frac{D(D-1)}{2}$, one that follows the pattern indicated by 
+For the second degree case, we seek a map from matrix indices $(i, j)$ (with $i < j$ and $0 \le i < D$) to numbers $f(i, j)$ with $0 \le f(i, j) < \frac{D(D-1)}{2}$, one that follows the pattern indicated by
 \begin{align}
@@
 
-\section{Analysis}
-\subsection{Analytical}
+\section{Complexity Analysis}
+%\subsection{Analytical}
 
@@
 
-\subsection{Empirical}
-\section{Conclusion}
+%\subsection{Empirical}
+%\section{Conclusion}
 

### 2016-02-18 paper/workshop/sparse_poly.tex
Much reduced section on formula derivations
Removed most algebra and put many displayed equations inline.

@@
 
-For the second degree case, we seek a map from matrix indices $(i, j)$ (with $i < j$ and $0 \le i < D$) to numbers $f(i, j)$ with $0 \le f(i, j) < \frac{D(D-1)}{2}$, one that follows the pattern indicated by
+For the second degree case, we seek a map from matrix indices $(i, j)$ (with $0 \le i < j < D$ ) to numbers $f(i, j)$ with $0 \le f(i, j) < \frac{D(D-1)}{2}$, one that follows the pattern indicated by
 \begin{align}
@@
 \end{align}
-where the entry in row $i$, column $j$, displays the value $f(i, j)$.
-
-To simplify slightly, we introduce a notation for the $n$th triangular number, 
-\begin{align}
-T_2(n) = \frac{n(n+1)}{2}
-\end{align}
-\noindent
-The subscript $2$ indicates that these are triangles in two dimensions; we'll use $T_3(n)$ to indicate the $n$th tetrahedral number, and so on for higher dimensions.
-
-Observe that in Equation~\ref{eq:4x4mat}, each entry in column $j$ (for $j > 0$) lies in the range
-\begin{align}
-T_2(j-1) &\le e < T_2(j).
-\end{align}
-\noindent
-And in fact, the entry in the $i$th row of that column is just $i + T_2(j-1)$. Thus we have
-\begin{align}
+where the entry in row $i$, column $j$, displays the value $f(i, j)$. We let $T_2(n) = \frac{1}{2} n(n+1)$ 
+be the $n$th triangular number; then in Equation~\ref{eq:4x4mat}, column $j$ (for $j > 0$) contains entries with  
+$T_2(j-1) \le e < T_2(j)$; the entry in the $i$th row is just $i + T_2(j-1)$. Thus we have
+$
 f(i, j) 
-&= i + T_2(j-1)\\
-&= i + \frac{(j-1)j}{2}\\
-&=  \frac{2i + j^2-j}{2}.
-\end{align}
-
-For instance, in column $j = 2$ in our example (the \emph{third} column), the entries range from $1$ to $2$, while $T_2(j-1) = T_2(1) = 1$ and $T_2(j) = T_2(2) = 3$, and the entry in column $j = 2$, row $i = 1$ is
+= i + T_2(j-1) =  \frac{1}{2}(2i + j^2-j).$
+For instance, in column $j = 2$ in our example (the \emph{third} column), the entry in row $i = 1$ is 
 $i + T_2(j-1) = 1 + 1 = 2$.
 
-\subsubsection{Other indices}
 With one-based indexing in both the domain and codomain, the formula above becomes
-\begin{align}
-f_1(i, j) &= 1+ f(i-1, j-1) \\
-&= 1+ f(i-1, j-1) \\
-& = \frac{2 + 2(i-1) + (j-1)^2-(j-1)}{2}\\
-& = \frac{2i + j^2 - 3j + 2}{2}
-\end{align}
+$f_1(i, j)  = \frac{1}{2}(2i + j^2 - 3j + 2).$
 
-\subsubsection{Polynomial Features}
-For polynommial features, we seek a map from matrix indices $(i, j)$ (with $i \le j$ and $0 \le i < D$) to numbers $g(i, j)$ with $0 \le f(i, j) < \frac{D(D+1)}{2}$, one that follows the pattern indicated by
-\begin{align}
-\begin{bmatrix}
- 0 & 1 & 3 & 6 \\
- x & 2 & 4 & 7\\
- x & x & 5 & 8 \\
- x & x & x & 9
-\end{bmatrix}
-\label{eq:4x4mat-poly}
-\end{align}
-i.e., essentially the same task as before, except that the diagonal is included.
-One can regard all but the last column of entries in Equation~\ref{eq:4x4mat-poly} as corresponding to the entries in Equation~\ref{eq:4x4mat}, but shifted to the left.
-Thus the formula for $g(i, j)$ is simply the formula for $f$, shifted by 1, i.e.,
-\begin{align}
-g(i, j) &= f(i, j+1)  \\
-&=  \frac{2i + (j+1)^2-(j+1)}{2}\\
-&=  \frac{2i + j^2 + j + 1)}{2}.
-\end{align}
-Alternatively, we can write this as
-\begin{align}
-g(i, j) &= i + T_2(j),
-\end{align}
-\noindent
-and get the same result.
+For \emph{polynomial} features, we seek a similar map $g$, one that also handles the case $i = j$. In this case, a similar analysis yields
+$ g(i, j) = i + T_2(j) = \frac{1}{2} (2i + j^2 + j + 1).$
 
-\subsubsection{Higher dimensions}
-To handle three-way interactions, we need to map triples of indices in a 3-index array to a flat list, and similarly for higher-order interactions.
+To handle \emph{three-way interactions}, we need to map triples of indices in a 3-index array to a flat list, and similarly for higher-order interactions.
+For this, we'll need the tetrahedral numbers $T_3(n) = \sum_{i=1}^n T_{2}(n) =
+\frac{1}{6}(n^3 + 3n^2 + 2n)$.
 
-For three indices, $i,j,k$, with $i < j < k$ and $0 \le i,j,k < D$, we have a similar recurrence. Calling the mapping $h$, we have 
+For three indices, $i,j,k$, with $0 \le i < j < k < D$, we have a similar recurrence. Calling the mapping $h$, we have 
 \begin{align}
@@
 \end{align}
-and from this, the generalization to higher dimensions is straightforward. The formulas for ``higher triangular numbers'', i.e., those defined by
+and from this the generalization to higher dimensions is straightforward. The formulas for ``higher triangular numbers'', i.e., those defined by
 \begin{align}
@@
 \end{align}
-for $k > 1$ can be determined inductively. For $k = 3$, the result is 
-\begin{align}
-T_3(n) &= \sum_{i=1}^n T_{2}(n)\\
-&= \frac{n^3 + 3n^2 + 2n}{6},
-\end{align}
-so that the formula for 3-way interactions, with zero-based indexing, becomes 
+for $k > 1$ can be determined inductively.
+
+The explicit formula for 3-way interactions, with zero-based indexing, is 
 \begin{align}
@@
 \end{align}
-\subsubsection{Higher-dimension polynomial features}
-For the case where we include the diagonal in higher dimensions, we must shift $j$ by $1$, $k$ by $2$, and so on, and the formula becomes
-\begin{align}
-\ell(i,j,k) =  T_1(i) + T_2(j) + T_3(k),
-\end{align}
-with analogous formulas for higher degree polynomial interactions.
 
@@
 
-Calculating $k$-degree polynomial features via our method for a vector of dimensionality $D$ and density $d$ requires $dD$ choose $k$ (with repetition) products.
-The big O of the algorithm is therefore given by
+Calculating $k$-degree polynomial features via our method for a vector of dimensionality $D$ and density $d$ requires $\binom{dD}{k}$ (with repetition) products.
+The complexity of the algorithm, for fixed $k \ll dD$, is therefore
 \begin{align}

### 2016-03-01 directly_on_csr/__init__.py
wow is it fast\!

@@ -1,0 +1,1 @@
+from in_cython import SecDegPolyFeats

### 2016-03-01 directly_on_csr/in_cython.pyx

@@ -34,64 +34,45 @@ class SecDegPolyFeats(BaseEstimator, TransformerMixin):
     cdef np.ndarray[INDEX_T, ndim=1] new_indptr = np.ndarray(shape=indptr.shape[0], dtype=np.int32, order='C')
   
     new_indptr[0] = 0
-    cdef INDEX_T ind = 0, start, stop, num_cols, k1, k2, col1, col2
-    cdef DATA_T data1, poly_data
+    cdef INDEX_T ind = 0, start, stop, num_cols, k1, k2, col2, original_D = X.shape[1]
     i = 0
     
-    """
-    for j in xrange(X.shape[0]):
-      for i in xrange(j+1):
-        X[i, j] = counter
-        counter += 1
-    """
-  
     #Calculate the poly features
     while i < indptr.shape[0]-1:
       start = indptr[i]
-      stop = indptr[i+1]
-    
+      stop = indptr[i+1]    
       num_cols = 0
     
       k1 = start
       while k1 < stop:
         new_data[ind] = data[k1]
         new_indices[ind] = indices[k1]
-      
         ind += 1
         num_cols += 1
         k1 += 1
     
-      k1 = start
-      while k1 < stop:
-        col1 = indices[k1]
-        data1 = data[k1]
+      k2 = start
+      while k2 < stop:
+        col2 = indices[k2]
+        data2 = data[k2]
       
         #Add the poly features
-        k2 = k1
-        while k2 < stop:
-          col2 = indices[k2]
-          poly_col = (2*col1+col2**2+col2+1)/2
-          poly_data = data1*data[k2]
-        
+        k1 = start
+        while k1 < k2+1:
           #Now put everything in its right place
-          new_data[ind] = poly_data
-          new_indices[ind] = poly_col
+          new_data[ind] = data[k1]*data2
+          new_indices[ind] = (2*indices[k1]+col2**2+col2+1)/2 + original_D
           ind += 1
           num_cols += 1
-          k2 += 1
-        k1 += 1
+          k1 += 1
+        k2 += 1
+    
       new_indptr[i+1] = new_indptr[i] + num_cols
       i += 1
-  
-    print 'new_indptr', new_indptr
-    print 'new_indices', new_indices
-    print 'new_data', new_data
-    
[…]

### 2016-03-06 paper/workshop/code/sparse_poly_test.py
modulizing

@@ -1,0 +1,68 @@
+from scipy.sparse import random
+from sklearn.preprocessing import PolynomialFeatures
+from sparse_polynomial_features import SparsePolynomialFeatures
+from time import time
+import numpy as np
+import matplotlib.pyplot as plt
+from scipy.sparse import vstack
+
+"""
+b: blue
+g: green
+r: red
+c: cyan
+m: magenta
+y: yellow
+k: black
+w: white
+"""
+
+colors = ['b', 'r', 'm', 'y', 'm', 'c', 'g', 'k']
+
+def plot_it(degree):
+  Ds = np.arange(1, 6, 1)
+  densities = np.arange(.1, 1.1, .1)#[0.0.1, .2, .3, .4, .5]
+  times_sparse = [[0. for d in Ds] for d in densities]
+  times_dense = [0. for d in Ds]
+  iters = 5
+
+  for iter in xrange(iters):
+    for col, D in enumerate(Ds):
+        for j, density in enumerate(densities):
+          X = random(100000, D, density).tocsr()
+          X_d = X.toarray()
+          
+          for name, l, trans in (('sparse', times_sparse, SparsePolynomialFeatures), ('dense', times_dense, PolynomialFeatures)):
+
+            
+            if name == 'sparse':
+              t = trans(degree)
+              a = time()
+              t.fit_transform(X)
+              b = time()
+              l[j][col] += b-a
+            else:
+              t = trans(degree, include_bias=False)
+              a = time()
+              t.fit_transform(X_d)
+              b = time()
+              l[col] += b-a
+          
+  times_sparse = np.array(times_sparse) / iters
+  times_dense = np.array(times_dense) / iters
+  
+  plt.hold = True
+  #Plot sparse
+  for density, times, c in zip(densities, times_sparse, colors):
+    plt.plot(Ds, times, '%s:' % (c,), label='Sparse d=%s' % density)
+  
+  plt.plot(Ds, times_dense, 'k', label='scikit-learn')
+  plt.xlabel('Dimensionality of Feature Matrix')
+  plt.ylabel('Time to Compute Polynomial Features (seconds)')
+  plt.title('Speed of scikit-learn vs Sparse Method (degree %s)' % (degree,))
+  plt.legend(loc=2)
+  plt.savefig('D_vs_time.pdf')
+  
+
+if __name__ == '__main__':
+  plot_it(3)

### 2016-03-06 setup.py

@@ -3,14 +3,14 @@ from distutils.core import setup
 from os.path import realpath
 
 setup(
-    name="sparse_interaction",
-    version="0.0.0",
-    packages=['sparse_interaction/'],
+    name="sparse_polynomial_features",
+    version="1.0.0",
+    packages=['sparse_polynomial_features/'],
     description="Compute second degree interaction features on a sparse matrix.",
     author="Andrew Nystrom",
     author_email="AWNystrom@gmail.com",
-    url="https://github.com/AWNystrom/SparseInteraction",
-    keywords=["sparse", "feature", "interaction", "combination", "combinations"],
+    url="https://github.com/AWNystrom/SparsePolynomialFeatures",
+    keywords=["sparse", "feature", "features", "polynomial", "interaction", "combination", "combinations"],
     license="Apache 2.0",
     long_description=open(realpath('README.md')).read(),
     classifiers=["Programming Language :: Python",
@@ -21,6 +21,6 @@ setup(
                  "Development Status :: 4 - Beta",
                  "Intended Audience :: Developers"
                  ],
-    install_requires=['numpy', 'sklearn', 'scipy'],
-    ext_modules=cythonize("./sparse_interaction/sparse_interaction.pyx")
+    install_requires=['numpy', 'scipy'],
+    ext_modules=cythonize("./sparse_polynomial_features/sparse_polynomial_features.pyx")
     )

### 2016-03-06 sparse_polynomial_features/__init__.py

@@ -1,0 +1,1 @@
+from sparse_polynomial_features import SparsePolynomialFeatures

### 2016-03-06 sparse_polynomial_features/sparse_polynomial_features.pyx

@@ -1,0 +1,333 @@
+#!python
+#cython: language_level=2, boundscheck=False, cdivision=True, overflowcheck=False, wraparound=False, initializedcheck=False
+
+from __future__ import division
+import numpy as np
+from scipy.sparse import csr_matrix
+from sklearn.base import BaseEstimator, TransformerMixin
+cimport numpy as np
+
+
+ctypedef np.float64_t DATA_T
+ctypedef np.int32_t INDEX_T
+
+__all__ = ['SparsePolynomialFeatures']
+    
+def sec_deg_poly_feats(X):
+    cdef np.ndarray[DATA_T, ndim=1] data = X.data
+    cdef np.ndarray[INDEX_T, ndim=1] indices = X.indices, indptr = X.indptr
+  
+    #Count the number of nonzero items in each row
+    cdef INDEX_T poly_nz_count = 0, i = 0, D
+  
+    while i < indptr.shape[0]-1:
+      D = indptr[i+1] - indptr[i]
+      poly_nz_count += D + <INDEX_T>((D**2+D)/2)
+      i += 1
+  
+    #Make the arrays that will form the new CSR matrix
+    cdef np.ndarray[DATA_T, ndim=1] new_data = np.ndarray(shape=poly_nz_count, dtype=np.float64, order='C')
+    cdef np.ndarray[INDEX_T, ndim=1] new_indices = np.ndarray(shape=poly_nz_count, dtype=np.int32, order='C')
+    cdef np.ndarray[INDEX_T, ndim=1] new_indptr = np.ndarray(shape=indptr.shape[0], dtype=np.int32, order='C')
+  
+    new_indptr[0] = 0
+    cdef INDEX_T ind = 0, start, stop, num_cols, k1, k2, col2, original_D = X.shape[1]
+    cdef DATA_T data2
+    i = 0
+    
+    #Calculate the poly features
+    while i < indptr.shape[0]-1:
+      start = indptr[i]
+      stop = indptr[i+1]    
+      num_cols = 0
+    
+      k1 = start
+      while k1 < stop:
+        new_data[ind] = data[k1]
+        new_indices[ind] = indices[k1]
+        ind += 1
+        num_cols += 1
+        k1 += 1
+    
+      k2 = start
+      while k2 < stop:
+        col2 = indices[k2]
+        data2 = data[k2]
+      
+        #Add the poly features
+        k1 = start
+        while k1 < k2+1:
+          #Now put everything in its right place
[…]

### 2016-03-06 old/c_module/poly.py
cleaning

@@ -1,0 +1,12 @@
+from __future__ import division
+import numpy as np
+cimport numpy as np
+
+ctypedef np.float64_t DTYPE_t
+ctypedef np.int_t INDEX_t
+
+cdef second_deg_poly_float(np.ndarray[INDEX_t, ndim=2] rows, 
+                           np.ndarray[INDEX_t, ndim=2] cols, 
+                           np.ndarray[DTYPE_t, ndim=2] data):
+
+  

### 2016-03-06 old/compare.py

@@ -1,0 +1,26 @@
+import sparse_interaction
+from sklearn.datasets import fetch_20newsgroups_vectorized
+from sklearn.cross_validation import StratifiedKFold
+from sklearn.linear_model import SGDClassifier
+from sklearn.metrics import f1_score
+
+dat = fetch_20newsgroups_vectorized()
+X = dat.data
+Y = dat.target
+cv = StratifiedKFold(Y)
+
+X = X[:, :20000]
+
+si = sparse_interaction.SparseInteractionFeatures()
+X_i = si.transform(X)
+
+scores, scores_i = [], []
+
+clf = SGDClassifier(penalty='l1', n_iter=10)
+
+for train, test in cv:
+    clf.fit(X[train], Y[train])
+    scores.append(f1_score(Y[test], clf.predict(X[test]), average='macro', pos_label=None))
+    clf.fit(X_i[train], Y[train])
+    scores_i.append(f1_score(Y[test], clf.predict(X_i[test]), average='macro', pos_label=None))
+print sum(scores), sum(scores_i)

### 2016-03-06 old/new_sparse_poly.py

@@ -1,0 +1,139 @@
+from sklearn.base import TransformerMixin, BaseEstimator
+import numpy as np
+import scipy as sp
+from scipy.misc import comb
+from scipy.sparse import csr_matrix, coo_matrix
+from code import interact
+
+#TODO add include_bias support
+
+class SparsePolynomialFeatures(TransformerMixin, BaseEstimator):
+    def __init__(self, degree, interaction_only=False, include_bias=False):
+    
+        if degree > 3:
+            raise Exception("SparsePolynomialFeatures currently supports degrees of 2 or 3.")
+            
+        self.degree = degree
+        self.interaction_only = interaction_only
+        self.include_bias = include_bias
+        self.__poly_functs = [self.__second_deg_poly, self.__third_deg_poly]
+        self.__inter_functs = [self.__second_deg_inter, self.__third_deg_inter]
+    
+    def fit(self, X, **fit_params):
+        return self
+        
+    def transform(self, X, Y=None, **transform_params):
+    
+        X = X.tocoo() #Put back into original format. Also put the new matrix into the same format.
+    
+        #How big will the output matrix be?
+        N, D = X.shape
+    
+        #start degree at 2 since we're going to just copy the first degree features
+        sizes = [int(comb(D, k)) for k in xrange(2, self.degree+1)] if self.interaction_only else [int(comb(D+k-1, D-1)) for k in xrange(2, self.degree+1)]
+        print 'the sizes:', sizes
+        stop_inds = np.cumsum(sizes)+D #non-inclusive
+        offset = D
+        
+        rows, cols, data = list(X.row), list(X.col), list(X.data) #use linked lists? Contiguous memory might be faster.
+        cur_ind = 0
+        
+        X = X.tocsr()
+        
+        funct_list = self.__inter_functs if self.interaction_only else self.__poly_functs
+        
+        for size, stop_ind, degree, funct in zip(sizes, stop_inds, xrange(2, self.degree+1), funct_list):
+            funct(X, D, offset, rows, cols, data)
+            offset += size
+        
+        return coo_matrix((data, (rows, cols)), shape=(N, D+sum(sizes))).tocsr()
+        
+    """
+    Each polynomial has its own method. I could have put it all into one, but
+    this would cause lots of function calls, which would add some time overhead.
+    """
+    
+    def __second_deg_inter(self, X, D, offset, rows, cols, data):
[…]

### 2016-03-06 old/old_setup.py

@@ -1,0 +1,26 @@
+from Cython.Build import cythonize
+from distutils.core import setup
+from os.path import realpath
+
+setup(
+    name="sparse_interaction",
+    version="0.0.0",
+    packages=['sparse_interaction/'],
+    description="Compute second degree interaction features on a sparse matrix.",
+    author="Andrew Nystrom",
+    author_email="AWNystrom@gmail.com",
+    url="https://github.com/AWNystrom/SparseInteraction",
+    keywords=["sparse", "feature", "interaction", "combination", "combinations"],
+    license="Apache 2.0",
+    long_description=open(realpath('README.md')).read(),
+    classifiers=["Programming Language :: Python",
+                 "Programming Language :: Python :: 2.7",
+                 "Programming Language :: Python :: 2",
+                 "License :: OSI Approved :: Apache Software License",
+                 "Operating System :: OS Independent",
+                 "Development Status :: 4 - Beta",
+                 "Intended Audience :: Developers"
+                 ],
+    install_requires=['numpy', 'sklearn', 'scipy'],
+    ext_modules=cythonize("./sparse_interaction/sparse_interaction.pyx")
+    )

### 2016-03-06 old/outer_product_baseline.py

@@ -1,0 +1,15 @@
+from numpy import outer, triu_indices
+from scipy.sparse import csr_matrix, coo_matrix, triu
+from code import interact
+
+def second_deg_poly_features(X):
+  D = X.shape[1]
+  D2 = (D**2+D)/2+D
+  print D2
+  X2 = coo_matrix(X.tocsr(), shape=(X.shape[0], D2)).tocsr()
+  for i, row in enumerate(X):
+    interact(local=locals())
+    nzrows, nzcols = triu(outer(row.T,row)[0][0]).flatten()
+    print r.shape
+    X2[i, D:] = r
+  return X2

### 2016-03-06 old/regression.py

@@ -1,0 +1,37 @@
+from sklearn.linear_model import LinearRegression
+from sklearn.preprocessing import PolynomialFeatures
+import directly_on_csr
+from scipy.sparse import csr_matrix
+import matplotlib.pyplot as plt
+import numpy as np
+
+X = np.arange(0, 1*np.pi, 0.01)
+X = X.reshape(X.shape[0], 1)
+#X_poly = PolynomialFeatures(degree=2, include_bias=False).fit_transform(X)
+X_poly = directly_on_csr.SecDegPolyFeats().fit_transform(csr_matrix(X)).toarray()
+Y = np.sin(X).flatten()
+noise = np.random.normal(0, 0.1, len(Y))
+print Y.shape, noise.shape
+Y = Y+noise
+
+clf1 = LinearRegression()
+clf1.fit(X, Y)
+pred1 = clf1.predict(X)
+
+clf2 = LinearRegression()
+clf2.fit(X_poly, Y)
+pred2 = clf2.predict(X_poly)
+
+fig, ax = plt.subplots(1)
+ax.scatter(X, Y, alpha=0.3)
+ax.plot(X, pred1, label="Linear Prediction")
+ax.plot(X, pred2, label="Polynomial Prediction")
+ax.legend()
+ax.set_xlabel(r'$x$')
+ax.set_ylabel(r'$y$')
+textstr = r'$y = sin(x) + \mathit{N}(0, 0.1)$'
+props = dict(boxstyle='round', facecolor='wheat', alpha=0.5)
+ax.text(0.05, 0.95, textstr, transform=ax.transAxes, fontsize=14,
+        verticalalignment='top', bbox=props)
+ax.set_title('Goodness of fit of polynomial features to a trigonomic function')
+plt.show()

### 2016-03-06 old/sparse_interaction/__init__.py

@@ -1,0 +1,1 @@
+from sparse_interaction import SparsePolynomialFeatures

### 2016-03-06 old/sparse_interaction/sparse_interaction.pyx

@@ -1,0 +1,107 @@
+#!python
+#cython: language_level=2, boundscheck=False, cdivision=True, overflowcheck=False, wraparound=False, initializedcheck=False
+
+from __future__ import division
+import numpy as np
+cimport numpy as np
+import cython
+from sklearn.base import TransformerMixin, BaseEstimator
+from scipy.sparse import coo_matrix
+
+from libc.math cimport pow
+
+ctypedef np.uint64_t INDEXDTYPE
+
+class SparsePolynomialFeatures(TransformerMixin, BaseEstimator):
+    """
+    Compute interaction features on a sparse (CSR) matrix while keeping its format sparse.
+    """
+    
+    def __init__(self, interaction_only=False):
+        self.interaction_only = interaction_only
+    
+    def fit(self, X, Y=None, **fit_params):
+        return self
+    
+    @cython.cdivision(True)
+    def transform(self, X, int interaction_only=False, int combine=True, **transform_params):
+                                                        
+        cdef INDEXDTYPE a, b, x, y, row_num, interaction_col, n_cols_X, output_dim, nnz, num_interactions = 0, min_col_count
+        cdef int combo_j_offset, combo_i_offset
+        
+        n_cols_X = X.shape[1]
+        if interaction_only == 1:
+            combo_j_offset = 1
+            combo_i_offset = -1
+            output_dim = <INDEXDTYPE>((n_cols_X**2-n_cols_X) / 2) #upper triangular
+            min_col_count = 2
+            for nnz in np.bincount(X.nonzero()[0]):
+                num_interactions += <INDEXDTYPE>((pow(nnz,2)-nnz)/2)
+        else:
+            combo_j_offset = 0
+            combo_i_offset = 0
+            min_col_count = 1
+            output_dim = <INDEXDTYPE>((n_cols_X**2+n_cols_X) / 2) #diagonal counted
+            for nnz in np.bincount(X.nonzero()[0]):
+                num_interactions += <INDEXDTYPE>((pow(nnz,2)+nnz)/2)
+        
+        if combine:
+            num_interactions += X.nnz
+            output_dim += X.shape[1]
+            
+        if num_interactions == 0:
+            return X
+        
+        cdef np.ndarray[INDEXDTYPE, ndim=1, mode='c'] rows = np.ndarray(shape=num_interactions, dtype=np.uint64, order='C')
+        cdef np.ndarray[INDEXDTYPE, ndim=1, mode='c'] cols = np.ndarray(shape=num_interactions, dtype=np.uint64, order='C')
[…]

### 2016-03-06 old/test.py

@@ -1,0 +1,33 @@
+import pyximport; pyximport.install()
+import sparse_interaction
+from scipy.sparse import csr_matrix, rand
+from time import time
+from sklearn.preprocessing import PolynomialFeatures
+from code import interact
+from numpy import  array_equal
+interaction = sparse_interaction.SparseInteractionFeatures()
+
+X = rand(10, 10000, 0.2).tocsr()#csr_matrix([[1,1,0,0],[0,0,1,1],[1,0,1,0],[0,1,0,1], [1,1,1,0], [0,1,1,0]])#
+raw_input('made matrix')
+#X = csr_matrix([[1,1,0,0],[0,0,1,1],[1,0,1,0],[0,1,0,1], [1,1,1,0], [0,1,1,0]])##rand(1000, 4, 0.5).tocsr()#
+
+theirs = PolynomialFeatures(interaction_only=True)
+t0 = time()
+B = interaction.fit_transform(X)
+
+t1 = time()
+raw_input('transformed matrix')
+print 'Mine:', t1-t0, B.shape
+
+t0 = time()
+C = theirs.fit_transform(X.toarray())
+t1 = time()
+print 'Theirs:', t1-t0, C.shape
+
+#Get rid of the dummy feature sklearn adds
+C = C[:,1:]
+
+print B.toarray()
+print C
+
+print (B==C).sum() == B.shape[0]*B.shape[1] and B.shape==C.shape

### 2016-03-06 sparse_polynomial_features/pure_python.py

@@ -1,0 +1,124 @@
+from scipy.sparse import csr_matrix
+from numpy import array, empty, int32, zeros
+
+def second_deg(X):
+  data, indices, indptr = X.data, X.indices, X.indptr
+  
+  #Count the number of nonzero items in each row
+  poly_nz_count = 0
+  for i in xrange(indptr.shape[0]-1):
+    D = indptr[i+1] - indptr[i]
+    poly_nz_count += D + (D**2+D)/2
+  
+  #Make the arrays that will form the new CSR matrix
+  new_data = zeros(shape=(poly_nz_count,), dtype=X.dtype, order='C')
+  new_indices = zeros(shape=(poly_nz_count,), dtype='int32', order='C')
+  new_indptr = zeros(shape=(X.indptr.shape[0],), dtype='int32', order='C')
+  
+  print 'poly_nz_count', poly_nz_count
+  
+  new_indptr[0] = 0
+  
+  ind = 0
+  
+  original_D = X.shape[1]
+  
+  #Calculate the poly features
+  for i in xrange(indptr.shape[0]-1):
+    start = indptr[i]
+    stop = indptr[i+1]
+    
+    cols = indices[start:stop]
+    
+    num_cols = 0
+    
+    for k1 in xrange(start, stop):
+      new_data[ind] = data[k1]
+      new_indices[ind] = indices[k1]
+      print ind
+      ind += 1
+      num_cols += 1
+    
+    for k2 in xrange(start, stop):
+      col2 = indices[k2]
+      data2 = data[k2]
+      
+      #Add the poly features
+      for k1 in xrange(start, k2+1):
+        col1 = indices[k1]
+        data1 = data[k1]
+        poly_col = (2*col1+col2**2+col2+1)/2 + original_D
+        poly_data = data1*data2
+        
+        #Now put everything in its right place
+        print ind
+        new_data[ind] = poly_data
+        new_indices[ind] = poly_col
+        ind += 1
+        num_cols += 1
+    
+    new_indptr[i+1] = new_indptr[i] + num_cols
+  
+  print
+  print 'data', new_data
+  print
+  print 'indices', new_indices
+  print
+  print 'indptr', new_indptr
+  print
+  
+  A = csr_matrix([])
+  A.data = new_data
+  A.indices = new_indices
+  A.indptr = new_indptr
[…]

### 2016-03-06 test.py
Delete test.py

@@ -1,33 +1,0 @@
-import pyximport; pyximport.install()
-import sparse_interaction
-from scipy.sparse import csr_matrix, rand
-from time import time
-from sklearn.preprocessing import PolynomialFeatures
-from code import interact
-from numpy import  array_equal
-interaction = sparse_interaction.SparseInteractionFeatures()
-
-X = rand(10, 10000, 0.2).tocsr()#csr_matrix([[1,1,0,0],[0,0,1,1],[1,0,1,0],[0,1,0,1], [1,1,1,0], [0,1,1,0]])#
-raw_input('made matrix')
-#X = csr_matrix([[1,1,0,0],[0,0,1,1],[1,0,1,0],[0,1,0,1], [1,1,1,0], [0,1,1,0]])##rand(1000, 4, 0.5).tocsr()#
-
-theirs = PolynomialFeatures(interaction_only=True)
-t0 = time()
-B = interaction.fit_transform(X)
-
-t1 = time()
-raw_input('transformed matrix')
-print 'Mine:', t1-t0, B.shape
-
-t0 = time()
-C = theirs.fit_transform(X.toarray())
-t1 = time()
-print 'Theirs:', t1-t0, C.shape
-
-#Get rid of the dummy feature sklearn adds
-C = C[:,1:]
-
-print B.toarray()
-print C
-
-print (B==C).sum() == B.shape[0]*B.shape[1] and B.shape==C.shape

### 2016-03-06 sparse_interaction/sparse_interaction.pyx
Delete sparse_interaction.pyx

@@ -1,107 +1,0 @@
-#!python
-#cython: language_level=2, boundscheck=False, cdivision=True, overflowcheck=False, wraparound=False, initializedcheck=False
-
-from __future__ import division
-import numpy as np
-cimport numpy as np
-import cython
-from sklearn.base import TransformerMixin, BaseEstimator
-from scipy.sparse import coo_matrix
-
-from libc.math cimport pow
-
-ctypedef np.uint64_t INDEXDTYPE
-
-class SparsePolynomialFeatures(TransformerMixin, BaseEstimator):
-    """
-    Compute interaction features on a sparse (CSR) matrix while keeping its format sparse.
-    """
-    
-    def __init__(self, interaction_only=False):
-        self.interaction_only = interaction_only
-    
-    def fit(self, X, Y=None, **fit_params):
-        return self
-    
-    @cython.cdivision(True)
-    def transform(self, X, int interaction_only=False, int combine=True, **transform_params):
-                                                        
-        cdef INDEXDTYPE a, b, x, y, row_num, interaction_col, n_cols_X, output_dim, nnz, num_interactions = 0, min_col_count
-        cdef int combo_j_offset, combo_i_offset
-        
-        n_cols_X = X.shape[1]
-        if interaction_only == 1:
-            combo_j_offset = 1
-            combo_i_offset = -1
-            output_dim = <INDEXDTYPE>((n_cols_X**2-n_cols_X) / 2) #upper triangular
-            min_col_count = 2
-            for nnz in np.bincount(X.nonzero()[0]):
-                num_interactions += <INDEXDTYPE>((pow(nnz,2)-nnz)/2)
-        else:
-            combo_j_offset = 0
-            combo_i_offset = 0
-            min_col_count = 1
-            output_dim = <INDEXDTYPE>((n_cols_X**2+n_cols_X) / 2) #diagonal counted
-            for nnz in np.bincount(X.nonzero()[0]):
-                num_interactions += <INDEXDTYPE>((pow(nnz,2)+nnz)/2)
-        
-        if combine:
-            num_interactions += X.nnz
-            output_dim += X.shape[1]
-            
-        if num_interactions == 0:
-            return X
-        
-        cdef np.ndarray[INDEXDTYPE, ndim=1, mode='c'] rows = np.ndarray(shape=num_interactions, dtype=np.uint64, order='C')
-        cdef np.ndarray[INDEXDTYPE, ndim=1, mode='c'] cols = np.ndarray(shape=num_interactions, dtype=np.uint64, order='C')
[…]

### 2016-03-06 directly_on_csr/__init__.py
Delete __init__.py

@@ -1,1 +1,0 @@
-from in_cython import SecDegPolyFeats

### 2016-03-06 directly_on_csr/in_cython.pyx
Delete in_cython.pyx

@@ -1,78 +1,0 @@
-#!python
-#cython: language_level=2, boundscheck=False, cdivision=True, overflowcheck=False, wraparound=False, initializedcheck=False
-
-from __future__ import division
-import numpy as np
-from scipy.sparse import csr_matrix
-from sklearn.base import BaseEstimator, TransformerMixin
-cimport numpy as np
-
-
-ctypedef np.float64_t DATA_T
-ctypedef np.int32_t INDEX_T
-
-class SecDegPolyFeats(BaseEstimator, TransformerMixin):
-
-  def fit(self, X, Y=None, **fit_params):
-    return self
-    
-  def transform(self, X, **transform_params):
-    cdef np.ndarray[DATA_T, ndim=1] data = X.data
-    cdef np.ndarray[INDEX_T, ndim=1] indices = X.indices, indptr = X.indptr
-  
-    #Count the number of nonzero items in each row
-    cdef INDEX_T poly_nz_count = 0, i = 0, D
-  
-    while i < indptr.shape[0]-1:
-      D = indptr[i+1] - indptr[i]
-      poly_nz_count += D + <INDEX_T>((D**2+D)/2)
-      i += 1
-  
-    #Make the arrays that will form the new CSR matrix
-    cdef np.ndarray[DATA_T, ndim=1] new_data = np.ndarray(shape=poly_nz_count, dtype=np.float64, order='C')
-    cdef np.ndarray[INDEX_T, ndim=1] new_indices = np.ndarray(shape=poly_nz_count, dtype=np.int32, order='C')
-    cdef np.ndarray[INDEX_T, ndim=1] new_indptr = np.ndarray(shape=indptr.shape[0], dtype=np.int32, order='C')
-  
-    new_indptr[0] = 0
-    cdef INDEX_T ind = 0, start, stop, num_cols, k1, k2, col2, original_D = X.shape[1]
-    i = 0
-    
-    #Calculate the poly features
-    while i < indptr.shape[0]-1:
-      start = indptr[i]
-      stop = indptr[i+1]    
-      num_cols = 0
-    
-      k1 = start
-      while k1 < stop:
-        new_data[ind] = data[k1]
-        new_indices[ind] = indices[k1]
-        ind += 1
-        num_cols += 1
-        k1 += 1
-    
-      k2 = start
-      while k2 < stop:
-        col2 = indices[k2]
-        data2 = data[k2]
-      
-        #Add the poly features
-        k1 = start
[…]

### 2016-03-06 compare.py
Delete compare.py

@@ -1,26 +1,0 @@
-import sparse_interaction
-from sklearn.datasets import fetch_20newsgroups_vectorized
-from sklearn.cross_validation import StratifiedKFold
-from sklearn.linear_model import SGDClassifier
-from sklearn.metrics import f1_score
-
-dat = fetch_20newsgroups_vectorized()
-X = dat.data
-Y = dat.target
-cv = StratifiedKFold(Y)
-
-X = X[:, :20000]
-
-si = sparse_interaction.SparseInteractionFeatures()
-X_i = si.transform(X)
-
-scores, scores_i = [], []
-
-clf = SGDClassifier(penalty='l1', n_iter=10)
-
-for train, test in cv:
-    clf.fit(X[train], Y[train])
-    scores.append(f1_score(Y[test], clf.predict(X[test]), average='macro', pos_label=None))
-    clf.fit(X_i[train], Y[train])
-    scores_i.append(f1_score(Y[test], clf.predict(X_i[test]), average='macro', pos_label=None))
-print sum(scores), sum(scores_i)

### 2016-03-06 regression.py
Delete regression.py

@@ -1,34 +1,0 @@
-from sklearn.linear_model import LinearRegression
-from sklearn.preprocessing import PolynomialFeatures
-import matplotlib.pyplot as plt
-import numpy as np
-
-X = np.arange(0, 1*np.pi, 0.01)
-X = X.reshape(X.shape[0], 1)
-X_poly = PolynomialFeatures(degree=2, include_bias=False).fit_transform(X)
-Y = np.sin(X).flatten()
-noise = np.random.normal(0, 0.1, len(Y))
-print Y.shape, noise.shape
-Y = Y+noise
-
-clf1 = LinearRegression()
-clf1.fit(X, Y)
-pred1 = clf1.predict(X)
-
-clf2 = LinearRegression()
-clf2.fit(X_poly, Y)
-pred2 = clf2.predict(X_poly)
-
-fig, ax = plt.subplots(1)
-ax.scatter(X, Y, alpha=0.3)
-ax.plot(X, pred1, label="Linear Prediction")
-ax.plot(X, pred2, label="Polynomial Prediction")
-ax.legend()
-ax.set_xlabel(r'$x$')
-ax.set_ylabel(r'$y$')
-textstr = r'$y = sin(x) + \mathit{N}(0, 0.1)$'
-props = dict(boxstyle='round', facecolor='wheat', alpha=0.5)
-ax.text(0.05, 0.95, textstr, transform=ax.transAxes, fontsize=14,
-        verticalalignment='top', bbox=props)
-ax.set_title('Goodness of fit of polynomial features to a trigonomic function')
-plt.show()

### 2016-03-06 time_space_compare.py
Delete time_space_compare.py

@@ -1,27 +1,0 @@
-from memory_profiler import profile
-from sparse_interaction import SparsePolynomialFeatures
-from sklearn.preprocessing import PolynomialFeatures
-from cPickle import load
-from scipy.sparse import csr_matrix
-from sklearn.datasets import fetch_20newsgroups_vectorized
-from time import time
-
-#@profile
-def sparse(X):
-    si = SparsePolynomialFeatures()
-    X_i = si.transform(X)
-    
-#@profile
-def dense(X):
-    poly = PolynomialFeatures(include_bias=False)
-    X_i = poly.fit_transform(X)
-
-if __name__ == '__main__':
-    #X_filename = 'connect_4/X'
-    #X = load(open(X_filename))
-    #X_sparse = csr_matrix(X)
-    X_sparse = fetch_20newsgroups_vectorized().data
-    X = X_sparse.toarray()
-
-    a=time(); dense(X); print time()-a
-    #a=time(); sparse(X_sparse); print time()-a

### 2016-03-06 random_matrix_compare.py
Delete random_matrix_compare.py

@@ -1,158 +1,0 @@
-from time import time
-from scipy.sparse import rand
-from sparse_interaction import SparsePolynomialFeatures
-import matplotlib.pyplot as plt
-from sklearn.preprocessing import PolynomialFeatures
-from code import interact
-from numpy import arange, median
-from memory_profiler import profile
-import StringIO
-
-N = 1000
-iters=1
-
-sp = SparsePolynomialFeatures()
-poly = PolynomialFeatures(include_bias=False)
-
-dD_to_times = {}
-other_d_to_times={}
-
-min_D = 1
-D_step = 100
-max_D = 500+D_step
-
-
-min_d = 0.2
-d_step = .2
-max_d = 1.
-
-def speed_test():
-    for i in xrange(iters):
-        for d in arange(min_d, max_d, d_step):
-            for D in arange(min_D, max_D, D_step):
-                X = rand(N, D, d).tocsr()
-                X_d = X.toarray()
-        
-                a = time()
-                A = sp.transform(X)
-                t = time()-a
-                if d not in dD_to_times:
-                    dD_to_times[d] = {}
-            
-                dD_to_times[d][D] = dD_to_times[d].get(D, 0) + t
-                print i, d, D, t
-            
-                if d == max_d-d_step:
-                    a = time()
-                    B = poly.fit_transform(X_d)
-                    t = time()-a
-                    other_d_to_times[D] = other_d_to_times.get(D, 0) + t
-
-
-    for d in sorted(dD_to_times.keys()):
-        for D in dD_to_times[d].keys():
-            dD_to_times[d][D] /= iters
-        pairs = dD_to_times[d].items()
-        pairs.sort()
-        x, y = zip(*pairs)
-        plt.plot(x, y, label='d=%s'%(d,))
-    
-    for k, val in other_d_to_times.iteritems():
-        other_d_to_times[k] /= iters
-
-    pairs = other_d_to_times.items()
-    pairs.sort()
-    x, y = zip(*pairs)
-    plt.plot(x, y, label='scikit-learn', color='red', marker='*')
-
-    plt.title('Times of Polynomial Feature Generation on Random Matrices With %s Rows' % (N,))
-    plt.xlabel('Dimensionality')
-    plt.ylabel('Time to Calculate (seconds)')
-    plt.legend(loc='best')
-    plt.xlim([0, max_D-D_step])
-    plt.grid(True)
[…]

### 2016-03-06 new_sparse_poly.py
Delete new_sparse_poly.py

@@ -1,139 +1,0 @@
-from sklearn.base import TransformerMixin, BaseEstimator
-import numpy as np
-import scipy as sp
-from scipy.misc import comb
-from scipy.sparse import csr_matrix, coo_matrix
-from code import interact
-
-#TODO add include_bias support
-
-class SparsePolynomialFeatures(TransformerMixin, BaseEstimator):
-    def __init__(self, degree, interaction_only=False, include_bias=False):
-    
-        if degree > 3:
-            raise Exception("SparsePolynomialFeatures currently supports degrees of 2 or 3.")
-            
-        self.degree = degree
-        self.interaction_only = interaction_only
-        self.include_bias = include_bias
-        self.__poly_functs = [self.__second_deg_poly, self.__third_deg_poly]
-        self.__inter_functs = [self.__second_deg_inter, self.__third_deg_inter]
-    
-    def fit(self, X, **fit_params):
-        return self
-        
-    def transform(self, X, Y=None, **transform_params):
-    
-        X = X.tocoo() #Put back into original format. Also put the new matrix into the same format.
-    
-        #How big will the output matrix be?
-        N, D = X.shape
-    
-        #start degree at 2 since we're going to just copy the first degree features
-        sizes = [int(comb(D, k)) for k in xrange(2, self.degree+1)] if self.interaction_only else [int(comb(D+k-1, D-1)) for k in xrange(2, self.degree+1)]
-        print 'the sizes:', sizes
-        stop_inds = np.cumsum(sizes)+D #non-inclusive
-        offset = D
-        
-        rows, cols, data = list(X.row), list(X.col), list(X.data) #use linked lists? Contiguous memory might be faster.
-        cur_ind = 0
-        
-        X = X.tocsr()
-        
-        funct_list = self.__inter_functs if self.interaction_only else self.__poly_functs
-        
-        for size, stop_ind, degree, funct in zip(sizes, stop_inds, xrange(2, self.degree+1), funct_list):
-            funct(X, D, offset, rows, cols, data)
-            offset += size
-        
-        return coo_matrix((data, (rows, cols)), shape=(N, D+sum(sizes))).tocsr()
-        
-    """
-    Each polynomial has its own method. I could have put it all into one, but
-    this would cause lots of function calls, which would add some time overhead.
-    """
-    
-    def __second_deg_inter(self, X, D, offset, rows, cols, data):
[…]

### 2016-10-30 paper/Final/sparse_poly.tex
final version of paper inprogress

\documentclass{article} % For LaTeX2e

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{clrscode3e}
\usepackage{esvect}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{mathtools}
\usepackage{times}
\usepackage{url}

\title{Efficient Calculation of Polynomial Features on Sparse Matrices}

%\author{Andrew Nystrom \\
%Savvysherpa Inc.\\
%6200 Shingle Creek Pkwy \\
%Suite 400 \\
%Minneapolis, MN 55430, USA \\
%\texttt{awnystrom@gmail.com} \\
%\And
%John F. Hughes \\
%Department of Computer Science \\
%Brown University \\
%Providence, RI \\
%\texttt{jfh@cs.brown.edu} \\
%}

\author{
  Nystrom, Andrew\\
  \texttt{awnystrom@gmail.com}
  \and
  Hughes, John\\
  \texttt{jfh@cs.brown.edu}
}

% The \author macro works with any number of authors. There are two commands
% used to separate the names and addresses of multiple authors: \And and \AND.
%
% Using \And between authors leaves it to \LaTeX{} to determine where to break
% the lines. Using \AND forces a linebreak at that point. So, if \LaTeX{}
% puts 3 of 4 authors names on the first line, and the last on the second
% line, try using \AND instead of \And before the third author name.

%\newcommand{\fix}{\marginpar{FIX}}
%\newcommand{\new}{\marginpar{NEW}}
\newcommand{\matr}[1]{\bm{#1}}

\begin{document}

\maketitle

\begin{abstract}
We provide an algorithm for polynomial feature expansion that operates directly on and produces a compressed sparse row matrix without any densification.
For a vector of dimension $D$ and density $d$, the algorithm has time and space complexity $O(d^kD^k)$ where $k$ is the polynomial order, which is an improvement by a factor $d^k$ over the standard method.
\end{abstract}

% this is a test-change for Spike to see whether Git seems to work for him. Ignore.

\section{Introduction}

Polynomial feature expansion has long been used in statistics to approximate nonlinear functions~\cite{gergonne1974application, smith1918standard}.
The compressed sparse row (CSR) matrix format is a widely-used data structure to hold design matrices for statistics and machine learning applications.
However, polynomial expansions cannot be performed directly on sparse CSR matrices, or any sparse matrix format for that matter, without intermediate densification steps.
This densification not only adds extra overhead, but causes combinations of features that have a product of zero to be computed, then put back into a sparse format, which is futile.

We provide an algorithm that allows CSR matrices to be the input of a polynomial feature expansion without any densification.
The algorithm leverages the CSR format to only compute products of features that result in a nonzero value.
This exploits the sparsity of the data to achieve an improved time complexity of $O(d^kD^k)$ on each vector of the matrix where $k$ is the degree of the expansion, $D$ is the dimensionality, and $d$ is the density.
The standard algorithm has time complexity $O(D^k)$.
Since $0 \le d \le 1$, our algorithm is a significant improvement.
While the algorithm we lay out is uses CSR matrices, it could be slightly modified to operate on other sparse formats.

\section{Preliminaries}
A matrix is uppercase and bold, like this: $\bm{A}$.
In this the context of this paper, row vectors are discussed, but not collum vectors.
Therefore, the distinction between the two is unimportant.
A row vector $\bm{a}_i$ is the $i^{th}$ row of $\bm{A}$ and $\bm{a}$, without a subscript, is a vector not necessarily related to $\bm{A}$.

A compressed sparse row (CSR) matrix representation of $\bm{A}$ consists of three vectors: $\bm{c}$, $\bm{d}$, and $\bm{p}$.
$\bm{c}$ and $\bm{d}$ hold the column indices and data values, respectively, of all nonzero elements of $\bm{A}$.
The values of $\bm{p}$ index $\bm{c}$ and $\bm{d}$ and is itself indexed by rows of $\bm{A}$ such that the nonzero columns of $\bm{a}_i$ are $\bm{c}_{\bm{p}_i:\bm{p}_{i+1}}$ and the nonzero data elements are $\bm{d}_{\bm{p}_i:\bm{p}_{i+1}}$.

\section{Motivation}
In this section, an algorithm for computing polynomial feature expansions on dense matrices will be given.
Next, the algorithm will be modified slightly to operate on a CSR matrix in order to expose its infeasibility.
We will show how the algorithm would be feasible with an added component and derive that component in the following section.

\subsection{Dense Expansion Algorithm}
A natural way to calculate polynomial features for a matrix $\bm{A}$ is to walk down its rows and, for each row, take products of all $k$ combinations of elements.
In order to determine in which column $\bm{A}^k_i$ products of elements in $\bm{A}_i$ belong, a simple counter can be set to zero for each row of $bm{A}$ and incremented efter each polynomial feature is generated.
This counter gives the column of $bm{A}^k_i$ into which each expansion feature belongs.

\begin{codebox}
\footnotesize
\Procname{$\proc{Second Order Dense Polynomial Expansion Algorithm}(\bm{A})$}
    \li $N \gets$ row count of $\bm{A}$
    \li $D \gets$ column count of $\bm{A}$
    \li $\bm{A}^k$ $\gets$ empty $N \times \binom{D}{2}$ matrix
    \li \For $i \gets 0 \To N-1$ \Do
    \li     $c_p \gets 0$
    \li     \For $j_1 \gets 0 \To D-1$ \Do
    \li         \For $j_2 \gets j_1 \To D-1$ \Do
    \li             $\bm{A}^k_{i{c_p}} \gets \bm{A}_{ij_1} \cdot \bm{A}_{ij_2}$
    \li             $c_p \gets c_p + 1$
                \End
            \End
       	\End
\end{codebox}

\subsection{Imperfect CSR Expansion Algorithm}
\label{sec:final-algo}
Now consider how this algorithm might be modified to accept a CSR matrix.
Instead of walking directly down rows of $bm{A}$, we will walk down sections of $\bm{c}$ and $\bm{d}$ partitioned by $\bm{p}$, and instead of inserting polynomial features into $\bm{A}^k$, we will insert column numbers into $\bm{c}^k$ and data elements into $\bm{d}^k$.

\begin{codebox}
\footnotesize
\Procname{$\proc{Incomplete Second Order CSR Polynomial Expansion Algorithm}(\bm{A})$}
    \li $N \gets$ row count of $\bm{A}$
    \li $\bm{p}^k$ $\gets$ vector of size $N+1$
    \li $\bm{p}^k_0 \gets 0$
    \li $nnz^k \gets 0$
    \li \For $i \gets 0 \To N-1$ \Do
    \li     $i_{start} \gets \bm{p}_i$
    \li     $i_{stop} \gets \bm{p}_{i+1}$
    \li     $\bm{c}_i \gets \bm{c}_{i_{start}:i_{stop}}$
    \li     $nnz^k_i \gets \binom{|\bm{c}_i|}{2}$
    \li     $nnz^k \gets nnz^k + nnz^k_i$
    \li     $\bm{p}^k_{i+1} \gets \bm{p}^k_i + nnz^k_i$
        \End
    \zi     
    \zi \Comment Build up the elements of $\bm{p}^k$, $\bm{c}^k$, and $\bm{d}^k$
    \li $\bm{p}^k$ $\gets$ vector of size $N+1$
    \li $\bm{c}^k$ $\gets$ vector of size $nnz^k$
    \li $\bm{d}^k$ $\gets$ vector of size $nnz^k$
    \li $n \gets 0$
    \li \For $i \gets 0 \To N-1$ \Do
    \li     $i_{start} \gets \bm{p}_i$
    \li     $i_{stop} \gets \bm{p}_{i+1}$
    \li     $\bm{c}_i \gets \bm{c}_{i_{start}:i_{stop}}$
    \li     $\bm{d}_i \gets \bm{d}_{i_{start}:i_{stop}}$
    \li     \For $c_1 \gets 0 \To |\bm{c}_i|-1$ \Do
    \li         \For $c_2 \gets c_1 \To |\bm{c}_i|-1$ \Do
%    \li             $j_0 \gets \bm{c}_{c_0}$
%    \li             $j_1 \gets \bm{c}_{c_1}$
    \li             $\bm{d}^k_{n} \gets \bm{d}_{c_0} \cdot \bm{d}_{c_1}$
    \li             $\bm{c}^k_{n} = ?$ \label{li:set_ck}
    \li             $n \gets n + 1$
                \End
            \End
       	\End
\end{codebox}

Note that in the algorithm adapted for CSR matrices, the counter is not reset for each row.
This is because there is only one $\bm{c}^k$ and $\bm{d}^k$ vector for all of $\bm{A}^k$, not one for each row.

However, note line \ref{li:set_ck}.
Given the arbitrary columns involved in a polynomial feature of $\bm{A}_i$, there is no way to determine what the corresponding column of $\bm{A}^k_i$.
In order to achieve a general algorithm, we require a mapping from columns of $\bm{A}$ a column of $\bm{A}^k$.
If there are $D$ columns of $\bm{A}$ and $\binom{D}{k}$ columns of $\bm{A}^k$, this can be accomplished by a bijective mapping of the following form:

\begin{equation}
(j_0, j_1, \dots, j_{k-1}) \rightarrowtail \hspace{-1.9ex} \twoheadrightarrow p_{j_0j_1 \dots i_{k-1}} \in \{0,1,\dots,\binom{D}{k}\} 
\end{equation}

such that $ 0 \le j_0 \le j_1 \le \dots \le j_{k-1} < D$
where $(j_0, j_1, \dots, j_{k-1})$ are elements of $\bm{c}$ and $p_{j_0j_1 \dots i_{k-1}}$ is an element of $\bm{c}^k$. %column indicies of a row vector $\vec{x}$ of an $N \times D$ input matrix, and $p_{i_0i_1 \dots i_{k-1}}$ is a column index into the polynomial expansion vector for $\vec{x}$ where the product of elements corresponding to indices $i_0, i_1, \dots, i_{k-1}$ will be stored.

\section{Construction of Mapping}
For the second degree case, we seek a map from matrix indices $(i, j)$ (with $0 \le i < j < D$ ) to numbers $f(i, j)$ with $0 \le f(i, j) < \frac{D(D-1)}{2}$, one that follows the pattern indicated by 
\begin{align}
\begin{bmatrix}
x & 0 & 1 & 3 \\
x & x & 2 & 4 \\
x & x & x & 5 \\
x & x & x & x
\end{bmatrix}
\label{eq:4x4mat}
\end{align}
where the entry in row $i$, column $j$, displays the value $f(i, j)$. We let $T_2(n) = \frac{1}{2} n(n+1)$ 
be the $n$th triangular number; then in Equation~\ref{eq:4x4mat}, column $j$ (for $j > 0$) contains entries with  
$T_2(j-1) \le e < T_2(j)$; the entry in the $i$th row is just $i + T_2(j-1)$. Thus we have
$
f(i, j) 
= i + T_2(j-1) =  \frac{1}{2}(2i + j^2-j).$
For instance, in column $j = 2$ in our example (the \emph{third} column), the entry in row $i = 1$ is 
$i + T_2(j-1) = 1 + 1 = 2$.

With one-based indexing in both the domain and codomain, the formula above becomes
$f_1(i, j)  = \frac{1}{2}(2i + j^2 - 3j + 2).$

For \emph{polynomial} features, we seek a similar map $g$, one that also handles the case $i = j$. In this case, a similar analysis yields
$ g(i, j) = i + T_2(j) = \frac{1}{2} (2i + j^2 + j + 1).$

To handle \emph{three-way interactions}, we need to map triples of indices in a 3-index array to a flat list, and similarly for higher-order interactions. For this, we'll need the tetrahedral numbers $T_3(n) = \sum_{i=1}^n T_{2}(n) = 
\frac{1}{6}(n^3 + 3n^2 + 2n)$.

For three indices, $i,j,k$, with $0 \le i < j < k < D$, we have a similar recurrence. Calling the mapping $h$, we have 
\begin{align}
h(i,j,k) = i + T_2(j-1) + T_3(k-2);
\end{align}
if we define $T_1(i) = i$, then this has the very regular form
\begin{align}
h(i,j,k) =  T_1(i) + T_2(j-1) + T_3(k-2);
\end{align}
and from this the generalization to higher dimensions is straightforward. The formulas for ``higher triangular numbers'', i.e., those defined by
\begin{align}
T_k(n) &= \sum_{i=1}^n T_{k-1}(n)
\end{align}
for $k > 1$ can be determined inductively.

The explicit formula for 3-way interactions, with zero-based indexing, is 
\begin{align}
h(i, j, k) &= 1 + (i-1) + \frac{(j-1)j}{2} + \\
& \frac{(k-2)^3 + 3(k-2)^2 + 2(k-2)}{6}. 
\end{align}

\subsubsection{Polynomial Features}
For polynommial features, we seek a map from matrix indices $(i, j)$ (with $i \le j$ and $0 \le i < D$) to numbers $g(i, j)$ with $0 \le f(i, j) < \frac{D(D+1)}{2}$, one that follows the pattern indicated by 
\begin{align}
\begin{bmatrix}
 0 & 1 & 3 & 6 \\
 x & 2 & 4 & 7\\
 x & x & 5 & 8 \\
 x & x & x & 9
\end{bmatrix}
\label{eq:4x4mat-poly}
\end{align}
i.e., essentially the same task as before, except that the diagonal is included. One can regard all but the last column of entries in Equation~\ref{eq:4x4mat-poly} as corresponding to the entries in Equation~\ref{eq:4x4mat}, but shifted to the left. Thus the formula for $g(i, j)$ is simply the formula for $f$, shifted by 1, i.e., 
\begin{align}
g(i, j) &= f(i, j+1)  \\
&=  \frac{2i + (j+1)^2-(j+1)}{2}\\
&=  \frac{2i + j^2 + j + 1)}{2}.
\end{align}
Alternatively, we can write this as
\begin{align}
g(i, j) &= i + T_2(j),
\end{align}
\noindent
and get the same result.

\subsubsection{Higher dimensions}
To handle three-way interactions, we need to map triples of indices in a 3-index array to a flat list, and similarly for higher-order interactions.

For three indices, $i,j,k$, with $i < j < k$ and $0 \le i,j,k < D$, we have a similar recurrence. Calling the mapping $h$, we have 
\begin{align}
h(i,j,k) = i + T_2(j-1) + T_3(k-2);
\end{align}
if we define $T_1(i) = i$, then this has the very regular form
\begin{align}
h(i,j,k) =  T_1(i) + T_2(j-1) + T_3(k-2);
\end{align}
and from this, the generalization to higher dimensions is straightforward. The formulas for ``higher triangular numbers'', i.e., those defined by
\begin{align}
T_k(n) &= \sum_{i=1}^n T_{k-1}(n)
\end{align}
for $k > 1$ can be determined inductively. For $k = 3$, the result is 
\begin{align}
T_3(n) &= \sum_{i=1}^n T_{2}(n)\\
&= \frac{n^3 + 3n^2 + 2n}{6},
\end{align}
so that the formula for 3-way interactions, with zero-based indexing, becomes 
\begin{align}
h(i, j, k) &= 1 + (i-1) + \frac{(j-1)j}{2} + \\
& \frac{(k-2)^3 + 3(k-2)^2 + 2(k-2)}{6}. 
\end{align}
\subsubsection{Higher-dimension polynomial features}
For the case where we include the diagonal in higher dimensions, we must shift $j$ by $1$, $k$ by $2$, and so on, and the formula becomes
\begin{align}
\ell(i,j,k) =  T_1(i) + T_2(j) + T_3(k),
\end{align}
with analogous formulas for higher degree polynomial interactions.

\subsection{Inversion}
We not only want to be able to compute the ``flat index'' $f(i,j)$ corresponding to a particular index pair $(i, j)$, we'd like, given a value $N = f(i, j)$, to be able to determine $i$ and $j$. We know, given $N$, that $f(0, j) \le N \le f(j,j-1)$, i.e., that 
\begin{align}
\frac{j^2 - 3j + 2}{2} &\le N \le \frac{2(j-1) + j^2 - 3j + 2}{2}\\
{j^2 - 3j + 2} &\le 2N \le {j^2 - j}.
\end{align}
If we find the (real number) value $j_{*}$ for which $j_{*}^2 - j_{*} = 2N$, then we know that $j = \lfloor j_{*}\rfloor$. Finding $j_{*}$ amounts to solving a quadratic, whence
\begin{align}
j_{*} = \frac{1 \pm \sqrt{1 + 8N}}{2}.
\end{align}
Clearly the larger value, 
\begin{align}
j_{*} = \frac{1 + \sqrt{1 + 8N}}{2},
\end{align}
is the one we seek; we set $j = \lfloor j_{*}\rfloor$, and compute $i = N - \frac{j^2 - j}{2}$. 
In practice, repeatedly computing square roots may be annoying; instead, when we compute $N = f(i,j)$, we can record the values of $i$ and $j$ along with the interaction features, making inversion a simple matter of looking up these values.

Similarly, for polynomial features, to invert 
\begin{align}
g(i, j) &=  \frac{2i + j^2 + j + 1)}{2}
\end{align}
we know that if $N = g(i,j)$, then $g(0,j) \le N \le g(j,j)$, which tells us that 
\begin{align}
2N \le j^2 + 3j + 1, \text{ so}\\
0 \le j^2 + 3j + (1-N^2)
\end{align}. 
Solving for $j$ when this last inequality is an equality, we get 
\begin{align}
j_{*} = \frac{-3 \pm \sqrt{ 5 - N^2 }}{2}; 
\end{align}
Once again, we set $j = \lfloor j_{*}\rfloor$, and compute $i = N - \frac{j^2 + j + 1}{2}$.

Inverting the higher-order formulas, which involve cubics, quartics, etc., is infeasible, but storing the inverse map as described above may be practical, especially for sparse feature sets.

\section{Final CSR Expansion Algorithm}
With the mapping from columns of $\bm{A}$ to a column of $\bm{A}^k$, we can now write the final form of the innermost loop of the algorithm from \ref{sec:final-algo}.
Let the mapping for $k=2$ be denoted $PolyCol_2$.
Then the innermost loop becomes:

\begin{codebox}
\footnotesize
    \zi         \For $c_2 \gets c_1 \To |\bm{c}_i|-1$ \Do
    \zi             $j_0 \gets \bm{c}_{c_0}$
    \zi             $j_1 \gets \bm{c}_{c_1}$
    \zi             $c_p \gets PolyCol_2(j_0, j_1)$
    \zi             $\bm{d}^k_{n} \gets \bm{d}_{c_0} \cdot \bm{d}_{c_1}$
    \zi             $\bm{c}^k_{n} = c_p$
    \zi             $n \gets n + 1$
                \End
\end{codebox}

\section{Time Complexity}
\subsection{Analytical}
Calculating $k$-degree polynomial features via our method for a row $\bm{a}_i$ of a matrix of dimensionality $D$ and density $d$ requires $\binom{dD}{k}$ (with repetition) products, assuming the nonzero elements of $\bm{A}$ are uniformly distributed.
The complexity of the algorithm, for fixed $k \ll dD$, is therefore
\begin{align}
O\left(\binom{dD+k-1}{k}\right) & = O\left(\frac{(dD+k-1)!}{k!(dD-1)!}\right)\\
& = O\left(\frac{(dD+k-1)(dD+k-2) \dots (dD)}{k!}\right)\\
& = O\left((dD+k-1)(dD+k-2) \dots (dD)\right) \mbox{ for } k \ll dD\\
& = O\left(d^kD^k\right)
\end{align}

\subsection{Empirical}
\section{Conclusion}

\bibliography{sparse_poly}
\bibliographystyle{iclr2016_workshop}

\end{document}

### 2016-11-01 paper/Final/sparse_poly.tex
fixed bibliography

@@
 
-\usepackage{amsmath}
 \usepackage{amssymb}
@@
 \usepackage{clrscode3e}
-\usepackage{esvect}
-\usepackage[T1]{fontenc}
+\usepackage{float}
+\usepackage{mathtools}
+
+%icrl 2017
+\usepackage{iclr2017_conference,times}
 \usepackage{hyperref}
-\usepackage{mathtools}
-\usepackage{times}
 \usepackage{url}
@@
 \author{
-  Nystrom, Andrew\\
-  \texttt{awnystrom@gmail.com}
+  Nystrom, Andrew \\
+  \texttt{nystrom@google.com}
   \and
-  Hughes, John\\
+  Hughes, John \\
   \texttt{jfh@cs.brown.edu}
@@
 
-%\newcommand{\fix}{\marginpar{FIX}}
-%\newcommand{\new}{\marginpar{NEW}}
-\newcommand{\matr}[1]{\bm{#1}}
+\newcommand{\fix}{\marginpar{FIX}}
+\newcommand{\new}{\marginpar{NEW}}
 
@@
 We provide an algorithm for polynomial feature expansion that operates directly on and produces a compressed sparse row matrix without any densification.
-For a vector of dimension $D$ and density $d$, the algorithm has time and space complexity $O(d^kD^k)$ where $k$ is the polynomial order, which is an improvement by a factor $d^k$ over the standard method.
+For a vector of dimension $D$, density $d$, and degree $k$ the algorithm has time and space complexity $O(d^kD^k)$ where $k$ is the polynomial order, which is an improvement by a factor $d^k$ over the standard method.
 \end{abstract}
-
-% this is a test-change for Spike to see whether Git seems to work for him. Ignore.
 
@@
 \section{Preliminaries}
-A matrix is uppercase and bold, like this: $\bm{A}$.
-In this the context of this paper, row vectors are discussed, but not collum vectors.
-Therefore, the distinction between the two is unimportant.
-A row vector $\bm{a}_i$ is the $i^{th}$ row of $\bm{A}$ and $\bm{a}$, without a subscript, is a vector not necessarily related to $\bm{A}$.
+A matrix $\bm{A}$ is uppercase and bold.
+A row vector $\bm{a}_i$ has a subscript and is the $i^{th}$ row of $\bm{A}$, whereas $\bm{a}$, without a subscript, is a vector not necessarily related to $\bm{A}$.
 
@@
 The values of $\bm{p}$ index $\bm{c}$ and $\bm{d}$ and is itself indexed by rows of $\bm{A}$ such that the nonzero columns of $\bm{a}_i$ are $\bm{c}_{\bm{p}_i:\bm{p}_{i+1}}$ and the nonzero data elements are $\bm{d}_{\bm{p}_i:\bm{p}_{i+1}}$.
+Since only nonzero elements of each row are held, the dimensionality must also be stored since it cannot be implicitly gotten from the length of $\bm{c}$.
+
+Scalers, vectors, and matrices are often referenced with the superscript $k$.
+This is not to be interpreted as an exponent, but to indicate that it is the analogous aspect of that which procedes it, but in its polynomial expansion form.
+For example, $\bm{c}^k$ is the vector that holds columns for nonzero values in $\bm{A}$'s polynomial feature expansion CSR representation.
+
+For simplicity, all polynomial expansions are assumed to be of the second degree.
+We do provide an algorithm for third degree expansions and derive the big O time complexity of the general case.
 
@@
 
-Note that in the algorithm adapted for CSR matrices, the counter is not reset for each row.
-This is because there is only one $\bm{c}^k$ and $\bm{d}^k$ vector for all of $\bm{A}^k$, not one for each row.
+The crux of the problem can be found on line \ref{li:set_ck}.
+Given the arbitrary columns involved in a polynomial feature of $\bm{A}_i$, there is no way to determine what the corresponding column of $\bm{A}^k_i$.
+We cannot simply reset a counter for each row as we did in the dense algorithm.
+This is because only columns corresponding with nonzero values are stored.
+Any time a column is implicitly skipped that would have held a zero value, the counter would err.
 
-However, note line \ref{li:set_ck}.
-Given the arbitrary columns involved in a polynomial feature of $\bm{A}_i$, there is no way to determine what the corresponding column of $\bm{A}^k_i$.
 In order to achieve a general algorithm, we require a mapping from columns of $\bm{A}$ a column of $\bm{A}^k$.
@@
 \section{Construction of Mapping}
+Within this section, $i$, $j$, and $k$ denote column indices.
 For the second degree case, we seek a map from matrix indices $(i, j)$ (with $0 \le i < j < D$ ) to numbers $f(i, j)$ with $0 \le f(i, j) < \frac{D(D-1)}{2}$, one that follows the pattern indicated by
@@
 
-\subsubsection{Polynomial Features}
-For polynommial features, we seek a map from matrix indices $(i, j)$ (with $i \le j$ and $0 \le i < D$) to numbers $g(i, j)$ with $0 \le f(i, j) < \frac{D(D+1)}{2}$, one that follows the pattern indicated by
-\begin{align}
-\begin{bmatrix}
- 0 & 1 & 3 & 6 \\
- x & 2 & 4 & 7\\
- x & x & 5 & 8 \\
- x & x & x & 9
-\end{bmatrix}
-\label{eq:4x4mat-poly}
-\end{align}
-i.e., essentially the same task as before, except that the diagonal is included.
-One can regard all but the last column of entries in Equation~\ref{eq:4x4mat-poly} as corresponding to the entries in Equation~\ref{eq:4x4mat}, but shifted to the left.
-Thus the formula for $g(i, j)$ is simply the formula for $f$, shifted by 1, i.e.,
-\begin{align}
-g(i, j) &= f(i, j+1)  \\
-&=  \frac{2i + (j+1)^2-(j+1)}{2}\\
-&=  \frac{2i + j^2 + j + 1)}{2}.
-\end{align}
-Alternatively, we can write this as
-\begin{align}
-g(i, j) &= i + T_2(j),
-\end{align}
-\noindent
-and get the same result.
-
-\subsubsection{Higher dimensions}
-To handle three-way interactions, we need to map triples of indices in a 3-index array to a flat list, and similarly for higher-order interactions.
-
-For three indices, $i,j,k$, with $i < j < k$ and $0 \le i,j,k < D$, we have a similar recurrence. Calling the mapping $h$, we have 
-\begin{align}
-h(i,j,k) = i + T_2(j-1) + T_3(k-2);
-\end{align}
-if we define $T_1(i) = i$, then this has the very regular form
-\begin{align}
-h(i,j,k) =  T_1(i) + T_2(j-1) + T_3(k-2);
-\end{align}
-and from this, the generalization to higher dimensions is straightforward. The formulas for ``higher triangular numbers'', i.e., those defined by
-\begin{align}
-T_k(n) &= \sum_{i=1}^n T_{k-1}(n)
-\end{align}
-for $k > 1$ can be determined inductively. For $k = 3$, the result is 
-\begin{align}
-T_3(n) &= \sum_{i=1}^n T_{2}(n)\\
-&= \frac{n^3 + 3n^2 + 2n}{6},
-\end{align}
-so that the formula for 3-way interactions, with zero-based indexing, becomes 
-\begin{align}
-h(i, j, k) &= 1 + (i-1) + \frac{(j-1)j}{2} + \\
-& \frac{(k-2)^3 + 3(k-2)^2 + 2(k-2)}{6}. 
-\end{align}
-\subsubsection{Higher-dimension polynomial features}
-For the case where we include the diagonal in higher dimensions, we must shift $j$ by $1$, $k$ by $2$, and so on, and the formula becomes
-\begin{align}
-\ell(i,j,k) =  T_1(i) + T_2(j) + T_3(k),
-\end{align}
-with analogous formulas for higher degree polynomial interactions.
-
-\subsection{Inversion}
-We not only want to be able to compute the ``flat index'' $f(i,j)$ corresponding to a particular index pair $(i, j)$, we'd like, given a value $N = f(i, j)$, to be able to determine $i$ and $j$.
-We know, given $N$, that $f(0, j) \le N \le f(j,j-1)$, i.e., that
-\begin{align}
-\frac{j^2 - 3j + 2}{2} &\le N \le \frac{2(j-1) + j^2 - 3j + 2}{2}\\
-{j^2 - 3j + 2} &\le 2N \le {j^2 - j}.
-\end{align}
-If we find the (real number) value $j_{*}$ for which $j_{*}^2 - j_{*} = 2N$, then we know that $j = \lfloor j_{*}\rfloor$. Finding $j_{*}$ amounts to solving a quadratic, whence
-\begin{align}
-j_{*} = \frac{1 \pm \sqrt{1 + 8N}}{2}.
-\end{align}
-Clearly the larger value, 
-\begin{align}
-j_{*} = \frac{1 + \sqrt{1 + 8N}}{2},
-\end{align}
-is the one we seek; we set $j = \lfloor j_{*}\rfloor$, and compute $i = N - \frac{j^2 - j}{2}$. 
-In practice, repeatedly computing square roots may be annoying; instead, when we compute $N = f(i,j)$, we can record the values of $i$ and $j$ along with the interaction features, making inversion a simple matter of looking up these values.
-
-Similarly, for polynomial features, to invert 
-\begin{align}
-g(i, j) &=  \frac{2i + j^2 + j + 1)}{2}
-\end{align}
-we know that if $N = g(i,j)$, then $g(0,j) \le N \le g(j,j)$, which tells us that 
-\begin{align}
-2N \le j^2 + 3j + 1, \text{ so}\\
-0 \le j^2 + 3j + (1-N^2)
-\end{align}. 
-Solving for $j$ when this last inequality is an equality, we get 
-\begin{align}
-j_{*} = \frac{-3 \pm \sqrt{ 5 - N^2 }}{2}; 
-\end{align}
-Once again, we set $j = \lfloor j_{*}\rfloor$, and compute $i = N - \frac{j^2 + j + 1}{2}$.
-
-Inverting the higher-order formulas, which involve cubics, quartics, etc., is infeasible, but storing the inverse map as described above may be practical, especially for sparse feature sets.
-
 \section{Final CSR Expansion Algorithm}
 With the mapping from columns of $\bm{A}$ to a column of $\bm{A}^k$, we can now write the final form of the innermost loop of the algorithm from \ref{sec:final-algo}.
-Let the mapping for $k=2$ be denoted $PolyCol_2$.
+Let the mapping for $k=2$ be denoted $PolyCol^2$.
 Then the innermost loop becomes:
@@
     \zi             $j_1 \gets \bm{c}_{c_1}$
-    \zi             $c_p \gets PolyCol_2(j_0, j_1)$
+    \zi             $c_p \gets PolyCol^2(j_0, j_1)$
     \zi             $\bm{d}^k_{n} \gets \bm{d}_{c_0} \cdot \bm{d}_{c_1}$
@@
 
+The algorithm can be generalized to higher degrees by simply adding more nested loops and using higher order mappings.
+
 \section{Time Complexity}
 \subsection{Analytical}
-Calculating $k$-degree polynomial features via our method for a row $\bm{a}_i$ of a matrix of dimensionality $D$ and density $d$ requires $\binom{dD}{k}$ (with repetition) products, assuming the nonzero elements of $\bm{A}$ are uniformly distributed.
+
+Calculating $k$-degree polynomial features via our method for a vector of dimensionality $D$ and density $d$ requires $\binom{dD}{k}$ (with repetition) products.
 The complexity of the algorithm, for fixed $k \ll dD$, is therefore
@@
 \subsection{Empirical}
+To demonstrate how our algorithm scales with the density of a matrix, we compare it to the traditional polynomial expansion algorithm in the popular machine library scikit-learn \cite{scikit-learn} in the task of generating second degree polynomial expansions.
+Matrices of size $100 \times 5000$ were randomly generated with densities of $0.2$, $0.4$, $0.6$, $0.8$, and $1.0$.
+Thirty matrices of each density were randomly generated, and the mean times (gray) of each algorithm were plotted.
+The red or blue width around the mean marks the third standard deviation from the mean.
+
+The standard algorithm's runtime stays constant no matter the density of the matrix.
+This is because it does not avoid products that result in zero, but simply multiplies all second order combinations of features.
+Our algorithm scales quadratically with respect to the density.
+If the task were third degree expansions rather than second, the plot would show cubic scaling.
+
+The fact that our algorithm is approximately $6.5$ times faster than the scikit-learn algorithm on $100 \times 5000$ matrices that are entirely dense is likely an language implementation difference.
+What matters is that the time of our algorithm increases quadratically with respect to the density in accordance with the big O analysis.
+
+\begin{figure}[H]
+    \centering
+    \includegraphics[scale=0.6]{density_vs_time}
+    \caption{This plot shows how our algorithm (bottom) scales with the density of a matrix compared to a traditional polynomial feature expansion method (top).
+             The task was a second degree expansion, which is why the time of our algorithm scales quadratically with the density.}
+    \label{fig:benchmark}
+\end{figure}
+
 \section{Conclusion}
+We have developed an algorithm for performing polynomial feature expansions on CSR matrices that scales polynomially with respect to the density of the matrix.
+While the areas within machine learning this work touches is not the flavor of the day, they are workhorses of industry and will likely save much coal from being unnecessarily burnt.
 
 \bibliography{sparse_poly}
-\bibliographystyle{iclr2016_workshop}
+\bibliographystyle{iclr2017_conference}
 

### 2016-11-04 paper/Final/sparse_poly.tex
typos and minor changes

@@
 \usepackage{float}
+\usepackage{footmisc}
 \usepackage{mathtools}
@@
 
+\newcommand*\samethanks[1][\value{footnote}]{\footnotemark[#1]}
+\renewcommand{\thefootnote}{\fnsymbol{thanks}}
+
 \author{
   Nystrom, Andrew \\
-  \texttt{nystrom@google.com}
+  \texttt{awnystrom@gmail.com}
+  \thanks{Now at Google}
+  \ 
+  \thanks{The authors contributed equally important and fundamental aspects of this work.}
   \and
@@
   \texttt{jfh@cs.brown.edu}
+  \footnotemark[2]
 }
@@
 We provide an algorithm for polynomial feature expansion that operates directly on and produces a compressed sparse row matrix without any densification.
-For a vector of dimension $D$, density $d$, and degree $k$ the algorithm has time and space complexity $O(d^kD^k)$ where $k$ is the polynomial order, which is an improvement by a factor $d^k$ over the standard method.
+For a vector of dimension $D$, density $d$, and degree $k$ the algorithm has time complexity $O(d^kD^k)$ where $k$ is the polynomial order, which is an improvement by a factor $d^k$ over the standard method.
 \end{abstract}
@@
 Since $0 \le d \le 1$, our algorithm is a significant improvement.
-While the algorithm we lay out is uses CSR matrices, it could be slightly modified to operate on other sparse formats.
+While the algorithm we lay out uses CSR matrices, it could be modified to operate on other sparse formats.
 
@@
 For simplicity, all polynomial expansions are assumed to be of the second degree.
-We do provide an algorithm for third degree expansions and derive the big O time complexity of the general case.
+We do provide an algorithm for third degree expansions, and derive the big O time complexity of the general case.
+We have also developed an algorithm for second and third degree interaction features (combinations without repetition), which can be found in the implementation.
 
@@
     \li     $\bm{c}_i \gets \bm{c}_{i_{start}:i_{stop}}$
-    \li     $nnz^k_i \gets \binom{|\bm{c}_i|}{2}$
+    \li     $nnz^k_i \gets \binom{|\bm{c}_i|}{2}$ \label{li:row_nnz_count}
     \li     $nnz^k \gets nnz^k + nnz^k_i$
@@
 \begin{equation}
-(j_0, j_1, \dots, j_{k-1}) \rightarrowtail \hspace{-1.9ex} \twoheadrightarrow p_{j_0j_1 \dots i_{k-1}} \in \{0,1,\dots,\binom{D}{k}\} 
+(j_0, j_1, \dots, j_{k-1}) \rightarrowtail \hspace{-1.9ex} \twoheadrightarrow p_{j_0j_1 \dots i_{k-1}} \in \{0,1,\dots,\binom{D}{k}-1\} 
 \end{equation}
@@
 With the mapping from columns of $\bm{A}$ to a column of $\bm{A}^k$, we can now write the final form of the innermost loop of the algorithm from \ref{sec:final-algo}.
-Let the mapping for $k=2$ be denoted $PolyCol^2$.
+Let the mapping for $k=2$ be denoted $h^2$.
 Then the innermost loop becomes:
@@
 
-The algorithm can be generalized to higher degrees by simply adding more nested loops and using higher order mappings.
+The algorithm can be generalized to higher degrees by simply adding more nested loops, using higher order mappings, modifying the output dimensionality, and adjusting the counting of nonzero polynomial features in line \ref{li:row_nnz_count}.
+%For interaction features, the interaction mappings can be used in lieu of the polynomial mappings with the additional change of the output dimensionality and the number of nonzero features in each row (combinations without repetition instead of with).
 
@@
 The red or blue width around the mean marks the third standard deviation from the mean.
+The time to densify the input to the standard algorithm was not counted.
 
@@
 
-The fact that our algorithm is approximately $6.5$ times faster than the scikit-learn algorithm on $100 \times 5000$ matrices that are entirely dense is likely an language implementation difference.
+The fact that our algorithm is approximately $6.5$ times faster than the scikit-learn algorithm on $100 \times 5000$ matrices that are entirely dense is likely a language implementation difference.
 What matters is that the time of our algorithm increases quadratically with respect to the density in accordance with the big O analysis.
@@
 We have developed an algorithm for performing polynomial feature expansions on CSR matrices that scales polynomially with respect to the density of the matrix.
-While the areas within machine learning this work touches is not the flavor of the day, they are workhorses of industry and will likely save much coal from being unnecessarily burnt.
+While the areas within machine learning this work touches are not en vogue, they are workhorses of industry.
+This improvement could therefore spare the burning of much fossil fuel.
 

### 2017-02-19 paper/Final/sparse_poly_icml.tex
icml

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%% ICML 2017 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Use the following line _only_ if you're still using LaTeX 2.09.
%\documentstyle[icml2017,epsf,natbib]{article}
% If you rely on Latex2e packages, like most moden people use this:
\documentclass{article}

% use Times
\usepackage{times}
% For figures
\usepackage{graphicx} % more modern
%\usepackage{epsfig} % less modern
\usepackage{subfigure}

% For citations
\usepackage{natbib}

% For algorithms
\usepackage{algorithm}
\usepackage{algorithmic}

\usepackage{amssymb}
\usepackage{bm}
\usepackage{clrscode3e}
\usepackage{float}
\usepackage{footmisc}
\usepackage{mathtools}
\usepackage{amsfonts}
\usepackage{amssymb}

% As of 2011, we use the hyperref package to produce hyperlinks in the
% resulting PDF.  If this breaks your system, please commend out the
% following usepackage line and replace \usepackage{icml2017} with
% \usepackage[nohyperref]{icml2017} above.
\usepackage{hyperref}

% Packages hyperref and algorithmic misbehave sometimes.  We can fix
% this with the following command.
\newcommand{\theHalgorithm}{\arabic{algorithm}}

% Employ the following version of the ``usepackage'' statement for
% submitting the draft version of the paper for review.  This will set
% the note in the first column to ``Under review.  Do not distribute.''
\usepackage{icml2017}

% Employ this version of the ``usepackage'' statement after the paper has
% been accepted, when creating the final version.  This will set the
% note in the first column to ``Proceedings of the...''
%\usepackage[accepted]{icml2017}

% The \icmltitle you define below is probably too long as a header.
% Therefore, a short form for the running title is supplied here:
\icmltitlerunning{Leveraging Matrix Sparsity in Polynomial Expansions}

\begin{document}

\twocolumn[
\icmltitle{Leveraging Matrix Sparsity in Polynomial Expansions}

% It is OKAY to include author information, even for blind
% submissions: the style file will automatically remove it for you
% unless you've provided the [accepted] option to the icml2017
% package.

% list of affiliations. the first argument should be a (short)
% identifier you will use later to specify author affiliations
% Academic affiliations should list Department, University, City, Region, Country
% Industry affiliations should list Company, City, Region, Country

% you can specify symbols, otherwise they are numbered in order
% ideally, you should not use this facility. affiliations will be numbered
% in order of appearance and this is the preferred way.
\icmlsetsymbol{equal}{*}

\begin{icmlauthorlist}
\icmlauthor{Aeiau Zzzz}{equal,to}
\icmlauthor{Bauiu C.~Yyyy}{equal,to,goo}
\icmlauthor{Cieua Vvvvv}{goo}
\icmlauthor{Iaesut Saoeu}{ed}
\icmlauthor{Fiuea Rrrr}{to}
\icmlauthor{Tateu H.~Yasehe}{ed,to,goo} 
\icmlauthor{Aaoeu Iasoh}{goo}
\icmlauthor{Buiui Eueu}{ed}
\icmlauthor{Aeuia Zzzz}{ed}
\icmlauthor{Bieea C.~Yyyy}{to,goo}
\icmlauthor{Teoau Xxxx}{ed}
\icmlauthor{Eee Pppp}{ed}
\end{icmlauthorlist}

\icmlaffiliation{to}{University of Torontoland, Torontoland, Canada}
\icmlaffiliation{goo}{Googol ShallowMind, New London, Michigan, USA}
\icmlaffiliation{ed}{University of Edenborrow, Edenborrow, United Kingdom}

\icmlcorrespondingauthor{Cieua Vvvvv}{c.vvvvv@googol.com}
\icmlcorrespondingauthor{Eee Pppp}{ep@eden.co.uk}

% You may provide any keywords that you 
% find helpful for describing your paper; these are used to populate 
% the "keywords" metadata in the PDF but will not be shown in the document
\icmlkeywords{boring formatting information, machine learning, ICML}

\vskip 0.3in
]

% this must go after the closing bracket ] following \twocolumn[ ...

% This command actually creates the footnote in the first column
% listing the affiliations and the copyright notice.
% The command takes one argument, which is text to display at the start of the footnote.
% The \icmlEqualContribution command is standard text for equal contribution.
% Remove it (just {}) if you do not need this facility.

%\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
\printAffiliationsAndNotice{\icmlEqualContribution} % otherwise use the standard text.
%\footnotetext{hi}

\begin{abstract}
We provide an algorithm for polynomial feature expansion that both operates on and produces a compressed sparse row matrix without any densification.
For a vector of dimension $D$, density $d$, and degree $k$ the algorithm has time complexity $O(d^kD^k)$ where $k$ is the polynomial-feature order; this  is an improvement by a factor $d^k$ over the standard method.
\end{abstract}

\section{Introduction}

Feature mappings are an intra-instance transformation, usually referred to as $\phi(\vec{x})$, that map instance vectors to higher dimensional spaces in which they are more linearly separable, allowing linear models to capture non-linearities \cite{yuan2012recent}.
A widely used feature mapping is the polynomial expansion, which generates a new feature for each product of all $k$-order combinations in the original feature space.
A $k$-order polynomial expansion of the feature space allows a linear model to learn polynomial relationships between dependent and independant variables.
This mapping was first utilized in a published experiment by Joseph Diaz Gergonne in 1815 \cite{gergonne1974application, smith1918standard}.

While other methods for capturing nonlinearities have been developed, such as kernels (the direct offspring of feature mappings), trees, generative models, and neural networks, feature mappings are still a popular tool \cite{barker200114, chang2010training, shaw2006intellectual}.
The instance-independant nature of feature mappings allows them to pair well with linear parameter estimation techniques such as stochastic gradient descent, making them a candidate for certain types of large-scale machine learning problems when $D < N$.

The compressed sparse row (CSR) matrix format \cite(saad1994sparskit) is widely used \cite{liu2012sparse, bulucc2009parallel, bell2008efficient, white1997improving} and supported \cite{eigenweb, bastien2012theano, koenker2003sparsem}, and is considered the standard data structure for sparse, computationally heavy tasks.
However, polynomial expansions cannot be performed directly on sparse CSR matrices, nor on any sparse matrix format, without intermediate densification steps.
This densification not only adds extra overhead, but wastefully computes combinations of features that have a product of zero, which are then discarded during conversion into a sparse format.

We provide an algorithm that allows CSR matrices to be the input of a polynomial feature expansion without any densification.
The algorithm leverages the CSR format to only compute products of features that result in nonzero values.
This exploits the sparsity of the data to achieve an improved time complexity of $O(d^kD^k)$ on each vector of the matrix where $k$ is the degree of the expansion, $D$ is the dimensionality, and $d$ is the density.
The standard algorithm has time complexity $O(D^k)$.
Since $0 \le d \le 1$, our algorithm is a significant improvement.
While the algorithm we describe uses CSR matrices, it could be modified to operate on other sparse formats.

\section{Preliminaries}
Matrices are denoted by uppercase bold letters thus: $\bm{A}$. 
The $i$the row of $\bm{A}$ is written $\bm{a}_i$. All vectors are written in bold, and  $\bm{a}$, with no subscript, is a vector.

A compressed sparse row (CSR) matrix representation of an $r$-row matrix $\bm{A}$ consists of three vectors: $\bm{c}$, $\bm{d}$, and $\bm{p}$ and a single number: the number of columns of $\bm{A}$. The vectors
$\bm{c}$ and $\bm{d}$ contain the same number of elements, and hold the column indices and data values, respectively, of all nonzero elements of $\bm{A}$.
The vector $\bm{p}$ has $r$ entries. The values in $\bm{p}$ index both $\bm{c}$ and $\bm{d}$. The $i$th entry $\bm{p}_i$ of $\bm{p}$ tells where
the data describing nonzero columns of $\bm{a}_i$ are within the other two vectors: $\bm{c}_{\bm{p}_i:\bm{p}_{i+1}}$ contain the column indices of those entries; $\bm{d}_{\bm{p}_i:\bm{p}_{i+1}}$ contain the entries themselves.
Since only nonzero elements of each row are held, the overall number of columns of $\bm{A}$  must also be stored, since it cannot be derived
from the other data.

Scalars, vectors, and matrices are often referenced with the superscript $k$.
This is not to be interpreted as an exponent, but to indicate that it is the analogous aspect of that which procedes it, but in its polynomial expansion form.
For example, $\bm{c}^2$ is the vector that holds columns for nonzero values in $\bm{A}$'s quadratic feature expansion CSR representation.

For simplicity in the presentation, we work with polynomial expansions of degree 2, but continue to use the exponent $k$ to show how the ideas apply in the general case.

We do provide an algorithm for third degree expansions, and derive the big-O time complexity of the general case.

We have also developed an algorithm for second and third degree interaction features (combinations without repetition), which can be found in the implementation.

\section{Motivation}
In this section, we present a strawman algorithm for computing polynomial feature expansions on dense matrices.
We then modify the algorithm  slightly to operate on a CSR matrix, in order to expose its infeasibility in that context.
We then show how the algorithm would be feasible with an added component, which we then derive in the following section.

\subsection{Dense Expansion Algorithm}
A natural way to calculate polynomial features for a matrix $\bm{A}$ is to walk down its rows and, for each row, take products of all $k$-combinations of elements.
To determine in which column of $\bm{A}^k_i$ products of elements in $\bm{A}_i$ belong, a simple counter can be set to zero for each row of $\bm{A}$ and incremented efter each polynomial feature is generated.
This counter gives the column of $\bm{A}^k_i$ into which each expansion feature belongs.

\begin{codebox}
\footnotesize
\Procname{$\proc{Dense Second Order Algorithm}(\bm{A})$}
    \li $N \gets$ row count of $\bm{A}$
    \li $D \gets$ column count of $\bm{A}$
    \li $\bm{A}^k$ $\gets$ empty $N \times \binom{D}{2}$ matrix
    \li \For $i \gets 0 \To N-1$ \Do
    \li     $c_p \gets 0$
    \li     \For $j_1 \gets 0 \To D-1$ \Do
    \li         \For $j_2 \gets j_1 \To D-1$ \Do
    \li             $\bm{A}^k_{i{c_p}} \gets \bm{A}_{ij_1} \cdot \bm{A}_{ij_2}$
    \li             $c_p \gets c_p + 1$
                \End
            \End
       	\End
\end{codebox}

\subsection{Imperfect CSR Expansion Algorithm}
\label{sec:final-algo}
Now consider how this algorithm might be modified to accept a CSR matrix.
Instead of walking directly down rows of $\bm{A}$, we will walk down sections of $\bm{c}$ and $\bm{d}$ partitioned by $\bm{p}$, and instead of inserting polynomial features into $\bm{A}^k$, we will insert column numbers into $\bm{c}^k$ and data elements into $\bm{d}^k$.

\begin{codebox}
\footnotesize
\Procname{$\proc{Incomplete Sparse Second Order ALgorithm}(\bm{A})$}
    \li $N \gets$ row count of $\bm{A}$
    \li $\bm{p}^k$ $\gets$ vector of size $N+1$
    \li $\bm{p}^k_0 \gets 0$
    \li $nnz^k \gets 0$
    \li \For $i \gets 0 \To N-1$ \Do
    \li     $i_{start} \gets \bm{p}_i$
    \li     $i_{stop} \gets \bm{p}_{i+1}$
    \li     $\bm{c}_i \gets \bm{c}_{i_{start}:i_{stop}}$
    \li     $nnz^k_i \gets \binom{|\bm{c}_i|}{2}$ \label{li:row_nnz_count}
    \li     $nnz^k \gets nnz^k + nnz^k_i$
    \li     $\bm{p}^k_{i+1} \gets \bm{p}^k_i + nnz^k_i$
        \End
    \zi     
    \zi \Comment Build up the elements of $\bm{p}^k$, $\bm{c}^k$, and $\bm{d}^k$
    \li $\bm{p}^k$ $\gets$ vector of size $N+1$
    \li $\bm{c}^k$ $\gets$ vector of size $nnz^k$
    \li $\bm{d}^k$ $\gets$ vector of size $nnz^k$
    \li $n \gets 0$
    \li \For $i \gets 0 \To N-1$ \Do
    \li     $i_{start} \gets \bm{p}_i$
    \li     $i_{stop} \gets \bm{p}_{i+1}$
    \li     $\bm{c}_i \gets \bm{c}_{i_{start}:i_{stop}}$
    \li     $\bm{d}_i \gets \bm{d}_{i_{start}:i_{stop}}$
    \li     \For $c_1 \gets 0 \To |\bm{c}_i|-1$ \Do
    \li         \For $c_2 \gets c_1 \To |\bm{c}_i|-1$ \Do
%    \li             $j_0 \gets \bm{c}_{c_0}$
%    \li             $j_1 \gets \bm{c}_{c_1}$
    \li             $\bm{d}^k_{n} \gets \bm{d}_{c_0} \cdot \bm{d}_{c_1}$
    \li             $\bm{c}^k_{n} = ?$ \label{li:set_ck}
    \li             $n \gets n + 1$
                \End
            \End
       	\End
\end{codebox}

The crux of the problem is at line \ref{li:set_ck}.
Given the arbitrary columns involved in a polynomial feature of $\bm{A}_i$, we need to determine the corresponding column of $\bm{A}^k_i$.
We cannot simply reset a counter for each row as we did in the dense algorithm,  because only columns corresponding to nonzero values are stored.
Any time a column that would have held a zero value is implicitly skipped, the counter would err.

To develop a general algorithm, we require a mapping from columns of $\bm{A}$ to a column of $\bm{A}^k$.
If there are $D$ columns of $\bm{A}$ and $\binom{D}{k}$ columns of $\bm{A}^k$, this can be accomplished by a bijective mapping of the following form:

\begin{equation}
(j_0, j_1, \dots, j_{k-1}) \rightarrowtail \hspace{-1.9ex} \twoheadrightarrow p_{j_0j_1 \dots i_{k-1}} \in \{0,1,\dots,\binom{D}{k}-1\} 
\end{equation}

such that $ 0 \le j_0 \le j_1 \le \dots \le j_{k-1} < D$
where $(j_0, j_1, \dots, j_{k-1})$ are elements of $\bm{c}$ and $p_{j_0j_1 \dots i_{k-1}}$ is an element of $\bm{c}^k$. %column indicies of a row vector $\vec{x}$ of an $N \times D$ input matrix, and $p_{i_0i_1 \dots i_{k-1}}$ is a column index into the polynomial expansion vector for $\vec{x}$ where the product of elements corresponding to indices $i_0, i_1, \dots, i_{k-1}$ will be stored.

\section{Construction of Mapping}
Within this section, $i$, $j$, and $k$ denote column indices.
For the second degree case, we seek a map from matrix indices $(i, j)$ (with $0 \le i < j < D$ ) to numbers $f(i, j)$ with $0 \le f(i, j) < \frac{D(D-1)}{2}$, one that follows the pattern indicated by 
\begin{align}
\begin{bmatrix}
x & 0 & 1 & 3 \\
x & x & 2 & 4 \\
x & x & x & 5 \\
x & x & x & x
\end{bmatrix}
\label{eq:4x4mat}
\end{align}
where the entry in row $i$, column $j$, displays the value $f(i, j)$. We let $T_2(n) = \frac{1}{2} n(n+1)$ 
be the $n$th triangular number; then in Equation~\ref{eq:4x4mat}, column $j$ (for $j > 0$) contains entries with  
$T_2(j-1) \le e < T_2(j)$; the entry in the $i$th row is just $i + T_2(j-1)$. Thus we have
$
f(i, j) 
= i + T_2(j-1) =  \frac{1}{2}(2i + j^2-j).$
For instance, in column $j = 2$ in our example (the \emph{third} column), the entry in row $i = 1$ is 
$i + T_2(j-1) = 1 + 1 = 2$.

With one-based indexing in both the domain and codomain, the formula above becomes
$f_1(i, j)  = \frac{1}{2}(2i + j^2 - 3j + 2).$

For \emph{polynomial} features, we seek a similar map $g$, one that also handles the case $i = j$. In this case, a similar analysis yields
$ g(i, j) = i + T_2(j) = \frac{1}{2} (2i + j^2 + j + 1).$

To handle \emph{three-way interactions}, we need to map triples of indices in a 3-index array to a flat list, and similarly for higher-order interactions. For this, we'll need the tetrahedral numbers $T_3(n) = \sum_{i=1}^n T_{2}(n) = 
\frac{1}{6}(n^3 + 3n^2 + 2n)$.

For three indices, $i,j,k$, with $0 \le i < j < k < D$, we have a similar recurrence. Calling the mapping $h$, we have 
\begin{align}
h(i,j,k) = i + T_2(j-1) + T_3(k-2);
\end{align}
if we define $T_1(i) = i$, then this has the very regular form
\begin{align}
h(i,j,k) =  T_1(i) + T_2(j-1) + T_3(k-2);
\end{align}
and from this the generalization to higher dimensions is straightforward. The formulas for ``higher triangular numbers'', i.e., those defined by
\begin{align}
T_k(n) &= \sum_{i=1}^n T_{k-1}(n)
\end{align}
for $k > 1$ can be determined inductively.

The explicit formula for 3-way interactions, with zero-based indexing, is 
\begin{align}
h(i, j, k) &= 1 + (i-1) + \frac{(j-1)j}{2} + \\
& \frac{(k-2)^3 + 3(k-2)^2 + 2(k-2)}{6}. 
\end{align}

\section{Final CSR Expansion Algorithm}
With the mapping from columns of $\bm{A}$ to a column of $\bm{A}^k$, we can now write the final form of the innermost loop of the algorithm from \ref{sec:final-algo}.
Let the mapping for $k=2$ be denoted $h^2$.
Then the innermost loop becomes:

\begin{codebox}
\footnotesize
    \zi         \For $c_2 \gets c_1 \To |\bm{c}_i|-1$ \Do
    \zi             $j_0 \gets \bm{c}_{c_0}$
    \zi             $j_1 \gets \bm{c}_{c_1}$
    \zi             $c_p \gets h^2(j_0, j_1)$
    \zi             $\bm{d}^k_{n} \gets \bm{d}_{c_0} \cdot \bm{d}_{c_1}$
    \zi             $\bm{c}^k_{n} = c_p$
    \zi             $n \gets n + 1$
                \End
\end{codebox}

The algorithm can be generalized to higher degrees by simply adding more nested loops, using higher order mappings, modifying the output dimensionality, and adjusting the counting of nonzero polynomial features in line \ref{li:row_nnz_count}.
%For interaction features, the interaction mappings can be used in lieu of the polynomial mappings with the additional change of the output dimensionality and the number of nonzero features in each row (combinations without repetition instead of with).

\section{Time Complexity}
\subsection{Analytical}

Calculating $k$-degree polynomial features via our method for a vector of dimensionality $D$ and density $d$ requires $\binom{dD}{k}$ (with repetition) products.
The complexity of the algorithm, for fixed $k \ll dD$, is therefore
\begin{align}
& O\left(\binom{dD+k-1}{k}\right) = O\left(\frac{(dD+k-1)!}{k!(dD-1)!}\right)\\
& = O\left(\frac{(dD+k-1)(dD+k-2) \dots (dD)}{k!}\right)\\
& = O\left((dD+k-1)(dD+k-2) \dots (dD)\right)\\
& = O\left(d^kD^k\right)
\end{align}

For a matrix of size $N \times D$, the complexity is therefore $O\left(Nd^kD^k\right)$.
The dense algorithm does not leverage sparsity, and its complexity is $O\left(ND^k\right)$.
Since $0 \le d \le 1$, the sparse algorithm scales polynomially with the degree of the polynomial expansion.

\subsection{Empirical}

\section{Conclusion}
We have developed an algorithm for performing polynomial feature expansions on CSR matrices that scales polynomially with respect to the density of the matrix.
The areas within machine learning that this work touches are not en vogue, but they are workhorses of industry, and every improvement in core representations has an impact across a broad range of applications. 
%This improvement could therefore spare the burning of much fossil fuel.

\bibliography{sparse_poly_icml}
\bibliographystyle{icml2017}

\end{document}

% This document was modified from the file originally made available by
% Pat Langley and Andrea Danyluk for ICML-2K. This version was
% created by Lise Getoor and Tobias Scheffer, it was slightly modified  
% from the 2010 version by Thorsten Joachims & Johannes Fuernkranz, 
% slightly modified from the 2009 version by Kiri Wagstaff and 
% Sam Roweis's 2008 version, which is slightly modified from 
% Prasad Tadepalli's 2007 version which is a lightly 
% changed version of the previous year's version by Andrew Moore, 
% which was in turn edited from those of Kristian Kersting and 
% Codrina Lauth. Alex Smola contributed to the algorithmic style files.

### 2017-02-19 paper/Final/sparse_poly_icml.tex
John's changes

@@
 \begin{abstract}
-We provide an algorithm for polynomial feature expansion that both operates on and produces a compressed sparse row matrix without any densification.
-For a vector of dimension $D$, density $d$, and degree $k$ the algorithm has time complexity $O(d^kD^k)$ where $k$ is the polynomial-feature order; this  is an improvement by a factor $d^k$ over the standard method.
+We provide an algorithm for polynomial feature expansion that both operates on and produces a compressed sparse row matrix with no densification.
+For a vector of dimension $D$ and density $0 \le d \le 1$ the algorithm has time complexity $O(d^kD^k)$ where $k$ is the polynomial-feature order; this  is an improvement by a factor $d^k$ over the standard method.
 \end{abstract}
@@
 
-Feature mappings are an intra-instance transformation, usually referred to as $\phi(\vec{x})$, that map instance vectors to higher dimensional spaces in which they are more linearly separable, allowing linear models to capture non-linearities \cite{yuan2012recent}.
-A widely used feature mapping is the polynomial expansion, which generates a new feature for each product of all $k$-order combinations in the original feature space.
-A $k$-order polynomial expansion of the feature space allows a linear model to learn polynomial relationships between dependent and independant variables.
+Feature mappings are intra-instance transformations, usually denoted by $x \mapsto \phi(\vec{x})$, that map instance vectors to higher dimensional spaces in which they are more linearly separable, allowing linear models to capture non-linearities \cite{yuan2012recent}.
+A widely used feature mapping is \emph{polynomial expansion}, which produces a new feature for degree-$k$ monomial in the original features.
+(If the original features are $x,y,z$, the order-2 polynomial features are $xy, xz, xy$, and the order-3 ones are $x^2y, x^2z, xy^2, y^2 z, xz^2, yz^2,$ and $xyz$.)
+A $k$-order polynomial expansion of the feature space allows a linear model to learn polynomial relationships between dependent and independent variables.
 This mapping was first utilized in a published experiment by Joseph Diaz Gergonne in 1815 \cite{gergonne1974application, smith1918standard}.
@@
 While other methods for capturing nonlinearities have been developed, such as kernels (the direct offspring of feature mappings), trees, generative models, and neural networks, feature mappings are still a popular tool \cite{barker200114, chang2010training, shaw2006intellectual}.
-The instance-independant nature of feature mappings allows them to pair well with linear parameter estimation techniques such as stochastic gradient descent, making them a candidate for certain types of large-scale machine learning problems when $D < N$.
+The instance-independent nature of feature mappings allows them to pair well with linear parameter estimation techniques such as stochastic gradient descent, making them a candidate for certain types of large-scale machine learning problems when $D < N$.
 
@@
 However, polynomial expansions cannot be performed directly on sparse CSR matrices, nor on any sparse matrix format, without intermediate densification steps.
-This densification not only adds extra overhead, but wastefully computes combinations of features that have a product of zero, which are then discarded during conversion into a sparse format.
+This densification not only adds overhead, but wastefully computes combinations of features that have a product of zero, which are then discarded during conversion into a sparse format.
 
-We provide an algorithm that allows CSR matrices to be the input of a polynomial feature expansion without any densification.
+We provide an algorithm that allows CSR matrices to be the input of a polynomial feature expansion with no densification.
 The algorithm leverages the CSR format to only compute products of features that result in nonzero values.
@@
 Since $0 \le d \le 1$, our algorithm is a significant improvement.
-While the algorithm we describe uses CSR matrices, it could be modified to operate on other sparse formats.
+While the algorithm we describe uses CSR matrices, it can be readily adapted to other sparse formats.
 
 \section{Preliminaries}
-Matrices are denoted by uppercase bold letters thus: $\bm{A}$. 
+We denote matrices by uppercase bold letters thus: $\bm{A}$. 
 The $i$the row of $\bm{A}$ is written $\bm{a}_i$. All vectors are written in bold, and  $\bm{a}$, with no subscript, is a vector.
@@
 
-Scalars, vectors, and matrices are often referenced with the superscript $k$.
-This is not to be interpreted as an exponent, but to indicate that it is the analogous aspect of that which procedes it, but in its polynomial expansion form.
-For example, $\bm{c}^2$ is the vector that holds columns for nonzero values in $\bm{A}$'s quadratic feature expansion CSR representation.
+Scalars, vectors, and matrices are often decorated with a superscript $k$, which is not to be interpreted as an exponent, but instead as an indicator of polynomial expansion: 
+For example, if the CSR for $\bm{A}$ is $\bm{c}, \bm{d}, \bm{p}$, then $\bm{c}^2$ is the vector that holds columns for nonzero values in $\bm{A}$'s quadratic feature expansion CSR representation.
 
-For simplicity in the presentation, we work with polynomial expansions of degree 2, but continue to use the exponent $k$ to show how the ideas apply in the general case.
+To simplify the presentation, we work with polynomial expansions of degree 2, but continue to use the exponent $k$ to show how the ideas apply in the general case.
 
@@
 In this section, we present a strawman algorithm for computing polynomial feature expansions on dense matrices.
-We then modify the algorithm  slightly to operate on a CSR matrix, in order to expose its infeasibility in that context.
+We then modify the algorithm  slightly to operate on a CSR matrix, to expose its infeasibility in that context.
 We then show how the algorithm would be feasible with an added component, which we then derive in the following section.
@@
 A natural way to calculate polynomial features for a matrix $\bm{A}$ is to walk down its rows and, for each row, take products of all $k$-combinations of elements.
-To determine in which column of $\bm{A}^k_i$ products of elements in $\bm{A}_i$ belong, a simple counter can be set to zero for each row of $\bm{A}$ and incremented efter each polynomial feature is generated.
+To determine in which column of $\bm{A}^k_i$ products of elements in $\bm{A}_i$ belong, a simple counter can be set to zero for each row of $\bm{A}$ and incremented after each polynomial feature is generated.
 This counter gives the column of $\bm{A}^k_i$ into which each expansion feature belongs.
@@
 \footnotesize
-\Procname{$\proc{Incomplete Sparse Second Order ALgorithm}(\bm{A})$}
+\Procname{$\proc{Incomplete Sparse Second Order Algorithm}(\bm{A})$}
     \li $N \gets$ row count of $\bm{A}$
@@
 
-To develop a general algorithm, we require a mapping from columns of $\bm{A}$ to a column of $\bm{A}^k$.
+To develop a general algorithm, we require a mapping from a list of $k$ columns of $\bm{A}$ to a column of $\bm{A}^k$.
 If there are $D$ columns of $\bm{A}$ and $\binom{D}{k}$ columns of $\bm{A}^k$, this can be accomplished by a bijective mapping of the following form:
@@
 
-such that $ 0 \le j_0 \le j_1 \le \dots \le j_{k-1} < D$
-where $(j_0, j_1, \dots, j_{k-1})$ are elements of $\bm{c}$ and $p_{j_0j_1 \dots i_{k-1}}$ is an element of $\bm{c}^k$.
+where (i) $ 0 \le j_0 \le j_1 \le \dots \le j_{k-1} < D$, (ii) $(j_0, j_1, \dots, j_{k-1})$ are elements of $\bm{c}$ and (iii) $p_{j_0j_1 \dots i_{k-1}}$ is an element of $\bm{c}^k$.
 %column indicies of a row vector $\vec{x}$ of an $N \times D$ input matrix, and $p_{i_0i_1 \dots i_{k-1}}$ is a column index into the polynomial expansion vector for $\vec{x}$ where the product of elements corresponding to indices $i_0, i_1, \dots, i_{k-1}$ will be stored.
@@
 \section{Construction of Mapping}
+\textbf{\color{red} This looks like the wrong mapping, since it uses $ i < j$ rather than $i \le j$, i.e., it's "interaction features" rather than polynomial features.
+Am I missing something?
+I've tried to rewrite a bit --jfh}
+
 Within this section, $i$, $j$, and $k$ denote column indices.
-For the second degree case, we seek a map from matrix indices $(i, j)$ (with $0 \le i < j < D$ ) to numbers $f(i, j)$ with $0 \le f(i, j) < \frac{D(D-1)}{2}$, one that follows the pattern indicated by
+We start with the second degree case for \emph{interaction} features, because it's easiest to understand.
+We seek a map from matrix indices $(i, j)$ (with $0 \le i < j < D$ ) to numbers $f(i, j)$ with $0 \le f(i, j) < \frac{D(D-1)}{2}$, one that follows the pattern indicated by
 \begin{align}
@@
 
+\section{Discussion}
+A natural question to ask is ``Why not just use a decent hashing algorithm?'' There are two answers.
+The first is that doing so does not avoid the \emph{computation} of features that contain a zero factor, and that alone prevents a reduction in big-O performance.
+The second is that hashing is great when you don't have a clear understanding of the distribution of items, but our index-computing function is simpler to compute than most hash functions, and provides perfect storage density.
+
 \section{Conclusion}

### 2017-02-19 dense_polynomial_features_package/dense_polynomial_features/dense_polynomial_features.pyx
code

@@ -8,7 +8,7 @@ cimport numpy as np
 
 
 ctypedef np.float64_t DATA_T
-ctypedef np.int32_t INDEX_T
+ctypedef np.int64_t INDEX_T
 
 __all__ = ['DensePolynomialFeatures']
 

### 2017-02-19 dense_polynomial_features_package/setup.py

@@ -1,0 +1,26 @@
+from Cython.Build import cythonize
+from distutils.core import setup
+from os.path import realpath
+
+setup(
+    name="dense_polynomial_features",
+    version="1.0.0",
+    packages=['dense_polynomial_features/'],
+    description="Baseline.",
+    author="Andrew Nystrom",
+    author_email="AWNystrom@gmail.com",
+    url="https://github.com/AWNystrom/SparsePolynomialFeatures",
+    keywords=["sparse", "feature", "features", "polynomial", "interaction", "combination", "combinations"],
+    license="Apache 2.0",
+    long_description=open(realpath('README.md')).read(),
+    classifiers=["Programming Language :: Python",
+                 "Programming Language :: Python :: 2.7",
+                 "Programming Language :: Python :: 2",
+                 "License :: OSI Approved :: Apache Software License",
+                 "Operating System :: OS Independent",
+                 "Development Status :: 4 - Beta",
+                 "Intended Audience :: Developers"
+                 ],
+    install_requires=['numpy'],
+    ext_modules=cythonize("./dense_polynomial_features/dense_polynomial_features.pyx")
+    )

### 2017-02-19 sparse_polynomial_features/sparse_polynomial_features.pyx

@@ -9,13 +9,13 @@ cimport numpy as np
 
 
 ctypedef np.float64_t DATA_T
-ctypedef np.int32_t INDEX_T
+ctypedef np.int64_t INDEX_T
 
 __all__ = ['SparsePolynomialFeatures']
     
 def sec_deg_poly_feats(X):
     cdef np.ndarray[DATA_T, ndim=1] data = X.data
-    cdef np.ndarray[INDEX_T, ndim=1] indices = X.indices, indptr = X.indptr
+    cdef np.ndarray[np.int32_t, ndim=1] indices = X.indices, indptr = X.indptr
   
     #Count the number of nonzero items in each row
     cdef INDEX_T poly_nz_count = 0, i = 0, D
@@ -27,8 +27,8 @@ def sec_deg_poly_feats(X):
   
     #Make the arrays that will form the new CSR matrix
     cdef np.ndarray[DATA_T, ndim=1] new_data = np.ndarray(shape=poly_nz_count, dtype=np.float64, order='C')
-    cdef np.ndarray[INDEX_T, ndim=1] new_indices = np.ndarray(shape=poly_nz_count, dtype=np.int32, order='C')
-    cdef np.ndarray[INDEX_T, ndim=1] new_indptr = np.ndarray(shape=indptr.shape[0], dtype=np.int32, order='C')
+    cdef np.ndarray[INDEX_T, ndim=1] new_indices = np.ndarray(shape=poly_nz_count, dtype=np.int64, order='C')
+    cdef np.ndarray[INDEX_T, ndim=1] new_indptr = np.ndarray(shape=indptr.shape[0], dtype=np.int64, order='C')
   
     new_indptr[0] = 0
     cdef INDEX_T ind = 0, start, stop, num_cols, k1, k2, col2, original_D = X.shape[1]
@@ -77,7 +77,7 @@ def sec_deg_poly_feats(X):
 
 def sec_deg_inter_feats(X):
     cdef np.ndarray[DATA_T, ndim=1] data = X.data
-    cdef np.ndarray[INDEX_T, ndim=1] indices = X.indices, indptr = X.indptr
+    cdef np.ndarray[np.int32_t, ndim=1] indices = X.indices, indptr = X.indptr
   
     #Count the number of nonzero items in each row
     cdef INDEX_T poly_nz_count = 0, i = 0, D
@@ -89,8 +89,8 @@ def sec_deg_inter_feats(X):
   
[…]

### 2017-02-19 sparse_polynomial_features/test.py

@@ -1,0 +1,75 @@
+import pyximport; pyximport.install()
+from sparse_polynomial_features import SparsePolynomialFeatures
+from sklearn.preprocessing import PolynomialFeatures
+import unittest
+from scipy.sparse import random
+from code import interact
+import cPickle
+from numpy import array
+from numpy.random import choice
+from scipy.sparse import csr_matrix
+
+primes = cPickle.load(open('primes.pickle'))
+
+"""
+def test_this(rows, cols, density, interaction_only, degree):
+  print(rows, cols, density, interaction_only, degree)
+  X_sparse = random(rows, cols, density).tocsr()
+  X_dense = X_sparse.toarray()
+  
+  poly_d = PolynomialFeatures(degree=degree, interaction_only=interaction_only, include_bias=False).fit_transform(X_dense)
+  
+  try:
+    poly_s = SparsePolynomialFeatures(degree=degree, interaction_only=interaction_only).fit_transform(X_sparse).toarray()
+  except:
+    interact(local=locals())
+
+  f1 = get_val_dist(poly_d)
+  f2 = get_val_dist(poly_s)
+  
+  all_vals = set()
+  all_vals.update(f1.keys())
+  all_vals.update(f2.keys())
+  
+  for v in all_vals:
+    assert f1[v] == f2[v]
+
+def get_val_dist(X):
+  f = {}
+  for v in X.flatten():
+    f[v] = f.get(v, 0) + 1
+  return f
+
+for interaction_only in [True]:#[True, False]:
+  for degree in [2, 3]:
+    for density in [0., 0.5, 1.]:
+      for rows in [1, 10]:
+        for cols in [1, 10]:
+          test_this(rows, cols, density, interaction_only, degree)
+"""
+
+def via_primes(N, D, degree, interaction_only):
+
+  X_dense = array(choice(primes, N*D, replace=False), dtype=float).reshape((N, D))
+  X_sparse = csr_matrix(X_dense)
+  poly_d = PolynomialFeatures(degree=degree, interaction_only=interaction_only,
+                              include_bias=False).fit_transform(X_dense)
+  poly_s = SparsePolynomialFeatures(degree=degree, 
+                                    interaction_only=interaction_only).fit_transform(X_sparse).toarray()
+
+  #Figure out the mapping on the first one, then ensure that it holds for all the others.
[…]

### 2017-02-21 dense_polynomial_features_package/dense_polynomial_features/dense_polynomial_features.pyx
benchmarck version without return values

@@ -33,59 +33,61 @@ class DensePolynomialFeatures(BaseEstimator, TransformerMixin):
     cdef INDEX_T degree = <INDEX_T>(self.degree)
     cdef INDEX_T N = X.shape[0]
     cdef INDEX_T D = X.shape[1]
-    cdef INDEX_T i, j, row, poly_index
+    cdef INDEX_T i, j, row, poly_index, row_offset
     cdef poly_D = D + <INDEX_T>((D**2+D)/2)
-    cdef np.ndarray[DATA_T, ndim=2] poly_X = np.ndarray(shape=(N, poly_D), dtype=np.float64, order='C')
+    cdef np.ndarray[DATA_T, ndim=1] flat_poly_X = np.ndarray(shape=(N*poly_D, ), dtype=np.float64, order='C')
     cdef DATA_T data1
     
     row = 0
     while row < N:
       poly_index = 0
       i = 0
+      row_offset = row * poly_D
       while i < D:
         data1 = X[row, i]
-        poly_X[row, poly_index] = data1
+        flat_poly_X[row_offset + poly_index] = data1
         poly_index += 1
         j = 0
         while j <= i:
-          poly_X[row, poly_index] = data1 * X[row, j]
+          flat_poly_X[row_offset + poly_index] = data1 * X[row, j]
           poly_index += 1
           j += 1
         i += 1
       row += 1
       
-    return poly_X
+    #return flat_poly_X.reshape((N, poly_D))
   
   def transform3(self, X_):
     cdef np.ndarray[DATA_T, ndim=2] X = X_
     cdef INDEX_T degree = <INDEX_T>(self.degree)
     cdef INDEX_T N = X.shape[0]
     cdef INDEX_T D = X.shape[1]
-    cdef INDEX_T i, j, k, row, poly_index
+    cdef INDEX_T i, j, k, row, poly_index, row_offset
     cdef poly_D = D + <INDEX_T>((D**2+D)/2) + <INDEX_T>((D**3+3*D**2+2*D)/6)
-    cdef np.ndarray[DATA_T, ndim=2] poly_X = np.ndarray(shape=(N, poly_D), dtype=np.float64, order='C')
+    cdef np.ndarray[DATA_T, ndim=1] flat_poly_X = np.ndarray(shape=(N*poly_D, ), dtype=np.float64, order='C')
     cdef DATA_T data1, data2
     
     row = 0
     while row < N:
       poly_index = 0
       i = 0
[…]

### 2017-02-21 paper/Final/code/simple_plots.py

@@ -1,0 +1,209 @@
+from scipy.sparse import random, vstack
+from sparse_polynomial_features import SparsePolynomialFeatures
+from dense_polynomial_features import DensePolynomialFeatures as PolynomialFeatures
+from time import time
+import numpy as np
+import matplotlib.pyplot as plt
+from code import interact
+import cPickle
+from sys import argv
+import seaborn as sns
+
+np.random.seed(42)
+
+"""
+We have variables D, d, N, t
+
+Vary D, d, N while keeping the others constant and plot it vs time.
+
+Constant slices: D=500, N=1000, d=0.5
+
+Do this with k = 2, then all over with k=3
+
+This will yield 6 plots
+
+"""
+
+
+"""
+b: blue
+g: green
+r: red
+c: cyan
+m: magenta
+y: yellow
+k: black
+w: white
+"""
+
+poly_order = 2
+
+filename = 'order_%s_simple_data.pickle' % (poly_order,)
+colors = ['b', 'r', 'm', 'y', 'm', 'c', 'g', 'k']
+iters = 20
+
+density_steps = 10.
+ds = np.arange(density_steps + 1) / density_steps
+
+#if poly_order == 2:
+#  Ds = [1, 100, 200, 300, 400, 500, 600, 700, 800, 900, 1000]
+#  Ns = [1, 100, 200, 300, 400, 500, 600, 700, 800, 900, 1000]
+
+#  D_slice = int(np.median(Ds))
+#  d_slice = 0.2
+#  N_slice = int(np.median(Ns))
+  
+#elif poly_order == 3:
+
+if poly_order == 3:
+  Ns = [1, 100, 200, 300, 400, 500, 600, 700, 800, 900, 1000]
+  max_D = 250.
+  Ds = map(int, np.arange(0, max_D+max_D/10, max_D/10))
+  Ds[0] = 1
+  N_slice = int(np.median(Ds))
+elif poly_order == 2:
+  Ns = [1, 100, 200, 300, 400, 500, 600, 700, 800, 900, 1000]
+  max_D = 1000.
+  Ds = map(int, np.arange(0, max_D+max_D/10, max_D/10))
+  Ds[0] = 1
+  N_slice = int(max(Ds))
+
+D_slice = int(np.median(Ds))
+d_slice = 0.2
+
+#else:
+#  assert(False)
+
+# Key to dep_to_times is always (N, D, d)
+var_sets = [
+             {
[…]

### 2017-02-21 sparse_polynomial_features/sparse_polynomial_features.pyx

@@ -73,7 +73,7 @@ def sec_deg_poly_feats(X):
     A.indices = new_indices
     A.indptr = new_indptr
     A._shape = (X.shape[0], X.shape[1] + <INDEX_T>((X.shape[1]**2 + X.shape[1]) / 2))
-    return A
+    #return A
 
 def sec_deg_inter_feats(X):
     cdef np.ndarray[DATA_T, ndim=1] data = X.data
@@ -135,7 +135,7 @@ def sec_deg_inter_feats(X):
     A.indices = new_indices
     A.indptr = new_indptr
     A._shape = (X.shape[0], X.shape[1] + <INDEX_T>((X.shape[1]**2 - X.shape[1]) / 2))
-    return A
+    #return A
 
 
 def third_deg_poly_feats(X):
@@ -223,7 +223,7 @@ def third_deg_poly_feats(X):
     A.indptr = new_indptr
     D = X.shape[1]
     A._shape = (X.shape[0], D + <INDEX_T>((D**2 + D) / 2) + <INDEX_T>((D**3 + 3*D**2 + 2*D)/6))
-    return A
+    #return A
 
 def third_deg_inter_feats(X):
     cdef np.ndarray[DATA_T, ndim=1] data = X.data
@@ -310,7 +310,7 @@ def third_deg_inter_feats(X):
     A.indptr = new_indptr
     D = X.shape[1]
     A._shape = (X.shape[0], D + <INDEX_T>((D**2 - D) / 2) + <INDEX_T>((D**3 - 3*D**2 + 2*D)/6))
-    return A
+    #return A
 
 params_to_funct = {(2, False): sec_deg_poly_feats, 
                    (2, True): sec_deg_inter_feats,

### 2017-02-24 paper/Final/sparse_poly_icml.tex
almost done

@@
 \usepackage{bm}
-\usepackage{clrscode3e}
 \usepackage{float}
@@
 We provide an algorithm for polynomial feature expansion that both operates on and produces a compressed sparse row matrix with no densification.
-For a vector of dimension $D$ and density $0 \le d \le 1$ the algorithm has time complexity $O(d^kD^k)$ where $k$ is the polynomial-feature order; this  is an improvement by a factor $d^k$ over the standard method.
+For a vector of dimension $D$ and density $0 \le d \le 1$ the algorithm has time complexity $\Theta(d^kD^k)$ where $k$ is the polynomial-feature order; this  is an improvement by a factor $d^k$ over the standard method.
 \end{abstract}
@@
 While other methods for capturing nonlinearities have been developed, such as kernels (the direct offspring of feature mappings), trees, generative models, and neural networks, feature mappings are still a popular tool \cite{barker200114, chang2010training, shaw2006intellectual}.
-The instance-independent nature of feature mappings allows them to pair well with linear parameter estimation techniques such as stochastic gradient descent, making them a candidate for certain types of large-scale machine learning problems when $D < N$.
+The instance-independent nature of feature mappings allows them to pair well with linear parameter estimation techniques such as stochastic gradient descent, making them a candidate for certain types of large-scale machine learning problems when $D \ll N$.
 
-The compressed sparse row (CSR) matrix format \cite(saad1994sparskit) is widely used \cite{liu2012sparse, bulucc2009parallel, bell2008efficient, white1997improving} and supported \cite{eigenweb, bastien2012theano, koenker2003sparsem}, and is considered the standard data structure for sparse, computationally heavy tasks.
+The compressed sparse row (CSR) matrix format \cite{saad1994sparskit} is widely used \cite{liu2012sparse, bulucc2009parallel, bell2008efficient, white1997improving} and supported \cite{eigenweb, bastien2012theano, scikit-learn, koenker2003sparsem}, and is considered the standard data structure for sparse, computationally heavy tasks.
 However, polynomial expansions cannot be performed directly on sparse CSR matrices, nor on any sparse matrix format, without intermediate densification steps.
@@
 The algorithm leverages the CSR format to only compute products of features that result in nonzero values.
-This exploits the sparsity of the data to achieve an improved time complexity of $O(d^kD^k)$ on each vector of the matrix where $k$ is the degree of the expansion, $D$ is the dimensionality, and $d$ is the density.
-The standard algorithm has time complexity $O(D^k)$.
+This exploits the sparsity of the data to achieve an improved time complexity of $\Theta(d^kD^k)$ on each vector of the matrix where $k$ is the degree of the expansion, $D$ is the dimensionality, and $d$ is the density.
+The standard algorithm has time complexity $\Theta(D^k)$.
 Since $0 \le d \le 1$, our algorithm is a significant improvement.
@@
 We denote matrices by uppercase bold letters thus: $\bm{A}$. 
-The $i$the row of $\bm{A}$ is written $\bm{a}_i$. All vectors are written in bold, and  $\bm{a}$, with no subscript, is a vector.
+The $i^{th}$ the row of $\bm{A}$ is written $\bm{a}_i$. All vectors are written in bold, and  $\bm{a}$, with no subscript, is a vector. Non-bold letters are scalars.
 
@@
 
+It should be noted that in practice, as well as in our code and experiments, expansions for degrees $1, 2, \dots, k-1$ are also generated.
+The final design matrix is the augmentation of all such expansions.
+However, the algorithm descriptions in this paper omit these steps as they would become unnecessarily visually cumbersome.
+Extending them to include all degrees less than $k$ is trivial and does not affect the complexity of the algorithm as the terms that involve $k$ dominate.
+
 \subsection{Dense Expansion Algorithm}
@@
 
-\begin{codebox}
-\footnotesize
-\Procname{$\proc{Dense Second Order Algorithm}(\bm{A})$}
-    \li $N \gets$ row count of $\bm{A}$
-    \li $D \gets$ column count of $\bm{A}$
-    \li $\bm{A}^k$ $\gets$ empty $N \times \binom{D}{2}$ matrix
-    \li \For $i \gets 0 \To N-1$ \Do
-    \li     $c_p \gets 0$
-    \li     \For $j_1 \gets 0 \To D-1$ \Do
-    \li         \For $j_2 \gets j_1 \To D-1$ \Do
-    \li             $\bm{A}^k_{i{c_p}} \gets \bm{A}_{ij_1} \cdot \bm{A}_{ij_2}$
-    \li             $c_p \gets c_p + 1$
-                \End
-            \End
-       	\End
-\end{codebox}
+\begin{algorithm}[H]
+   \caption{Dense Second Order Polynomial Expansion}
+   \label{alg:Dense-Second-Order-Polynomial-Expansion}
+\begin{algorithmic}[1]
+   \STATE {\bfseries Input:} data $\bm{A}$, size $N \times D$
+   \STATE $\bm{A}^k$ $\gets$ empty $N \times \binom{D}{2}$ matrix
+   \FOR{$i \gets 0$ {\bfseries to} $N-1$}
+      \STATE $c_p \gets 0$
+      \FOR{$j_1 \gets 0$ {\bfseries to} $D-1$}
+          \FOR{$j_2 \gets j_1$ {\bfseries to} $D-1$}
+              \STATE $\bm{A}^k_{i{c_p}} \gets \bm{A}_{ij_1} \cdot \bm{A}_{ij_2}$
+              \STATE $c_p \gets c_p + 1$
+          \ENDFOR
+      \ENDFOR
+   \ENDFOR
+\end{algorithmic}
+\end{algorithm}
 
@@
 
-\begin{codebox}
-\footnotesize
-\Procname{$\proc{Incomplete Sparse Second Order Algorithm}(\bm{A})$}
-    \li $N \gets$ row count of $\bm{A}$
-    \li $\bm{p}^k$ $\gets$ vector of size $N+1$
-    \li $\bm{p}^k_0 \gets 0$
-    \li $nnz^k \gets 0$
-    \li \For $i \gets 0 \To N-1$ \Do
-    \li     $i_{start} \gets \bm{p}_i$
-    \li     $i_{stop} \gets \bm{p}_{i+1}$
-    \li     $\bm{c}_i \gets \bm{c}_{i_{start}:i_{stop}}$
-    \li     $nnz^k_i \gets \binom{|\bm{c}_i|}{2}$ \label{li:row_nnz_count}
-    \li     $nnz^k \gets nnz^k + nnz^k_i$
-    \li     $\bm{p}^k_{i+1} \gets \bm{p}^k_i + nnz^k_i$
-        \End
-    \zi     
-    \zi \Comment Build up the elements of $\bm{p}^k$, $\bm{c}^k$, and $\bm{d}^k$
-    \li $\bm{p}^k$ $\gets$ vector of size $N+1$
-    \li $\bm{c}^k$ $\gets$ vector of size $nnz^k$
-    \li $\bm{d}^k$ $\gets$ vector of size $nnz^k$
-    \li $n \gets 0$
-    \li \For $i \gets 0 \To N-1$ \Do
-    \li     $i_{start} \gets \bm{p}_i$
-    \li     $i_{stop} \gets \bm{p}_{i+1}$
-    \li     $\bm{c}_i \gets \bm{c}_{i_{start}:i_{stop}}$
-    \li     $\bm{d}_i \gets \bm{d}_{i_{start}:i_{stop}}$
-    \li     \For $c_1 \gets 0 \To |\bm{c}_i|-1$ \Do
-    \li         \For $c_2 \gets c_1 \To |\bm{c}_i|-1$ \Do
-%    \li             $j_0 \gets \bm{c}_{c_0}$
-%    \li             $j_1 \gets \bm{c}_{c_1}$
-    \li             $\bm{d}^k_{n} \gets \bm{d}_{c_0} \cdot \bm{d}_{c_1}$
-    \li             $\bm{c}^k_{n} = ?$ \label{li:set_ck}
-    \li             $n \gets n + 1$
-                \End
-            \End
-       	\End
-\end{codebox}
+\begin{algorithm}[H]
+   \caption{Incomplete Sparse Second Order Polynomial Expansion}
+   \label{alg:Incomplete-Sparse-Second-Order-Polynomial-Expansion}
+\begin{algorithmic}[1]
+   \STATE {\bfseries Input:} data $\bm{A}$, size $N \times D$
+   \STATE $\bm{p}^k$ $\gets$ vector of size $N+1$
+   \STATE $\bm{p}^k_0 \gets 0$
+   \STATE $nnz^k \gets 0$
+   \FOR{$i \gets 0$ {\bfseries to} $N-1$}
+      \STATE $i_{start} \gets \bm{p}_i$
+      \STATE $i_{stop} \gets \bm{p}_{i+1}$
+      \STATE $\bm{c}_i \gets \bm{c}_{i_{start}:i_{stop}}$
+      \STATE $nnz^k_i \gets \binom{|\bm{c}_i|}{2}$ \label{li:row_nnz_count}
+      \STATE $nnz^k \gets nnz^k + nnz^k_i$
+      \STATE $\bm{p}^k_{i+1} \gets \bm{p}^k_i + nnz^k_i$
+  \ENDFOR
 
-The crux of the problem is at line \ref{li:set_ck}.
+  \STATE $\bm{p}^k$ $\gets$ vector of size $N+1$
+  \STATE $\bm{c}^k$ $\gets$ vector of size $nnz^k$
+  \STATE $\bm{d}^k$ $\gets$ vector of size $nnz^k$
+  \STATE $n \gets 0$
+
+  \FOR {$i \gets 0$ {\bfseries to} $N-1$}
+      \STATE $i_{start} \gets \bm{p}_i$
+      \STATE $i_{stop} \gets \bm{p}_{i+1}$
+      \STATE $\bm{c}_i \gets \bm{c}_{i_{start}:i_{stop}}$
+      \STATE $\bm{d}_i \gets \bm{d}_{i_{start}:i_{stop}}$
+      \FOR {$c_1 \gets 0$ {\bfseries to} $|\bm{c}_i|-1$}
+          \FOR {$c_2 \gets c_1$ {\bfseries to} $|\bm{c}_i|-1$}
+              \STATE $\bm{d}^k_{n} \gets \bm{d}_{c_0} \cdot \bm{d}_{c_1}$
+              \STATE $\bm{c}^k_{n} = ?$ \label{li:set_ck}
+              \STATE $n \gets n + 1$
+          \ENDFOR
+      \ENDFOR
+  \ENDFOR
+\end{algorithmic}
+\end{algorithm}
+
+The crux of the problem is at line 25. %reference wasn't working \ref{li:set_ck}.
 Given the arbitrary columns involved in a polynomial feature of $\bm{A}_i$, we need to determine the corresponding column of $\bm{A}^k_i$.
@@
 \section{Construction of Mapping}
-\textbf{\color{red} This looks like the wrong mapping, since it uses $ i < j$ rather than $i \le j$, i.e., it's "interaction features" rather than polynomial features.
+%\textbf{\color{red} This looks like the wrong mapping, since it uses $ i < j$ rather than $i \le j$, i.e., it's "interaction features" rather than polynomial features.
 Am I missing something?
@@
 Within this section, $i$, $j$, and $k$ denote column indices.
-We start with the second degree case for \emph{interaction} features, because it's easiest to understand.
-We seek a map from matrix indices $(i, j)$ (with $0 \le i < j < D$ ) to numbers $f(i, j)$ with $0 \le f(i, j) < \frac{D(D-1)}{2}$, one that follows the pattern indicated by
+We start with the second degree case for \emph{interaction} features due to its relative simplicity as it is easiest to understand.
+We seek a map from matrix indices $(i, j)$ (with $0 \le i < j < D$ ) to numbers $f(i, j)$ with $0 \le f(i, j) < \frac{D(D-1)}{2}$ that follows the pattern indicated by
 \begin{align}
@@
 To handle \emph{three-way interactions}, we need to map triples of indices in a 3-index array to a flat list, and similarly for higher-order interactions.
-For this, we'll need the tetrahedral numbers $T_3(n) = \sum_{i=1}^n T_{2}(n) =
+For this, we require the tetrahedral numbers $T_3(n) = \sum_{i=1}^n T_{2}(n) =
 \frac{1}{6}(n^3 + 3n^2 + 2n)$.
@@
 
-\begin{codebox}
-\footnotesize
-    \zi         \For $c_2 \gets c_1 \To |\bm{c}_i|-1$ \Do
-    \zi             $j_0 \gets \bm{c}_{c_0}$
-    \zi             $j_1 \gets \bm{c}_{c_1}$
-    \zi             $c_p \gets h^2(j_0, j_1)$
-    \zi             $\bm{d}^k_{n} \gets \bm{d}_{c_0} \cdot \bm{d}_{c_1}$
-    \zi             $\bm{c}^k_{n} = c_p$
-    \zi             $n \gets n + 1$
-                \End
-\end{codebox}
+\begin{algorithm}[H]
+   \caption{Inner Loop of Completed Sparse Second Order Polynomial Expansion}
+   \label{alg:Inner-Loop-of-Completed-Sparse-Second-Order-Polynomial-Expansion}
+\begin{algorithmic}[1]
+  \FOR {$c_2 \gets c_1$ {\bfseries to} $|\bm{c}_i|-1$}
+      \STATE $j_0 \gets \bm{c}_{c_0}$
+      \STATE $j_1 \gets \bm{c}_{c_1}$
+      \STATE $c_p \gets h^2(j_0, j_1)$
+      \STATE $\bm{d}^k_{n} \gets \bm{d}_{c_0} \cdot \bm{d}_{c_1}$
+      \STATE $\bm{c}^k_{n} = c_p$
+      \STATE $n \gets n + 1$
+  \ENDFOR
+\end{algorithmic}
+\end{algorithm}
 
@@
 \subsection{Analytical}
+\label{sec:analytical}
 
@@
 \begin{align}
-& O\left(\binom{dD+k-1}{k}\right) = O\left(\frac{(dD+k-1)!}{k!(dD-1)!}\right)\\
-& = O\left(\frac{(dD+k-1)(dD+k-2) \dots (dD)}{k!}\right)\\
-& = O\left((dD+k-1)(dD+k-2) \dots (dD)\right)\\
-& = O\left(d^kD^k\right)
+& \Theta\left(\binom{dD+k-1}{k}\right) = \Theta\left(\frac{(dD+k-1)!}{k!(dD-1)!}\right)\\
+& = \Theta\left(\frac{(dD+k-1)(dD+k-2) \dots (dD)}{k!}\right)\\
+& = \Theta\left((dD+k-1)(dD+k-2) \dots (dD)\right)\\
+& = \Theta\left(d^kD^k\right)
 \end{align}
 
-For a matrix of size $N \times D$, the complexity is therefore $O\left(Nd^kD^k\right)$.
-The dense algorithm does not leverage sparsity, and its complexity is $O\left(ND^k\right)$.
+For a matrix of size $N \times D$, the complexity is therefore $\Theta\left(Nd^kD^k\right)$.
+The dense algorithm does not leverage sparsity, and its complexity is $\Theta\left(ND^k\right)$.
 Since $0 \le d \le 1$, the sparse algorithm scales polynomially with the degree of the polynomial expansion.
 
+\newpage
+
 \subsection{Empirical}
+To verify the average time complexity of our algorithm, we implemented both the sparse version and the baseline in the Cython programming language so that results would be directly comparable.
+We sought the relationships between runtime and the instance count ($N$), the instance dimensionality ($D$), and the instance density ($d$).
+
+To find these relationships, we individually varied $N$, $D$, and $d$ while holding the remaining two constant.
+For each of these configurations, we generated $20$ matrices and took the average time to reduce variance.
+The time to densify did not count against the dense algorithm.
+Figure-\ref{fig:all-vs-time} summarizes our findings.
+
+Varying the density ($d$) (column 1) shows that our algorithm scales polynomially with $d$, but that the baseline is unaffected by it.
+Both algorithms increase polynomially with the dimensionality ($D$), but ours at a significantly reduced rate.
+Likewise, both algorithms scale linearly with the instance count ($N$), but ours to a much lesser degree.
+
+Note that the point at which the sparse and dense algorithms intersect when varying $d$ is to the right of $0.5$, which is when a matrix technically becomes sparse.
+The point at which this happens will depend on $D$, but this shows that our algorithm is useful even for some dense matrices.
+
+\begin{figure*}[ht!]
+\vskip 0.2in
+\begin{center}
+\centerline{\includegraphics[width=\textwidth]{all_vs_time.png}}
+\caption{
+%This figure shows how the runtimes of the sparse and dense %algorithms scale with $d$, $D$, and $N$.
+%Each point is the average of 20 samples.
+%The time to densify did not count against the dense %algorithm.
+%The figure shows that our algorithm scales polynomially with %$d$ and $D$ and linearly with $N$, which is in accordance %with the analysis of section \ref{sec:analytical}.
+Summary performance comparison plots for quadratic (top) and cubic (bottom) cases showing how the algorithm's performance varies with $d$, $D$, and $N$; our sparse algorithm is shown in blue, the dense algorithm in red.
+Each point in each graph is an average of 20 runs, and the time used in densification is not included in the dense-algorithm timings.
+In the quadratic case, sparsity loses its advantage at about 67\%, and at about 77\% for the cubic case, though these precise intersections depend on $D$.
+In general, taking advantage of sparsity shows large benefits, so large that it's difficult to see that the performance does not actually change linearly with $D$ (column 2); figure \ref{fig:sparse_D_and_N_vs_time} gives further details.}
+\label{fig:all-vs-time}
+\end{center}
+\vskip -0.2in
+\end{figure*}
+
+\begin{figure*}[ht!]
+\vskip 0.2in
+\begin{center}
+\centerline{\includegraphics[width=\textwidth]{sparse_D_and_N_vs_time.png}}
+\caption{A closer view of only the sparse runtimes while varying $D$ (left) and $N$ (right) for $d = 0.2$.
+The left subplot shows that varying $D$ gives polynomial growth in runtime; quadratic for $k = 2$ (dashed line) and cubic for $k = 3$ (dotted line).
+These nonlinearities were not apparent in Figure \ref{fig:all-vs-time} due to the much greater runtimes of the dense algorithm.
+The right subplot shows linear growth in runtime for both.
+These findings are in accordance with the analysis of section \ref{sec:analytical}.}
+\label{fig:sparse_D_and_N_vs_time}
+\end{center}
+\vskip -0.2in
+\end{figure*}
 
@@
 A natural question to ask is ``Why not just use a decent hashing algorithm?'' There are two answers.
-The first is that doing so does not avoid the \emph{computation} of features that contain a zero factor, and that alone prevents a reduction in big-O performance.
+The first is that doing so does not avoid the \emph{computation} of features that contain a zero factor, and that alone prevents a reduction in big-theta performance.
 The second is that hashing is great when you don't have a clear understanding of the distribution of items, but our index-computing function is simpler to compute than most hash functions, and provides perfect storage density.
@@
 %This improvement could therefore spare the burning of much fossil fuel.
+
+\newpage
 

### 2018-03-07 paper/Final/sparse_poly_icml.tex
Major updates to paper, including reordering of indices in mapping to match common libraries.

@@
 \usepackage{amssymb}
+\usepackage{caption}
 
@@
 % the note in the first column to ``Under review.  Do not distribute.''
-\usepackage{icml2017}
+%\usepackage{icml2017}
 
@@
 % note in the first column to ``Proceedings of the...''
-%\usepackage[accepted]{icml2017}
+\usepackage[accepted]{icml2017}
 
@@
 % Therefore, a short form for the running title is supplied here:
-\icmltitlerunning{Leveraging Matrix Sparsity in Polynomial Expansions}
+\icmltitlerunning{Speeding Up Polynomial Feature Expansions of Compressed Sparse Row Matrices}
 
@@
 \twocolumn[
-\icmltitle{Leveraging Matrix Sparsity in Polynomial Expansions}
+\icmltitle{Speeding Up Polynomial Feature Expansions by Leveraging Density}
 
@@
 \begin{icmlauthorlist}
-\icmlauthor{Aeiau Zzzz}{equal,to}
-\icmlauthor{Bauiu C.~Yyyy}{equal,to,goo}
-\icmlauthor{Cieua Vvvvv}{goo}
-\icmlauthor{Iaesut Saoeu}{ed}
-\icmlauthor{Fiuea Rrrr}{to}
-\icmlauthor{Tateu H.~Yasehe}{ed,to,goo} 
-\icmlauthor{Aaoeu Iasoh}{goo}
-\icmlauthor{Buiui Eueu}{ed}
-\icmlauthor{Aeuia Zzzz}{ed}
-\icmlauthor{Bieea C.~Yyyy}{to,goo}
-\icmlauthor{Teoau Xxxx}{ed}
-\icmlauthor{Eee Pppp}{ed}
+\icmlauthor{Andrew Nystrom}{google}
+\icmlauthor{John Hughes}{brown}
 \end{icmlauthorlist}
 
-\icmlaffiliation{to}{University of Torontoland, Torontoland, Canada}
-\icmlaffiliation{goo}{Googol ShallowMind, New London, Michigan, USA}
-\icmlaffiliation{ed}{University of Edenborrow, Edenborrow, United Kingdom}
+\icmlaffiliation{google}{Google Inc.}
+\icmlaffiliation{brown}{Brown University}
 
-\icmlcorrespondingauthor{Cieua Vvvvv}{c.vvvvv@googol.com}
-\icmlcorrespondingauthor{Eee Pppp}{ep@eden.co.uk}
+\icmlcorrespondingauthor{Andrew Nystrom}{nystrom@google.com}
+\icmlcorrespondingauthor{John Hughes}{jfh@cs.brown.edu}
 
@@
 \begin{abstract}
-We provide an algorithm for polynomial feature expansion that both operates on and produces a compressed sparse row matrix with no densification.
-For a vector of dimension $D$ and density $0 \le d \le 1$ the algorithm has time complexity $\Theta(d^kD^k)$ where $k$ is the polynomial-feature order; this  is an improvement by a factor $d^k$ over the standard method.
+We provide an algorithm that speeds up polynomial and interaction feature expansion on sparse matrices.
+The algorithm operates on and produces a compressed sparse row matrix with no densification.
+It works by mapping (via simplex numbers) column indices with nonzero elements to a column index in the polynomial expansion vector, allowing for only the nonzero elements in a CSR matrix to be iterated over when computing products.
+For a vector of dimension $D$ and density $0 \le d \le 1$, the algorithm has time complexity $\Theta(d^KD^K)$ where $K$ is the polynomial-feature order; this  is an improvement by a factor $d^K$ over the standard method.
 \end{abstract}
@@
 
-Feature mappings are intra-instance transformations, usually denoted by $x \mapsto \phi(\vec{x})$, that map instance vectors to higher dimensional spaces in which they are more linearly separable, allowing linear models to capture non-linearities \cite{yuan2012recent}.
-A widely used feature mapping is \emph{polynomial expansion}, which produces a new feature for degree-$k$ monomial in the original features.
-(If the original features are $x,y,z$, the order-2 polynomial features are $xy, xz, xy$, and the order-3 ones are $x^2y, x^2z, xy^2, y^2 z, xz^2, yz^2,$ and $xyz$.)
-A $k$-order polynomial expansion of the feature space allows a linear model to learn polynomial relationships between dependent and independent variables.
+In machine learning and statistical modeling, feature mappings are intra-instance transformations, usually denoted by $x \mapsto \phi(\vec{x})$, that map instance vectors to higher dimensional spaces in which they are more linearly separable, allowing linear models to capture non-linearities \cite{yuan2012recent}.
+
+A well known and widely used feature mapping is the \emph{polynomial expansion}, which produces a new feature for each degree-$k$ monomial in the original features.
+(If the original features are $x,y,z$, the order-2 polynomial features are $x^2, y^2, z^2, xy, xz, xy$, and the order-3 ones are $x^3, y^3, z^3, x^2y, x^2z, xy^2, y^2 z, xz^2, yz^2,$ and $xyz$.)
+A $k$-order polynomial feature expansion of the feature space allows a linear model to learn polynomial relationships between dependent and independent variables.
 This mapping was first utilized in a published experiment by Joseph Diaz Gergonne in 1815 \cite{gergonne1974application, smith1918standard}.
@@
 The compressed sparse row (CSR) matrix format \cite{saad1994sparskit} is widely used \cite{liu2012sparse, bulucc2009parallel, bell2008efficient, white1997improving} and supported \cite{eigenweb, bastien2012theano, scikit-learn, koenker2003sparsem}, and is considered the standard data structure for sparse, computationally heavy tasks.
-However, polynomial expansions cannot be performed directly on sparse CSR matrices, nor on any sparse matrix format, without intermediate densification steps.
+However, polynomial feature expansions cannot be performed directly on CSR matrices, nor on any sparse matrix format, without intermediate densification steps.
 This densification not only adds overhead, but wastefully computes combinations of features that have a product of zero, which are then discarded during conversion into a sparse format.
 
-We provide an algorithm that allows CSR matrices to be the input of a polynomial feature expansion with no densification.
+We describe an algorithm that takes a CSR matrix as input and produces a CSR matrix for its degree-$K$ polynomial feature expansion with no intermediate densification.
 The algorithm leverages the CSR format to only compute products of features that result in nonzero values.
-This exploits the sparsity of the data to achieve an improved time complexity of $\Theta(d^kD^k)$ on each vector of the matrix where $k$ is the degree of the expansion, $D$ is the dimensionality, and $d$ is the density.
-The standard algorithm has time complexity $\Theta(D^k)$.
-Since $0 \le d \le 1$, our algorithm is a significant improvement.
+This exploits the sparsity of the data to achieve an improved time complexity of $\Theta(d^KD^K)$ on each vector of the matrix where $K$ is the degree of the expansion, $D$ is the dimensionality, and $d$ is the density.
+The standard algorithm has time complexity $\Theta(D^K)$.
+Since $0 \le d \le 1$, our algorithm is a significant improvement for small $d$.
 While the algorithm we describe uses CSR matrices, it can be readily adapted to other sparse formats.
 
-\section{Preliminaries}
+\section{Notation}
 We denote matrices by uppercase bold letters thus: $\bm{A}$. 
-The $i^{th}$ the row of $\bm{A}$ is written $\bm{a}_i$. All vectors are written in bold, and  $\bm{a}$, with no subscript, is a vector. Non-bold letters are scalars.
+The $i^{th}$ the row of $\bm{A}$ is written $\bm{a}_i$.
+All vectors are written in bold, and  $\bm{a}$, with no subscript, is a vector.
+Non-bold letters are scalars.
+We sometimes use `slice' notation on subscripts, so that $x_{2:5}$ indicates the second through fifth elements of the vector $x$.
 
-A compressed sparse row (CSR) matrix representation of an $r$-row matrix $\bm{A}$ consists of three vectors: $\bm{c}$, $\bm{d}$, and $\bm{p}$ and a single number: the number of columns of $\bm{A}$.
+A compressed sparse row (CSR) matrix representation of an $r$-row matrix $\bm{A}$ consists of three vectors: $\bm{c}$, $\bm{d}$, and $\bm{p}$ and a single number: the number $N$ of columns of $\bm{A}$.
 The vectors
@@
 the data describing nonzero columns of $\bm{a}_i$ are within the other two vectors: $\bm{c}_{\bm{p}_i:\bm{p}_{i+1}}$ contain the column indices of those entries; $\bm{d}_{\bm{p}_i:\bm{p}_{i+1}}$ contain the entries themselves.
-Since only nonzero elements of each row are held, the overall number of columns of $\bm{A}$  must also be stored, since it cannot be derived
+Since only nonzero elements of each row are held, the overall number $N$ of columns of $\bm{A}$  must also be stored, since it cannot be derived
 from the other data.
+
+Uppercase $K$ refers to the degree of a polynomial or interaction expansion.
+When used as a superscript, $K$ indicates that element below it is in a polynomially expanded context.
+For example, if $nnz_i$ is the number of nonezero elements in the vector $i$th row vector of some matrix, $nnz_i^K$ is the number of nonzero elements in the polynomial expansion of that row vector.
+If lowercase $k$ refers to a column index.
 
@@
 For example, if the CSR for $\bm{A}$ is $\bm{c}, \bm{d}, \bm{p}$, then $\bm{c}^2$ is the vector that holds columns for nonzero values in $\bm{A}$'s quadratic feature expansion CSR representation.
-
-To simplify the presentation, we work with polynomial expansions of degree 2, but continue to use the exponent $k$ to show how the ideas apply in the general case.
-
-We do provide an algorithm for third degree expansions, and derive the big-O time complexity of the general case.
-
-We have also developed an algorithm for second and third degree interaction features (combinations without repetition), which can be found in the implementation.
 
@@
 The final design matrix is the augmentation of all such expansions.
-However, the algorithm descriptions in this paper omit these steps as they would become unnecessarily visually cumbersome.
-Extending them to include all degrees less than $k$ is trivial and does not affect the complexity of the algorithm as the terms that involve $k$ dominate.
+However, the algorithm descriptions in this paper omit these steps as they would become unnecessarily visually and notationally cumbersome.
+Extending them to include all degrees less than $K$ is trivial and does not affect the complexity of the algorithm as the terms that involve $k$ dominate.
 
-\subsection{Dense Expansion Algorithm}
-A natural way to calculate polynomial features for a matrix $\bm{A}$ is to walk down its rows and, for each row, take products of all $k$-combinations of elements.
-To determine in which column of $\bm{A}^k_i$ products of elements in $\bm{A}_i$ belong, a simple counter can be set to zero for each row of $\bm{A}$ and incremented after each polynomial feature is generated.
-This counter gives the column of $\bm{A}^k_i$ into which each expansion feature belongs.
+\subsection{Dense Second Degree Polynomial Expansion Algorithm}
+A natural way to calculate polynomial features for a matrix $\bm{A}$ is to walk down its rows and, for each row, take products of all $K$-combinations of elements.
+To determine in which column of $\bm{A}^K_i$ products of elements in $\bm{A}_i$ belong, a simple counter can be set to zero for each row of $\bm{A}$ and incremented after each polynomial feature is generated.
+This counter gives the column of $\bm{A}^K_i$ into which each expansion feature belongs.
+This is shown in Algorithm \ref{alg:Dense-Second-Order-Polynomial-Expansion}.
 
-\begin{algorithm}[H]
+% BEGIN vrtically centered
+%\pagebreak
+%\hspace{0pt}
+%\vfill
+
+\begin{algorithm}%[H]
    \caption{Dense Second Order Polynomial Expansion}
@@
    \STATE {\bfseries Input:} data $\bm{A}$, size $N \times D$
-   \STATE $\bm{A}^k$ $\gets$ empty $N \times \binom{D}{2}$ matrix
+   \STATE $\bm{A}^K$ $\gets$ empty $N \times \binom{D}{2}$ matrix
    \FOR{$i \gets 0$ {\bfseries to} $N-1$}
@@
           \FOR{$j_2 \gets j_1$ {\bfseries to} $D-1$}
-              \STATE $\bm{A}^k_{i{c_p}} \gets \bm{A}_{ij_1} \cdot \bm{A}_{ij_2}$
+              \STATE $\bm{A}^K_{i{c_p}} \gets \bm{A}_{ij_1} \cdot \bm{A}_{ij_2}$
               \STATE $c_p \gets c_p + 1$
@@
 
-\subsection{Imperfect CSR Expansion Algorithm}
+%\vskip 1.5in
+
+\subsection{Incomplete Second Degree CSR Polynomial Expansion Algorithm}
 \label{sec:final-algo}
 Now consider how this algorithm might be modified to accept a CSR matrix.
-Instead of walking directly down rows of $\bm{A}$, we will walk down sections of $\bm{c}$ and $\bm{d}$ partitioned by $\bm{p}$, and instead of inserting polynomial features into $\bm{A}^k$, we will insert column numbers into $\bm{c}^k$ and data elements into $\bm{d}^k$.
+Instead of walking directly down rows of $\bm{A}$, we will walk down sections of $\bm{c}$ and $\bm{d}$ partitioned by $\bm{p}$, and instead of inserting polynomial features into $\bm{A}^K$, we will insert column numbers into $\bm{c}^K$ and data elements into $\bm{d}^K$.
+Throughout the algorithm, we use variables named $nnz$, with sub- or superscripts, to indicate the number of nonzero entries in either a matrix or a row of a matrix. 
+See Algorithm \ref{alg:Incomplete-Sparse-Second-Order-Polynomial-Expansion}.
 
-\begin{algorithm}[H]
-   \caption{Incomplete Sparse Second Order Polynomial Expansion}
+%\vfill
+%\hspace{0pt}
+%\pagebreak
+% END vrtically centered
+
+\begin{algorithm}%[H]
+   \caption{Sparse Second Order Polynomial Expansion}
    \label{alg:Incomplete-Sparse-Second-Order-Polynomial-Expansion}
@@
    \STATE {\bfseries Input:} data $\bm{A}$, size $N \times D$
-   \STATE $\bm{p}^k$ $\gets$ vector of size $N+1$
-   \STATE $\bm{p}^k_0 \gets 0$
-   \STATE $nnz^k \gets 0$
+   \STATE $\bm{p}^K$ $\gets$ vector of size $N+1$
+   \STATE $\bm{p}^K_0 \gets 0$
+   \STATE $nnz^K \gets 0$
    \FOR{$i \gets 0$ {\bfseries to} $N-1$}
@@
       \STATE $\bm{c}_i \gets \bm{c}_{i_{start}:i_{stop}}$
-      \STATE $nnz^k_i \gets \binom{|\bm{c}_i|}{2}$ \label{li:row_nnz_count}
-      \STATE $nnz^k \gets nnz^k + nnz^k_i$
-      \STATE $\bm{p}^k_{i+1} \gets \bm{p}^k_i + nnz^k_i$
+      \STATE $nnz^K_i \gets \binom{|\bm{c}_i|}{2}$ \label{li:row_nnz_count}
+      \STATE $nnz^K \gets nnz^K + nnz^K_i$
+      \STATE $\bm{p}^K_{i+1} \gets \bm{p}^K_i + nnz^K_i$
   \ENDFOR
 
-  \STATE $\bm{p}^k$ $\gets$ vector of size $N+1$
-  \STATE $\bm{c}^k$ $\gets$ vector of size $nnz^k$
-  \STATE $\bm{d}^k$ $\gets$ vector of size $nnz^k$
+  \STATE $\bm{p}^K$ $\gets$ vector of size $N+1$
+  \STATE $\bm{c}^K$ $\gets$ vector of size $nnz^K$
+  \STATE $\bm{d}^K$ $\gets$ vector of size $nnz^K$
   \STATE $n \gets 0$
@@
           \FOR {$c_2 \gets c_1$ {\bfseries to} $|\bm{c}_i|-1$}
-              \STATE $\bm{d}^k_{n} \gets \bm{d}_{c_0} \cdot \bm{d}_{c_1}$
-              \STATE $\bm{c}^k_{n} = ?$ \label{li:set_ck}
+              \STATE $\bm{d}^K_{n} \gets \bm{d}_{c_0} \cdot \bm{d}_{c_1}$
+              \STATE $\bm{c}^K_{n} = ?$ \label{li:set_ck}
               \STATE $n \gets n + 1$
@@
 
-The crux of the problem is at line 25. %reference wasn't working \ref{li:set_ck}.
-Given the arbitrary columns involved in a polynomial feature of $\bm{A}_i$, we need to determine the corresponding column of $\bm{A}^k_i$.
+The crux of the problem is at line 25 of Algorithm \ref{alg:Incomplete-Sparse-Second-Order-Polynomial-Expansion}.
+Given the arbitrary columns involved in a polynomial feature of $\bm{A}_i$, we need to determine the corresponding column of $\bm{A}^K_i$.
 We cannot simply reset a counter for each row as we did in the dense algorithm,  because only columns corresponding to nonzero values are stored.
@@
 To develop a general algorithm, we require a mapping from a list of $k$ columns of $\bm{A}$ to a column of $\bm{A}^k$.
-If there are $D$ columns of $\bm{A}$ and $\binom{D}{k}$ columns of $\bm{A}^k$, this can be accomplished by a bijective mapping of the following form:
+If there are $D$ columns of $\bm{A}$ and $\binom{D}{K}$ columns of $\bm{A}^K$, this can be accomplished by a bijective mapping of the following form:
 
 \begin{equation}
-(j_0, j_1, \dots, j_{k-1}) \rightarrowtail \hspace{-1.9ex} \twoheadrightarrow p_{j_0j_1 \dots i_{k-1}} \in \{0,1,\dots,\binom{D}{k}-1\} 
+(j_0, j_1, \dots, j_{K-1}) \rightarrowtail \hspace{-1.9ex} \twoheadrightarrow p_{j_0j_1 \dots i_{K-1}} \in \{0,1,\dots,\binom{D}{K}-1\} 
 \end{equation}
 
-where (i) $ 0 \le j_0 \le j_1 \le \dots \le j_{k-1} < D$, (ii) $(j_0, j_1, \dots, j_{k-1})$ are elements of $\bm{c}$ and (iii) $p_{j_0j_1 \dots i_{k-1}}$ is an element of $\bm{c}^k$.
+where (i) $ 0 \le j_0 \le j_1 \le \dots \le j_{K-1} < D$, (ii) $(j_0, j_1, \dots, j_{K-1})$ are elements of $\bm{c}$ and (iii) $p_{j_0j_1 \dots i_{K-1}}$ is an element of $\bm{c}^K$.
+(For interaction features, the constraint is $ 0 \le j_0 < j_1 < \dots < j_{K-1} < D$.)
+
+Stated more verbosely, we require a bijective mapping from tuples consisting of column indicies of the original input to where the column index of the corresponding product of features in the polynomial expansion.
+While any bijective mapping would suffice, a common order in which to produce polynomial features is $(0, 1), (0, 2), \dots, (0, D-1), (1, 1), (1, 2), \dots (1, D-1), \dots (D-1, D-1)$ for $K=2$ where the elements of the tuples are column indices.
+That is, the further to the right an index is, the sooner it is incremented.
+We refer to this order as \emph{big endian order} as it can be thought of as the way digits in a big endian number system are incremented when counting.
+If we want for our algorithm to be backwards compatible with existing models, the mapping must use the big endian ordering.
+
 %column indicies of a row vector $\vec{x}$ of an $N \times D$ input matrix, and $p_{i_0i_1 \dots i_{k-1}}$ is a column index into the polynomial expansion vector for $\vec{x}$ where the product of elements corresponding to indices $i_0, i_1, \dots, i_{k-1}$ will be stored.
 
-\section{Construction of Mapping}
+\section{Construction of Mappings}
 %\textbf{\color{red} This looks like the wrong mapping, since it uses $ i < j$ rather than $i \le j$, i.e., it's "interaction features" rather than polynomial features.
@@
 Within this section, $i$, $j$, and $k$ denote column indices.
-We start with the second degree case for \emph{interaction} features due to its relative simplicity as it is easiest to understand.
-We seek a map from matrix indices $(i, j)$ (with $0 \le i < j < D$ ) to numbers $f(i, j)$ with $0 \le f(i, j) < \frac{D(D-1)}{2}$ that follows the pattern indicated by
+We will construct mappings for second ($K=2$) and third ($K=3$) degree interaction and polynomial expansions.
+In order to accomplish this, we will require the triangle and tetrahedral numbers.
+We denote the $n$th triangle number as $T_2(n) = \frac{n(n+1)}{2}$ and the $n$th tetrahedral number as $T_3(n) = \frac{n(n+1)(n+2)}{6}$.
+
+For reference, we list the first four triangle and tetrahedral numbers in the following table:
+
+\begin{tabular}{| c | c | c |}
+  \hline
+  $n$ & $T_2(n)$ & $T_3(n)$ \\
+  \hline
+  0 & 0 & 0 \\
+  1 & 1 & 1 \\
+  2 & 3 & 4 \\
+  3 & 6 & 10 \\
+  4 & 10 & 20 \\
+  \hline  
+\end{tabular}
+
+\subsection{Second Degree Interaction Mapping}
+For second order interaction features, we require a bijective function that maps the elements of the ordered set
+\begin{equation}
+((0, 1), (0, 2), \dots, (1, 2), (1, 3), \dots, (D-2, D-1))
+\end{equation}
+to the elements of the ordered set
+\begin{equation}
+(0,1,\dots,\binom{D-1}{2}-1)
+\end{equation}
+
+For $D=4$, we can view the desired mapping $f$ as one that maps the coordinates of matrix cells to $0, 1, 2, 3$.
+If we fill the cells of the matrix with the codomain, the target matrix is as follows:
 \begin{align}
 \begin{bmatrix}
-x & 0 & 1 & 3 \\
-x & x & 2 & 4 \\
+x & 0 & 1 & 2 \\
+x & x & 3 & 4 \\
 x & x & x & 5 \\
@@
 \end{align}
-where the entry in row $i$, column $j$, displays the value $f(i, j)$. We let $T_2(n) = \frac{1}{2} n(n+1)$ 
-be the $n$th triangular number; then in Equation~\ref{eq:4x4mat}, column $j$ (for $j > 0$) contains entries with  
-$T_2(j-1) \le e < T_2(j)$; the entry in the $i$th row is just $i + T_2(j-1)$. Thus we have
-$
-f(i, j) 
-= i + T_2(j-1) =  \frac{1}{2}(2i + j^2-j).$
-For instance, in column $j = 2$ in our example (the \emph{third} column), the entry in row $i = 1$ is 
-$i + T_2(j-1) = 1 + 1 = 2$.
+where the entry in row $i$, column $j$, displays the value of $f(i, j)$.
 
-With one-based indexing in both the domain and codomain, the formula above becomes
-$f_1(i, j)  = \frac{1}{2}(2i + j^2 - 3j + 2).$
+It will be simpler to instead construct a preliminary mapping, $r(i, j)$ of the following form:
+\begin{align}
+\begin{bmatrix}
+x & 6 & 5 & 4 \\
+x & x & 3 & 2 \\
+x & x & x & 1 \\
+x & x & x & x
+\end{bmatrix}
+\label{eq:preliminary4x4}
+\end{align}
+and then subtract the preliminary mapping from the total number of elements in the codomain to create the final mapping.
+Note that in equation \ref{eq:preliminary4x4} we have the following relation:
 
-For \emph{polynomial} features, we seek a similar map $g$, one that also handles the case $i = j$. In this case, a similar analysis yields
-$ g(i, j) = i + T_2(j) = \frac{1}{2} (2i + j^2 + j + 1).$
+\begin{equation}
+T_2(D-i-2) < e^i \le T_2(D-i-1)
+\end{equation}
 
-To handle \emph{three-way interactions}, we need to map triples of indices in a 3-index array to a flat list, and similarly for higher-order interactions.
-For this, we require the tetrahedral numbers $T_3(n) = \sum_{i=1}^n T_{2}(n) =
-\frac{1}{6}(n^3 + 3n^2 + 2n)$.
+where $e^i$ is the value of any cell in row $i$ of equation \ref{eq:preliminary4x4}.
 
-For three indices, $i,j,k$, with $0 \le i < j < k < D$, we have a similar recurrence. Calling the mapping $h$, we have 
+Therefore, the following mapping will produce equation \ref{eq:preliminary4x4}:
+
+\begin{equation}
+r(i, j) = T_2(D-i-1) - (j - i - 1)
+\end{equation}
+
+We can now use this result to construct a mapping for equation \ref{4x4mat} by subtracting it from the size of the codomain:
+
+\begin{equation}
+f(i, j) = T_2(D-1) - [T_2(D-i-1) - (j - i - 1)]
+\end{equation}
+
+\subsection{Second Degree Polynomial Mapping}
+In this case, the target matrix is of the form
+
 \begin{align}
-h(i,j,k) = i + T_2(j-1) + T_3(k-2);
-\end{align}
-if we define $T_1(i) = i$, then this has the very regular form
-\begin{align}
-h(i,j,k) =  T_1(i) + T_2(j-1) + T_3(k-2);
-\end{align}
-and from this the generalization to higher dimensions is straightforward. The formulas for ``higher triangular numbers'', i.e., those defined by
-\begin{align}
-T_k(n) &= \sum_{i=1}^n T_{k-1}(n)
-\end{align}
-for $k > 1$ can be determined inductively.
-
-The explicit formula for 3-way interactions, with zero-based indexing, is 
-\begin{align}
-h(i, j, k) &= 1 + (i-1) + \frac{(j-1)j}{2} + \\
-& \frac{(k-2)^3 + 3(k-2)^2 + 2(k-2)}{6}. 
+\begin{bmatrix}
+0 & 1 & 2 & 3 \\
+x & 4 & 5 & 6 \\
+x & x & 7 & 8 \\
+x & x & x & 9
+\end{bmatrix}
+\label{eq:4x4mat}
 \end{align}
 
-\section{Final CSR Expansion Algorithm}
-With the mapping from columns of $\bm{A}$ to a column of $\bm{A}^k$, we can now write the final form of the innermost loop of the algorithm from \ref{sec:final-algo}.
-Let the mapping for $k=2$ be denoted $h^2$.
-Then the innermost loop becomes:
+A very similar analysis can be done for the $K=2$ case to yield
+
+\begin{equation}
+f(i, j) = T_2(D) - [T_2(D-i) - (j - i)]
+\end{equation}
+
+\subsection{Third Degree Interaction Mapping}
+For $K=3$, we can no longer view the necessary function as mapping matrix coordinates to cell values, but rather, $2x2$ tensor coordinates to cell values.
+For simplicity, we will instead list the column index tuples and their necessary mappings in a table.
+We shall consider the case of $D=5$.
+
+Again, it is simpler to find a mapping $r(i, j, k)$ that maps to the reverse the target indices (plus one) and create a final mapping by subtracting that mapping from the number of elements in the codomain.
+We therefore seek a preliminary mapping of the form
+
+\begin{tabular}{| c | c | c |}
+  \hline
+  $(i, k, j)$ & $r(i,j,k)$ & $f(i,j,k)$ \\
+  \hline
+  $(0, 1, 2)$ & 10 & 0 \\
+  $(0, 1, 3)$ & 9 & 1 \\
+  $(0, 1, 4)$ & 8 & 2 \\
+  $(0, 2, 3)$ & 7 & 3 \\
+  $(0, 2, 4)$ & 6 & 4 \\
+  $(0, 3, 4)$ & 5 & 5 \\
+  \hline
+  $(1, 2, 3)$ & 4 & 6 \\
+  $(1, 2, 4)$ & 3 & 7 \\
+  $(1, 3, 4)$ & 2 & 8 \\
+  \hline
+  $(2, 3, 4)$ & 1 & 9 \\
+  \hline  
+\end{tabular}
+
+The mapping has been partitioned according to the $i$ dimension.
+Note that within each partition is a mapping very similar to the $K=2$ equivalent, but with the indices shifted by a function of $T_3$.
+For example, when $i=0$, the indices are shifted by $T_3(2)$, when $i=1$, the shift is $T_3(1)$, and finally, when $i=2$, the shift is $T_3(0)$.
+The preliminary mapping is therefore
+\begin{equation}
+r(i, j, k) = T_3(D-i-3) + T_2(D-j-1) - (k-j-1)
+\end{equation}
+
+and the final mapping is therefore
+
+\begin{align}
+f(i, j, k) = T_3(D-2) - [&T_3(D-i-3) + \\
+                         &T_2(D-j-1) - \\
+                         &(k-j-1)]
+\end{align}
+
+\subsection{Third Degree Polynomial Mapping}
+The analysis of the $K=3$ polynomial case is very similar to that of the $K=3$ interaction case.
+However the target mapping now includes the edge of the simplex, as it included the diagonal of the matrix in the $K=2$ polynomial case.
+The analysis yields the mapping
+
+\begin{equation}
+f(i, j, k) = T_3(D) - [T_3(D-i-1) + T_2(D-j) - (k-j)]
+\end{equation}
+
+\section{Higher Order Mappings}
+It can be seen that mappings to higher orders can be constructed inductively.
+A $K$ degree mapping is a function of the $K$-simplex numbers and reparameterized versions of all lower order mappings.
+However, in practice, higher degrees are not often used as the dimensionality of the expanded vectors becomes prohibitively large.
+A second degree polynomial expansion of $D=1000$ would have $\binom{1000}{4} = 41,417,124,750$ dimensions.
+
+\section{Final CSR Polynomial Expansion Algorithm}
+With the mapping from columns of $\bm{A}$ to a column of $\bm{A}^K$, we can now write the final form of the innermost loop of the algorithm from \ref{sec:final-algo}.
+Let the polynomial mapping for $K=2$ be denoted $h^2$.
+Then the innermost loop can be completed as follows: %shown in Algorithm \ref{alg:Inner-Loop-of-Completed-Sparse-Second-Order-Polynomial-Expansion}.
 
 \begin{algorithm}[H]
-   \caption{Inner Loop of Completed Sparse Second Order Polynomial Expansion}
+   \caption*{Completed Inner Loop of Algorithm \ref{alg:Incomplete-Sparse-Second-Order-Polynomial-Expansion}}
    \label{alg:Inner-Loop-of-Completed-Sparse-Second-Order-Polynomial-Expansion}
@@
       \STATE $c_p \gets h^2(j_0, j_1)$
-      \STATE $\bm{d}^k_{n} \gets \bm{d}_{c_0} \cdot \bm{d}_{c_1}$
-      \STATE $\bm{c}^k_{n} = c_p$
+      \STATE $\bm{d}^K_{n} \gets \bm{d}_{c_0} \cdot \bm{d}_{c_1}$
+      \STATE $\bm{c}^K_{n} = c_p$
       \STATE $n \gets n + 1$
@@
 
+\section{Higher Degree and Interaction Algorithms}
+Most of the steps for higher degrees and interaction expansions (as opposed to polynomial) are the same as for the $K=2$ polynomial case.
+The differences are that for higher degrees, an extra loop is needed to iterate over another column index, and a different mapping is required.
+For interaction expansions, the column indices are never allowed to equal each other, so each loop executes one less time, and an interaction mapping is required.
+Also, for interaction expansions, the way $nnz$ is computed on line \ref{li:row_nnz_count} of algorithm \ref{alg:Incomplete-Sparse-Second-Order-Polynomial-Expansion}.
+Instead of $\binom{|\bm{c}_i|}{K}$, we have $\binom{|\bm{c}_i|-1}{K}$.
+
 \section{Time Complexity}
@@
 \begin{align}
-& \Theta\left(\binom{dD+k-1}{k}\right) = \Theta\left(\frac{(dD+k-1)!}{k!(dD-1)!}\right)\\
-& = \Theta\left(\frac{(dD+k-1)(dD+k-2) \dots (dD)}{k!}\right)\\
-& = \Theta\left((dD+k-1)(dD+k-2) \dots (dD)\right)\\
-& = \Theta\left(d^kD^k\right)
+& \Theta\left(\binom{dD+K-1}{k}\right) = \Theta\left(\frac{(dD+K-1)!}{K!(dD-1)!}\right)\\
+& = \Theta\left(\frac{(dD+K-1)(dD+K-2) \dots (dD)}{K!}\right)\\
+& = \Theta\left((dD+K-1)(dD+K-2) \dots (dD)\right)\\
+& = \Theta\left(d^KD^K\right)
 \end{align}
 
-For a matrix of size $N \times D$, the complexity is therefore $\Theta\left(Nd^kD^k\right)$.
-The dense algorithm does not leverage sparsity, and its complexity is $\Theta\left(ND^k\right)$.
+For a matrix of size $N \times D$, the complexity is therefore $\Theta\left(Nd^KD^K\right)$.
+The dense algorithm does not leverage sparsity, and its complexity is $\Theta\left(ND^K\right)$.
 Since $0 \le d \le 1$, the sparse algorithm scales polynomially with the degree of the polynomial expansion.
 
-\newpage
-
-\subsection{Empirical}
+\subsection{Empirical Results}
 To verify the average time complexity of our algorithm, we implemented both the sparse version and the baseline in the Cython programming language so that results would be directly comparable.
@@
 Varying the density ($d$) (column 1) shows that our algorithm scales polynomially with $d$, but that the baseline is unaffected by it.
-Both algorithms increase polynomially with the dimensionality ($D$), but ours at a significantly reduced rate.
+The runtimes for both algorithms increase polynomially with the dimensionality ($D$), but ours at a significantly reduced rate.
 Likewise, both algorithms scale linearly with the instance count ($N$), but ours to a much lesser degree.
@@
 \caption{A closer view of only the sparse runtimes while varying $D$ (left) and $N$ (right) for $d = 0.2$.
-The left subplot shows that varying $D$ gives polynomial growth in runtime; quadratic for $k = 2$ (dashed line) and cubic for $k = 3$ (dotted line).
+The left subplot shows that varying $D$ gives polynomial growth in runtime; quadratic for $K = 2$ (dashed line) and cubic for $K = 3$ (dotted line).
 These nonlinearities were not apparent in Figure \ref{fig:all-vs-time} due to the much greater runtimes of the dense algorithm.
@@
 
-\section{Discussion}
-A natural question to ask is ``Why not just use a decent hashing algorithm?'' There are two answers.
+%\section{Discussion}
+%A natural question to ask is ``Why not just use a decent hashing algorithm?'' There are two answers.
 The first is that doing so does not avoid the \emph{computation} of features that contain a zero factor, and that alone prevents a reduction in big-theta performance.

### 2018-03-08 sanity_check.py
adding a sanity check program to check the new mappings

@@ -1,0 +1,34 @@
+from itertools import combinations, combinations_with_replacement
+
+T2 = lambda n: (n * (n+1)) / 2
+T3 = lambda n: (n * (n+1) * (n+2)) / 6
+
+mappings = {'2i': lambda D, i, j: T2(D-1) - (T2(D-i-1) - (j-i-1)),
+            '2p': lambda D, i, j: T2(D) - (T2(D-i) - (j-i)),
+            '3i': lambda D, i, j, k: T3(D-2) - (T3(D-i-3) + T2(D-j-1) - (k-j-1)),
+            '3p': lambda D, i, j, k: T3(D) - (T3(D-i-1) + T2(D-j) - (k-j))}
+
+def DoIt(D, degree, kind):
+  assert kind in 'ip'
+  assert degree in (2, 3)
+  code = '%s%s' % (degree, kind)
+  mapping = mappings[code]
+  inds = range(D)
+  iter_funct = combinations if kind is 'i' else combinations_with_replacement
+  
+  expected = 0
+  for combo in iter_funct(inds, degree):
+    output_ind = mapping(D, *combo)
+    assert expected == output_ind
+    expected += 1
+  print expected
+  
+def DoAll():
+  for D in [5, 10, 20, 100]:
+    DoIt(D, 2, 'i')
+    DoIt(D, 2, 'p')
+    DoIt(D, 3, 'i')
+    DoIt(D, 3, 'p')
+
+if __name__ == '__main__':
+  DoAll()

### 2018-03-08 paper/Final/sparse_poly_icml.tex
fixed a lot of capitalization of K issues, removed some sentences, clarified some things.

@@
 % Therefore, a short form for the running title is supplied here:
-\icmltitlerunning{Speeding Up Polynomial Feature Expansions of Compressed Sparse Row Matrices}
+\icmltitlerunning{Speeding Up Polynomial Feature Expansions of Sparse CSR Matrices Using $K$-Simplex Numbers}
 
@@
 \twocolumn[
-\icmltitle{Speeding Up Polynomial Feature Expansions by Leveraging Density}
 
@@
 
-\begin{icmlauthorlist}
-\icmlauthor{Andrew Nystrom}{google}
-\icmlauthor{John Hughes}{brown}
-\end{icmlauthorlist}
-
-\icmlaffiliation{google}{Google Inc.}
-\icmlaffiliation{brown}{Brown University}
-
-\icmlcorrespondingauthor{Andrew Nystrom}{nystrom@google.com}
-\icmlcorrespondingauthor{John Hughes}{jfh@cs.brown.edu}
-
-% You may provide any keywords that you 
-% find helpful for describing your paper; these are used to populate 
-% the "keywords" metadata in the PDF but will not be shown in the document
-\icmlkeywords{boring formatting information, machine learning, ICML}
+\author{
+  Nystrom, Andrew\\
+  \texttt{nystrom@google.com}
+  \and
+  Hughes, John\\
+  \texttt{jfh@cs.brown.edu}
+}
+\title{Speeding Up Polynomial Feature Expansions of Sparse CSR Matrices Using $K$-Simplex Numbers}
+\maketitle
 
@@
 ]
+
+\date{}
 
@@
 %\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
-\printAffiliationsAndNotice{\icmlEqualContribution} % otherwise use the standard text.
+%\printAffiliationsAndNotice{\icmlEqualContribution} % otherwise use the standard text.
 %\footnotetext{hi}
@@
 The algorithm operates on and produces a compressed sparse row matrix with no densification.
-It works by mapping (via simplex numbers) column indices with nonzero elements to a column index in the polynomial expansion vector, allowing for only the nonzero elements in a CSR matrix to be iterated over when computing products.
+It works by defining a bijective function involving simplex numbers of column indices in the original matrix to column indices in the expanded matrix.
+This function allows for only the nonzero elements to be iterated over and multiplied together, greatly improving the expansion of sparse matrices in compressed sparse row format.
 For a vector of dimension $D$ and density $0 \le d \le 1$, the algorithm has time complexity $\Theta(d^KD^K)$ where $K$ is the polynomial-feature order; this  is an improvement by a factor $d^K$ over the standard method.
@@
 A well known and widely used feature mapping is the \emph{polynomial expansion}, which produces a new feature for each degree-$k$ monomial in the original features.
-(If the original features are $x,y,z$, the order-2 polynomial features are $x^2, y^2, z^2, xy, xz, xy$, and the order-3 ones are $x^3, y^3, z^3, x^2y, x^2z, xy^2, y^2 z, xz^2, yz^2,$ and $xyz$.)
-A $k$-order polynomial feature expansion of the feature space allows a linear model to learn polynomial relationships between dependent and independent variables.
+(If the original features are $x,y,z$, the order-2 polynomial features are $x^2, y^2, z^2, xy, xz, xy$, and the order-3 ones are $x^3, y^3, z^3, x^2y, x^2z, xy^2, y^2 z, xz^2, yz^2,$ and $xyz$.
+A $K$-order polynomial feature expansion of the feature space allows a linear model to learn polynomial relationships between dependent and independent variables.
 This mapping was first utilized in a published experiment by Joseph Diaz Gergonne in 1815 \cite{gergonne1974application, smith1918standard}.
@@
 
-Uppercase $K$ refers to the degree of a polynomial or interaction expansion.
-When used as a superscript, $K$ indicates that element below it is in a polynomially expanded context.
-For example, if $nnz_i$ is the number of nonezero elements in the vector $i$th row vector of some matrix, $nnz_i^K$ is the number of nonzero elements in the polynomial expansion of that row vector.
-If lowercase $k$ refers to a column index.
-
 Scalars, vectors, and matrices are often decorated with a superscript $k$, which is not to be interpreted as an exponent, but instead as an indicator of polynomial expansion: 
 For example, if the CSR for $\bm{A}$ is $\bm{c}, \bm{d}, \bm{p}$, then $\bm{c}^2$ is the vector that holds columns for nonzero values in $\bm{A}$'s quadratic feature expansion CSR representation.
+
+Uppercase $K$ refers to the degree of a polynomial or interaction expansion.
+When used as a superscript of a non-scalar, $K$ indicates that element below it is in a polynomially expanded context.
+For example, if $nnz_i$ is the number of nonezero elements in the vector $i$th row vector of some matrix, $nnz_i^K$ is the number of nonzero elements in the polynomial expansion of that row vector.
+Lowercase $k$ refers to a column index.
 
@@
 However, the algorithm descriptions in this paper omit these steps as they would become unnecessarily visually and notationally cumbersome.
-Extending them to include all degrees less than $K$ is trivial and does not affect the complexity of the algorithm as the terms that involve $k$ dominate.
+Extending them to include all degrees less than $K$ is trivial and does not affect the complexity of the algorithm as the terms that involve $K$ dominate.
 
@@
 
-To develop a general algorithm, we require a mapping from a list of $k$ columns of $\bm{A}$ to a column of $\bm{A}^k$.
+To develop a general algorithm, we require a mapping from a list of $K$ columns of $\bm{A}$ to a single column of $\bm{A}^K$.
 If there are $D$ columns of $\bm{A}$ and $\binom{D}{K}$ columns of $\bm{A}^K$, this can be accomplished by a bijective mapping of the following form:
@@
 Stated more verbosely, we require a bijective mapping from tuples consisting of column indicies of the original input to where the column index of the corresponding product of features in the polynomial expansion.
-While any bijective mapping would suffice, a common order in which to produce polynomial features is $(0, 1), (0, 2), \dots, (0, D-1), (1, 1), (1, 2), \dots (1, D-1), \dots (D-1, D-1)$ for $K=2$ where the elements of the tuples are column indices.
+While any bijective mapping would suffice, a common order in which to produce polynomial features is $(0, 1)$, $(0, 2)$, $\dots$, $(0, D-1)$, $(1, 1)$, $(1, 2)$, $\dots$, $(1, D-1)$, $\dots$, $(D-1, D-1)$ for $K=2$ where the elements of the tuples are column indices.
 That is, the further to the right an index is, the sooner it is incremented.
-We refer to this order as \emph{big endian order} as it can be thought of as the way digits in a big endian number system are incremented when counting.
-If we want for our algorithm to be backwards compatible with existing models, the mapping must use the big endian ordering.
+If we want for our algorithm to be backwards compatible with existing models, the mapping must use the this same ordering.
 
@@
 
-For reference, we list the first four triangle and tetrahedral numbers in the following table:
+For reference, we list the first five triangle and tetrahedral numbers in the following table:
 
@@
 
-Calculating $k$-degree polynomial features via our method for a vector of dimensionality $D$ and density $d$ requires $\binom{dD}{k}$ (with repetition) products.
-The complexity of the algorithm, for fixed $k \ll dD$, is therefore
+Calculating $K$-degree polynomial features via our method for a vector of dimensionality $D$ and density $d$ requires $\binom{dD}{K}$ products.
+The complexity of the algorithm, for fixed $K \ll dD$, is therefore
 \begin{align}
-& \Theta\left(\binom{dD+K-1}{k}\right) = \Theta\left(\frac{(dD+K-1)!}{K!(dD-1)!}\right)\\
+& \Theta\left(\binom{dD+K-1}{K}\right) = \Theta\left(\frac{(dD+K-1)!}{K!(dD-1)!}\right)\\
 & = \Theta\left(\frac{(dD+K-1)(dD+K-2) \dots (dD)}{K!}\right)\\
@@
 For a matrix of size $N \times D$, the complexity is therefore $\Theta\left(Nd^KD^K\right)$.
-The dense algorithm does not leverage sparsity, and its complexity is $\Theta\left(ND^K\right)$.
+The dense algorithm (algorithm \ref{alg:Dense-Second-Order-Polynomial-Expansion}) does not leverage sparsity, and its complexity is $\Theta\left(ND^K\right)$.
 Since $0 \le d \le 1$, the sparse algorithm scales polynomially with the degree of the polynomial expansion.
@@
 \subsection{Empirical Results}
-To verify the average time complexity of our algorithm, we implemented both the sparse version and the baseline in the Cython programming language so that results would be directly comparable.
+To empirically verify the average time complexity of our algorithm, we implemented both the sparse version and the baseline in the Cython programming language so that results would be directly comparable.
 We sought the relationships between runtime and the instance count ($N$), the instance dimensionality ($D$), and the instance density ($d$).

### 2018-03-08 paper/Final/sparse_poly_icml.tex
fixed tensor size

@@
 % Therefore, a short form for the running title is supplied here:
-\icmltitlerunning{Speeding Up Polynomial Feature Expansions of Sparse CSR Matrices Using $K$-Simplex Numbers}
+\icmltitlerunning{Leveraging Sparsity to Speed Up Polynomial Feature Expansions of CSR Matrices Using $K$-Simplex Numbers}
 
@@
 \twocolumn[
+\icmltitle{Leveraging Sparsity to Speed Up Polynomial Feature Expansions of CSR Matrices Using $K$-Simplex Numbers}
 
@@
 
-\author{
-  Nystrom, Andrew\\
-  \texttt{nystrom@google.com}
-  \and
-  Hughes, John\\
-  \texttt{jfh@cs.brown.edu}
-}
-\title{Speeding Up Polynomial Feature Expansions of Sparse CSR Matrices Using $K$-Simplex Numbers}
-\maketitle
+\begin{icmlauthorlist}
+\icmlauthor{Andrew Nystrom}{google}
+\icmlauthor{John Hughes}{brown}
+\end{icmlauthorlist}
+
+\icmlaffiliation{google}{Google Inc.}
+\icmlaffiliation{brown}{Brown University}
+
+\icmlcorrespondingauthor{Andrew Nystrom}{nystrom@google.com}
+\icmlcorrespondingauthor{John Hughes}{jfh@cs.brown.edu}
+
+% You may provide any keywords that you 
+% find helpful for describing your paper; these are used to populate 
+% the "keywords" metadata in the PDF but will not be shown in the document
+\icmlkeywords{csr, polynomial, sparse, expansion, triangle, tetrahedral, simplex}
 
@@
 ]
-
-\date{}
 
@@
 
-%\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
+\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
 %\printAffiliationsAndNotice{\icmlEqualContribution} % otherwise use the standard text.
@@
 \subsection{Third Degree Interaction Mapping}
-For $K=3$, we can no longer view the necessary function as mapping matrix coordinates to cell values, but rather, $2x2$ tensor coordinates to cell values.
+For $K=3$, we can no longer view the necessary function as mapping matrix coordinates to cell values, but rather, $DxDxD$ tensor coordinates to cell values.
 For simplicity, we will instead list the column index tuples and their necessary mappings in a table.

### 2018-03-16 paper/Final/sparse_poly_icml.tex
fourth degree, not second
using kappa to indicate polynomial context

@@
 
-\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
+%\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
 %\printAffiliationsAndNotice{\icmlEqualContribution} % otherwise use the standard text.
@@
 Uppercase $K$ refers to the degree of a polynomial or interaction expansion.
-When used as a superscript of a non-scalar, $K$ indicates that element below it is in a polynomially expanded context.
-For example, if $nnz_i$ is the number of nonezero elements in the vector $i$th row vector of some matrix, $nnz_i^K$ is the number of nonzero elements in the polynomial expansion of that row vector.
+When a superscript $\kappa$ (kappa) appears, it indicates that the element below it is in a polynomially expanded context of degree $K$.
+For example, if $nnz_i$ is the number of nonezero elements in the vector $i$th row vector of some matrix, $nnz_i^\kappa$ is the number of nonzero elements in the polynomial expansion of that row vector.
 Lowercase $k$ refers to a column index.
@@
 A natural way to calculate polynomial features for a matrix $\bm{A}$ is to walk down its rows and, for each row, take products of all $K$-combinations of elements.
-To determine in which column of $\bm{A}^K_i$ products of elements in $\bm{A}_i$ belong, a simple counter can be set to zero for each row of $\bm{A}$ and incremented after each polynomial feature is generated.
-This counter gives the column of $\bm{A}^K_i$ into which each expansion feature belongs.
+To determine in which column of $\bm{A}^\kappa_i$ products of elements in $\bm{A}_i$ belong, a simple counter can be set to zero for each row of $\bm{A}$ and incremented after each polynomial feature is generated.
+This counter gives the column of $\bm{A}^\kappa_i$ into which each expansion feature belongs.
 This is shown in Algorithm \ref{alg:Dense-Second-Order-Polynomial-Expansion}.
@@
    \STATE {\bfseries Input:} data $\bm{A}$, size $N \times D$
-   \STATE $\bm{A}^K$ $\gets$ empty $N \times \binom{D}{2}$ matrix
+   \STATE $\bm{A}^\kappa$ $\gets$ empty $N \times \binom{D}{2}$ matrix
    \FOR{$i \gets 0$ {\bfseries to} $N-1$}
@@
           \FOR{$j_2 \gets j_1$ {\bfseries to} $D-1$}
-              \STATE $\bm{A}^K_{i{c_p}} \gets \bm{A}_{ij_1} \cdot \bm{A}_{ij_2}$
+              \STATE $\bm{A}^\kappa_{i{c_p}} \gets \bm{A}_{ij_1} \cdot \bm{A}_{ij_2}$
               \STATE $c_p \gets c_p + 1$
@@
 Now consider how this algorithm might be modified to accept a CSR matrix.
-Instead of walking directly down rows of $\bm{A}$, we will walk down sections of $\bm{c}$ and $\bm{d}$ partitioned by $\bm{p}$, and instead of inserting polynomial features into $\bm{A}^K$, we will insert column numbers into $\bm{c}^K$ and data elements into $\bm{d}^K$.
+Instead of walking directly down rows of $\bm{A}$, we will walk down sections of $\bm{c}$ and $\bm{d}$ partitioned by $\bm{p}$, and instead of inserting polynomial features into $\bm{A}^\kappa$, we will insert column numbers into $\bm{c}^\kappa$ and data elements into $\bm{d}^\kappa$.
 Throughout the algorithm, we use variables named $nnz$, with sub- or superscripts, to indicate the number of nonzero entries in either a matrix or a row of a matrix. 
@@
    \STATE {\bfseries Input:} data $\bm{A}$, size $N \times D$
-   \STATE $\bm{p}^K$ $\gets$ vector of size $N+1$
-   \STATE $\bm{p}^K_0 \gets 0$
-   \STATE $nnz^K \gets 0$
+   \STATE $\bm{p}^\kappa$ $\gets$ vector of size $N+1$
+   \STATE $\bm{p}^\kappa_0 \gets 0$
+   \STATE $nnz^\kappa \gets 0$
    \FOR{$i \gets 0$ {\bfseries to} $N-1$}
@@
       \STATE $\bm{c}_i \gets \bm{c}_{i_{start}:i_{stop}}$
-      \STATE $nnz^K_i \gets \binom{|\bm{c}_i|}{2}$ \label{li:row_nnz_count}
-      \STATE $nnz^K \gets nnz^K + nnz^K_i$
-      \STATE $\bm{p}^K_{i+1} \gets \bm{p}^K_i + nnz^K_i$
+      \STATE $nnz^\kappa_i \gets \binom{|\bm{c}_i|}{2}$ \label{li:row_nnz_count}
+      \STATE $nnz^\kappa \gets nnz^\kappa + nnz^\kappa_i$
+      \STATE $\bm{p}^\kappa_{i+1} \gets \bm{p}^\kappa_i + nnz^\kappa_i$
   \ENDFOR
 
-  \STATE $\bm{p}^K$ $\gets$ vector of size $N+1$
-  \STATE $\bm{c}^K$ $\gets$ vector of size $nnz^K$
-  \STATE $\bm{d}^K$ $\gets$ vector of size $nnz^K$
+  \STATE $\bm{p}^\kappa$ $\gets$ vector of size $N+1$
+  \STATE $\bm{c}^\kappa$ $\gets$ vector of size $nnz^\kappa$
+  \STATE $\bm{d}^\kappa$ $\gets$ vector of size $nnz^\kappa$
   \STATE $n \gets 0$
@@
           \FOR {$c_2 \gets c_1$ {\bfseries to} $|\bm{c}_i|-1$}
-              \STATE $\bm{d}^K_{n} \gets \bm{d}_{c_0} \cdot \bm{d}_{c_1}$
-              \STATE $\bm{c}^K_{n} = ?$ \label{li:set_ck}
+              \STATE $\bm{d}^\kappa_{n} \gets \bm{d}_{c_0} \cdot \bm{d}_{c_1}$
+              \STATE $\bm{c}^\kappa_{n} = ?$ \label{li:set_ck}
               \STATE $n \gets n + 1$
@@
 The crux of the problem is at line 25 of Algorithm \ref{alg:Incomplete-Sparse-Second-Order-Polynomial-Expansion}.
-Given the arbitrary columns involved in a polynomial feature of $\bm{A}_i$, we need to determine the corresponding column of $\bm{A}^K_i$.
+Given the arbitrary columns involved in a polynomial feature of $\bm{A}_i$, we need to determine the corresponding column of $\bm{A}^\kappa_i$.
 We cannot simply reset a counter for each row as we did in the dense algorithm,  because only columns corresponding to nonzero values are stored.
@@
 
-To develop a general algorithm, we require a mapping from a list of $K$ columns of $\bm{A}$ to a single column of $\bm{A}^K$.
-If there are $D$ columns of $\bm{A}$ and $\binom{D}{K}$ columns of $\bm{A}^K$, this can be accomplished by a bijective mapping of the following form:
+To develop a general algorithm, we require a mapping from a list of $K$ columns of $\bm{A}$ to a single column of $\bm{A}^\kappa$.
+If there are $D$ columns of $\bm{A}$ and $\binom{D}{K}$ columns of $\bm{A}^\kappa$, this can be accomplished by a bijective mapping of the following form:
 
@@
 
-where (i) $ 0 \le j_0 \le j_1 \le \dots \le j_{K-1} < D$, (ii) $(j_0, j_1, \dots, j_{K-1})$ are elements of $\bm{c}$ and (iii) $p_{j_0j_1 \dots i_{K-1}}$ is an element of $\bm{c}^K$.
+where (i) $ 0 \le j_0 \le j_1 \le \dots \le j_{K-1} < D$, (ii) $(j_0, j_1, \dots, j_{K-1})$ are elements of $\bm{c}$ and (iii) $p_{j_0j_1 \dots i_{K-1}}$ is an element of $\bm{c}^\kappa$.
 (For interaction features, the constraint is $ 0 \le j_0 < j_1 < \dots < j_{K-1} < D$.)
@@
 However, in practice, higher degrees are not often used as the dimensionality of the expanded vectors becomes prohibitively large.
-A second degree polynomial expansion of $D=1000$ would have $\binom{1000}{4} = 41,417,124,750$ dimensions.
+A fourth degree polynomial expansion of a $D=1000$ vector would have $\binom{1000}{4} = 41,417,124,750$ dimensions.
 
 \section{Final CSR Polynomial Expansion Algorithm}
-With the mapping from columns of $\bm{A}$ to a column of $\bm{A}^K$, we can now write the final form of the innermost loop of the algorithm from \ref{sec:final-algo}.
+With the mapping from columns of $\bm{A}$ to a column of $\bm{A}^\kappa$, we can now write the final form of the innermost loop of the algorithm from \ref{sec:final-algo}.
 Let the polynomial mapping for $K=2$ be denoted $h^2$.
@@
       \STATE $c_p \gets h^2(j_0, j_1)$
-      \STATE $\bm{d}^K_{n} \gets \bm{d}_{c_0} \cdot \bm{d}_{c_1}$
-      \STATE $\bm{c}^K_{n} = c_p$
+      \STATE $\bm{d}^\kappa_{n} \gets \bm{d}_{c_0} \cdot \bm{d}_{c_1}$
+      \STATE $\bm{c}^\kappa_{n} = c_p$
       \STATE $n \gets n + 1$
@@
 
-Calculating $K$-degree polynomial features via our method for a vector of dimensionality $D$ and density $d$ requires $\binom{dD}{K}$ products.
+Calculating $K$-degree polynomial features via our method for a vector of dimensionality $D$ and density $d$ requires $T_K(dD)$ products.
 The complexity of the algorithm, for fixed $K \ll dD$, is therefore
 \begin{align}
-& \Theta\left(\binom{dD+K-1}{K}\right) = \Theta\left(\frac{(dD+K-1)!}{K!(dD-1)!}\right)\\
-& = \Theta\left(\frac{(dD+K-1)(dD+K-2) \dots (dD)}{K!}\right)\\
-& = \Theta\left((dD+K-1)(dD+K-2) \dots (dD)\right)\\
-& = \Theta\left(d^KD^K\right)
+&\Theta\left(T_K(dD)\right) =\\
+&\Theta\left(dD(dD+1)(dD+2)\dots(dD+K-1)\right) =\\
+&\Theta\left(d^KD^K\right)
 \end{align}

### 2018-04-11 paper/Final/nystrom_hughes_csr_polynomial_expansions.tex
Adding John Dibs's suggestions

--- 2018-03-16 paper/Final/sparse_poly_icml.tex
@@
 For a vector of dimension $D$ and density $0 \le d \le 1$, the algorithm has time complexity $\Theta(d^KD^K)$ where $K$ is the polynomial-feature order; this  is an improvement by a factor $d^K$ over the standard method.
+[Keywords: compressed sparse row, csr, feature expansion, feature mapping, polynomial expansion, sparse matrix]
 \end{abstract}
@@
 
-In machine learning and statistical modeling, feature mappings are intra-instance transformations, usually denoted by $x \mapsto \phi(\vec{x})$, that map instance vectors to higher dimensional spaces in which they are more linearly separable, allowing linear models to capture non-linearities \cite{yuan2012recent}.
+In machine learning and statistical modeling, feature mappings are intra-instance transformations, usually denoted by $x \mapsto \phi(\vec{x})$, that map instance vectors to higher dimensional spaces in which they are more linearly separable, allowing linear models to capture nonlinearities \cite{yuan2012recent}.
 
@@
 
-A compressed sparse row (CSR) matrix representation of an $r$-row matrix $\bm{A}$ consists of three vectors: $\bm{c}$, $\bm{d}$, and $\bm{p}$ and a single number: the number $N$ of columns of $\bm{A}$.
-The vectors
+A CSR matrix representation of an $r$-row matrix $\bm{A}$ consists of three vectors: $\bm{c}$, $\bm{d}$, and $\bm{p}$ and a single number: the number $N$ of columns of $\bm{A}$. The vectors
 $\bm{c}$ and $\bm{d}$ contain the same number of elements, and hold the column indices and data values, respectively, of all nonzero elements of $\bm{A}$.
@@
 \section{Motivation}
-In this section, we present a strawman algorithm for computing polynomial feature expansions on dense matrices.
+In this section we present a strawman algorithm for computing polynomial feature expansions on dense matrices.
 We then modify the algorithm  slightly to operate on a CSR matrix, to expose its infeasibility in that context.
-We then show how the algorithm would be feasible with an added component, which we then derive in the following section.
+We then show how the algorithm would be feasible with a bijective mapping from $k$-tuples of column indicies in the input matrix to column indices in the polynomial expansion matrix, which we then derive in the following section.
 
@@
 
-where (i) $ 0 \le j_0 \le j_1 \le \dots \le j_{K-1} < D$, (ii) $(j_0, j_1, \dots, j_{K-1})$ are elements of $\bm{c}$ and (iii) $p_{j_0j_1 \dots i_{K-1}}$ is an element of $\bm{c}^\kappa$.
+where (i) $ 0 \le j_0 \le j_1 \le \dots \le j_{K-1} < D$, (ii) $(j_0, j_1, \dots, j_{K-1})$ are elements of $\bm{c}$, and (iii) $p_{j_0j_1 \dots i_{K-1}}$ is an element of $\bm{c}^\kappa$.
 (For interaction features, the constraint is $ 0 \le j_0 < j_1 < \dots < j_{K-1} < D$.)
@@
 We will construct mappings for second ($K=2$) and third ($K=3$) degree interaction and polynomial expansions.
-In order to accomplish this, we will require the triangle and tetrahedral numbers.
+To accomplish this, we will require the triangle and tetrahedral numbers.
 We denote the $n$th triangle number as $T_2(n) = \frac{n(n+1)}{2}$ and the $n$th tetrahedral number as $T_3(n) = \frac{n(n+1)(n+2)}{6}$.
@@
 \begin{tabular}{| c | c | c |}
+  \caption{The first five triangle and tetrahedral numbers.}
   \hline
@@
 \subsection{Third Degree Interaction Mapping}
-For $K=3$, we can no longer view the necessary function as mapping matrix coordinates to cell values, but rather, $DxDxD$ tensor coordinates to cell values.
+For $K=3$ we can no longer view the necessary function as mapping matrix coordinates to cell values; rather, $DxDxD$ tensor coordinates to cell values.
 For simplicity, we will instead list the column index tuples and their necessary mappings in a table.
@@
 The analysis of the $K=3$ polynomial case is very similar to that of the $K=3$ interaction case.
-However the target mapping now includes the edge of the simplex, as it included the diagonal of the matrix in the $K=2$ polynomial case.
+However, the target mapping now includes the edge of the simplex, as it included the diagonal of the matrix in the $K=2$ polynomial case.
 The analysis yields the mapping
@@
 For interaction expansions, the column indices are never allowed to equal each other, so each loop executes one less time, and an interaction mapping is required.
-Also, for interaction expansions, the way $nnz$ is computed on line \ref{li:row_nnz_count} of algorithm \ref{alg:Incomplete-Sparse-Second-Order-Polynomial-Expansion}.
+Also, for interaction expansions, the way $nnz$ is computed on line \ref{li:row_nnz_count} of Algorithm \ref{alg:Incomplete-Sparse-Second-Order-Polynomial-Expansion}.
 Instead of $\binom{|\bm{c}_i|}{K}$, we have $\binom{|\bm{c}_i|-1}{K}$.
@@
 For a matrix of size $N \times D$, the complexity is therefore $\Theta\left(Nd^KD^K\right)$.
-The dense algorithm (algorithm \ref{alg:Dense-Second-Order-Polynomial-Expansion}) does not leverage sparsity, and its complexity is $\Theta\left(ND^K\right)$.
+The dense algorithm (Algorithm \ref{alg:Dense-Second-Order-Polynomial-Expansion}) does not leverage sparsity, and its complexity is $\Theta\left(ND^K\right)$.
 Since $0 \le d \le 1$, the sparse algorithm scales polynomially with the degree of the polynomial expansion.
@@
 Note that the point at which the sparse and dense algorithms intersect when varying $d$ is to the right of $0.5$, which is when a matrix technically becomes sparse.
-The point at which this happens will depend on $D$, but this shows that our algorithm is useful even for some dense matrices.
+The point at which this happens will depend on $D$, but the results show that our algorithm is useful even for some dense matrices.
 
@@
 
-\bibliography{sparse_poly_icml}
+\bibliography{nystrom_hughes_csr_polynomial_expansions}
 \bibliographystyle{icml2017}

### 2018-04-11 paper/Final/iccsda_sparse_poly.tex
iccsda ready

--- 2018-04-11 paper/Final/nystrom_hughes_csr_polynomial_expansions.tex
@@
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%% ICML 2017 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% bare_jrnl.tex
+%% V1.3
+%% 2007/01/11
+%% by Michael Shell
+%% see http://www.michaelshell.org/
+%% for current contact information.
+%%
+%% This is a skeleton file demonstrating the use of IEEEtran.cls
+%% (requires IEEEtran.cls version 1.7 or later) with an IEEE journal paper.
+%%
+%% Support sites:
+%% http://www.michaelshell.org/tex/ieeetran/
+%% http://www.ctan.org/tex-archive/macros/latex/contrib/IEEEtran/
+%% and
+%% http://www.ieee.org/
 
-% Use the following line _only_ if you're still using LaTeX 2.09.
-%\documentstyle[icml2017,epsf,natbib]{article}
-% If you rely on Latex2e packages, like most moden people use this:
-\documentclass{article}
+% *** Authors should verify (and, if needed, correct) their LaTeX system  ***
+% *** with the testflow diagnostic prior to trusting their LaTeX platform ***
+% *** with production work. IEEE's font choices can trigger bugs that do  ***
+% *** not appear when using other class files.                            ***
+% The testflow support page is at:
+% http://www.michaelshell.org/tex/testflow/
+
+%%*************************************************************************
+%% Legal Notice:
+%% This code is offered as-is without any warranty either expressed or
+%% implied; without even the implied warranty of MERCHANTABILITY or
+%% FITNESS FOR A PARTICULAR PURPOSE! 
+%% User assumes all risk.
+%% In no event shall IEEE or any contributor to this code be liable for
+%% any damages or losses, including, but not limited to, incidental,
+%% consequential, or any other damages, resulting from the use or misuse
+%% of any information contained here.
+%%
+%% All comments are the opinions of their respective authors and are not
+%% necessarily endorsed by the IEEE.
+%%
+%% This work is distributed under the LaTeX Project Public License (LPPL)
+%% ( http://www.latex-project.org/ ) version 1.3, and may be freely used,
+%% distributed and modified. A copy of the LPPL, version 1.3, is included
+%% in the base LaTeX documentation of all distributions of LaTeX released
+%% 2003/12/01 or later.
+%% Retain all contribution notices and credits.
+%% ** Modified files should be clearly indicated as such, including  **
+%% ** renaming them and changing author support contact information. **
+%%
+%% File list of work: IEEEtran.cls, IEEEtran_HOWTO.pdf, bare_adv.tex,
+%%                    bare_conf.tex, bare_jrnl.tex, bare_jrnl_compsoc.tex
+%%*************************************************************************
+
+% Note that the a4paper option is mainly intended so that authors in
+% countries using A4 can easily print to A4 and see how their papers will
+% look in print - the typesetting of the document will not typically be
+% affected with changes in paper size (but the bottom and side margins will).
+% Use the testflow package mentioned above to verify correct handling of
+% both paper sizes by the user's LaTeX system.
+%
+% Also note that the "draftcls" or "draftclsnofoot", not "draft", option
+% should be used if it is desired that the figures are to be displayed in
+% draft mode.
+%
+\documentclass[journal]{journal}
+%
+% If IEEEtran.cls has not been installed into the LaTeX system files,
+% manually specify the path to it like:
+% \documentclass[journal]{../sty/IEEEtran}
+
+% Some very useful LaTeX packages include:
+% (uncomment the ones you want to load)
+
+% *** MISC UTILITY PACKAGES ***
+%
+%\usepackage{ifpdf}
+% Heiko Oberdiek's ifpdf.sty is very useful if you need conditional
+% compilation based on whether the output is pdf or dvi.
+% usage:
+% \ifpdf
+%   % pdf code
+% \else
+%   % dvi code
+% \fi
+% The latest version of ifpdf.sty can be obtained from:
+% http://www.ctan.org/tex-archive/macros/latex/contrib/oberdiek/
+% Also, note that IEEEtran.cls V1.7 and later provides a builtin
+% \ifCLASSINFOpdf conditional that works the same way.
+% When switching from latex to pdflatex and vice-versa, the compiler may
+% have to be run twice to clear warning/error messages.
 
@@
 % For citations
-\usepackage{natbib}
+%\usepackage{natbib}
 
@@
 
-% Employ the following version of the ``usepackage'' statement for
-% submitting the draft version of the paper for review.  This will set
-% the note in the first column to ``Under review.  Do not distribute.''
-%\usepackage{icml2017}
+% *** CITATION PACKAGES ***
+%
+%\usepackage{cite}
+% cite.sty was written by Donald Arseneau
+% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
+% \cite{} output to follow that of IEEE. Loading the cite package will
+% result in citation numbers being automatically sorted and properly
+% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
+% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
+% \cite will automatically add leading space, if needed. Use cite.sty's
+% noadjust option (cite.sty V3.8 and later) if you want to turn this off.
+% cite.sty is already installed on most LaTeX systems. Be sure and use
+% version 4.0 (2003-05-27) and later if using hyperref.sty. cite.sty does
+% not currently provide for hyperlinked citations.
+% The latest version can be obtained at:
+% http://www.ctan.org/tex-archive/macros/latex/contrib/cite/
+% The documentation is contained in the cite.sty file itself.
 
-% Employ this version of the ``usepackage'' statement after the paper has
-% been accepted, when creating the final version.  This will set the
-% note in the first column to ``Proceedings of the...''
-\usepackage[accepted]{icml2017}
+% *** GRAPHICS RELATED PACKAGES ***
+%
+\ifCLASSINFOpdf
+  % \usepackage[pdftex]{graphicx}
+  % declare the path(s) where your graphic files are
+  % \graphicspath{{../pdf/}{../jpeg/}}
+  % and their extensions so you won't have to specify these with
+  % every instance of \includegraphics
+  % \DeclareGraphicsExtensions{.pdf,.jpeg,.png}
+\else
+  % or other class option (dvipsone, dvipdf, if not using dvips). graphicx
+  % will default to the driver specified in the system graphics.cfg if no
+  % driver is specified.
+  % \usepackage[dvips]{graphicx}
+  % declare the path(s) where your graphic files are
+  % \graphicspath{{../eps/}}
+  % and their extensions so you won't have to specify these with
+  % every instance of \includegraphics
+  % \DeclareGraphicsExtensions{.eps}
+\fi
+% graphicx was written by David Carlisle and Sebastian Rahtz. It is
+% required if you want graphics, photos, etc. graphicx.sty is already
+% installed on most LaTeX systems. The latest version and documentation can
+% be obtained at: 
+% http://www.ctan.org/tex-archive/macros/latex/required/graphics/
+% Another good source of documentation is "Using Imported Graphics in
+% LaTeX2e" by Keith Reckdahl which can be found as epslatex.ps or
+% epslatex.pdf at: http://www.ctan.org/tex-archive/info/
+%
+% latex, and pdflatex in dvi mode, support graphics in encapsulated
+% postscript (.eps) format. pdflatex in pdf mode supports graphics
+% in .pdf, .jpeg, .png and .mps (metapost) formats. Users should ensure
+% that all non-photo figures use a vector format (.eps, .pdf, .mps) and
+% not a bitmapped formats (.jpeg, .png). IEEE frowns on bitmapped formats
+% which can result in "jaggedy"/blurry rendering of lines and letters as
+% well as large increases in file sizes.
+%
+% You can find documentation about the pdfTeX application at:
+% http://www.tug.org/applications/pdftex
 
-% The \icmltitle you define below is probably too long as a header.
-% Therefore, a short form for the running title is supplied here:
-\icmltitlerunning{Leveraging Sparsity to Speed Up Polynomial Feature Expansions of CSR Matrices Using $K$-Simplex Numbers}
+% *** MATH PACKAGES ***
+%
+%\usepackage[cmex10]{amsmath}
+% A popular package from the American Mathematical Society that provides
+% many useful and powerful commands for dealing with mathematics. If using
+% it, be sure to load this package with the cmex10 option to ensure that
+% only type 1 fonts will utilized at all point sizes. Without this option,
+% it is possible that some math symbols, particularly those within
+% footnotes, will be rendered in bitmap form which will result in a
+% document that can not be IEEE Xplore compliant!
+%
+% Also, note that the amsmath package sets \interdisplaylinepenalty to 10000
+% thus preventing page breaks from occurring within multiline equations. Use:
+%\interdisplaylinepenalty=2500
+% after loading amsmath to restore such page breaks as IEEEtran.cls normally
+% does. amsmath.sty is already installed on most LaTeX systems. The latest
+% version and documentation can be obtained at:
+% http://www.ctan.org/tex-archive/macros/latex/required/amslatex/math/
+
+% *** SPECIALIZED LIST PACKAGES ***
+%
+%\usepackage{algorithmic}
+% algorithmic.sty was written by Peter Williams and Rogerio Brito.
+% This package provides an algorithmic environment fo describing algorithms.
+% You can use the algorithmic environment in-text or within a figure
+% environment to provide for a floating algorithm. Do NOT use the algorithm
+% floating environment provided by algorithm.sty (by the same authors) or
+% algorithm2e.sty (by Christophe Fiorio) as IEEE does not use dedicated
+% algorithm float types and packages that provide these will not provide
+% correct IEEE style captions. The latest version and documentation of
+% algorithmic.sty can be obtained at:
+% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithms/
+% There is also a support site at:
+% http://algorithms.berlios.de/index.html
+% Also of interest may be the (relatively newer and more customizable)
+% algorithmicx.sty package by Szasz Janos:
+% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithmicx/
+
+% *** ALIGNMENT PACKAGES ***
+%
+%\usepackage{array}
+% Frank Mittelbach's and David Carlisle's array.sty patches and improves
+% the standard LaTeX2e array and tabular environments to provide better
+% appearance and additional user controls. As the default LaTeX2e table
+% generation code is lacking to the point of almost being broken with
+% respect to the quality of the end results, all users are strongly
+% advised to use an enhanced (at the very least that provided by array.sty)
+% set of table tools. array.sty is already installed on most systems. The
+% latest version and documentation can be obtained at:
+% http://www.ctan.org/tex-archive/macros/latex/required/tools/
+
+%\usepackage{mdwmath}
+%\usepackage{mdwtab}
+% Also highly recommended is Mark Wooding's extremely powerful MDW tools,
+% especially mdwmath.sty and mdwtab.sty which are used to format equations
+% and tables, respectively. The MDWtools set is already installed on most
+% LaTeX systems. The lastest version and documentation is available at:
+% http://www.ctan.org/tex-archive/macros/latex/contrib/mdwtools/
+
+% IEEEtran contains the IEEEeqnarray family of commands that can be used to
+% generate multiline equations as well as matrices, tables, etc., of high
+% quality.
+
+%\usepackage{eqparbox}
+% Also of notable interest is Scott Pakin's eqparbox package for creating
+% (automatically sized) equal width boxes - aka "natural width parboxes".
+% Available at:
+% http://www.ctan.org/tex-archive/macros/latex/contrib/eqparbox/
+
+% *** SUBFIGURE PACKAGES ***
+%\usepackage[tight,footnotesize]{subfigure}
+% subfigure.sty was written by Steven Douglas Cochran. This package makes it
+% easy to put subfigures in your figures. e.g., "Figure 1a and 1b". For IEEE
+% work, it is a good idea to load it with the tight package option to reduce
+% the amount of white space around the subfigures. subfigure.sty is already
+% installed on most LaTeX systems. The latest version and documentation can
+% be obtained at:
+% http://www.ctan.org/tex-archive/obsolete/macros/latex/contrib/subfigure/
+% subfigure.sty has been superceeded by subfig.sty.
+
+%\usepackage[caption=false]{caption}
+%\usepackage[font=footnotesize]{subfig}
+% subfig.sty, also written by Steven Douglas Cochran, is the modern
+% replacement for subfigure.sty. However, subfig.sty requires and
+% automatically loads Axel Sommerfeldt's caption.sty which will override
+% IEEEtran.cls handling of captions and this will result in nonIEEE style
+% figure/table captions. To prevent this problem, be sure and preload
+% caption.sty with its "caption=false" package option. This is will preserve
+% IEEEtran.cls handing of captions. Version 1.3 (2005/06/28) and later 
+% (recommended due to many improvements over 1.2) of subfig.sty supports
+% the caption=false option directly:
+%\usepackage[caption=false,font=footnotesize]{subfig}
+%
+% The latest version and documentation can be obtained at:
+% http://www.ctan.org/tex-archive/macros/latex/contrib/subfig/
+% The latest version and documentation of caption.sty can be obtained at:
+% http://www.ctan.org/tex-archive/macros/latex/contrib/caption/
+
+% *** FLOAT PACKAGES ***
+%
+%\usepackage{fixltx2e}
+% fixltx2e, the successor to the earlier fix2col.sty, was written by
+% Frank Mittelbach and David Carlisle. This package corrects a few problems
+% in the LaTeX2e kernel, the most notable of which is that in current
+% LaTeX2e releases, the ordering of single and double column floats is not
+% guaranteed to be preserved. Thus, an unpatched LaTeX2e can allow a
+% single column figure to be placed prior to an earlier double column
+% figure. The latest version and documentation can be found at:
+% http://www.ctan.org/tex-archive/macros/latex/base/
+
+%\usepackage{stfloats}
+% stfloats.sty was written by Sigitas Tolusis. This package gives LaTeX2e
+% the ability to do double column floats at the bottom of the page as well
+% as the top. (e.g., "\begin{figure*}[!b]" is not normally possible in
+% LaTeX2e). It also provides a command:
+%\fnbelowfloat
+% to enable the placement of footnotes below bottom floats (the standard
+% LaTeX2e kernel puts them above bottom floats). This is an invasive package
+% which rewrites many portions of the LaTeX2e float routines. It may not work
+% with other packages that modify the LaTeX2e float routines. The latest
+% version and documentation can be obtained at:
+% http://www.ctan.org/tex-archive/macros/latex/contrib/sttools/
+% Documentation is contained in the stfloats.sty comments as well as in the
+% presfull.pdf file. Do not use the stfloats baselinefloat ability as IEEE
+% does not allow \baselineskip to stretch. Authors submitting work to the
+% IEEE should note that IEEE rarely uses double column equations and
+% that authors should try to avoid such use. Do not be tempted to use the
+% cuted.sty or midfloat.sty packages (also by Sigitas Tolusis) as IEEE does
+% not format its papers in such ways.
+
+%\ifCLASSOPTIONcaptionsoff
+%  \usepackage[nomarkers]{endfloat}
+% \let\MYoriglatexcaption\caption
+% \renewcommand{\caption}[2][\relax]{\MYoriglatexcaption[#2]{#2}}
+%\fi
+% endfloat.sty was written by James Darrell McCauley and Jeff Goldberg.
+% This package may be useful when used in conjunction with IEEEtran.cls'
+% captionsoff option. Some IEEE journals/societies require that submissions
+% have lists of figures/tables at the end of the paper and that
+% figures/tables without any captions are placed on a page by themselves at
+% the end of the document. If needed, the draftcls IEEEtran class option or
+% \CLASSINPUTbaselinestretch interface can be used to increase the line
+% spacing as well. Be sure and use the nomarkers option of endfloat to
+% prevent endfloat from "marking" where the figures would have been placed
+% in the text. The two hack lines of code above are a slight modification of
+% that suggested by in the endfloat docs (section 8.3.1) to ensure that
+% the full captions always appear in the list of figures/tables - even if
+% the user used the short optional argument of \caption[]{}.
+% IEEE papers do not typically make use of \caption[]'s optional argument,
+% so this should not be an issue. A similar trick can be used to disable
+% captions of packages such as subfig.sty that lack options to turn off
+% the subcaptions:
+% For subfig.sty:
+% \let\MYorigsubfloat\subfloat
+% \renewcommand{\subfloat}[2][\relax]{\MYorigsubfloat[]{#2}}
+% For subfigure.sty:
+% \let\MYorigsubfigure\subfigure
+% \renewcommand{\subfigure}[2][\relax]{\MYorigsubfigure[]{#2}}
+% However, the above trick will not work if both optional arguments of
+% the \subfloat/subfig command are used. Furthermore, there needs to be a
+% description of each subfigure *somewhere* and endfloat does not add
+% subfigure captions to its list of figures. Thus, the best approach is to
+% avoid the use of subfigure captions (many IEEE journals avoid them anyway)
+% and instead reference/explain all the subfigures within the main caption.
+% The latest version of endfloat.sty and its documentation can obtained at:
+% http://www.ctan.org/tex-archive/macros/latex/contrib/endfloat/
+%
+% The IEEEtran \ifCLASSOPTIONcaptionsoff conditional can also be used
+% later in the document, say, to conditionally put the References on a 
+% page by themselves.
+
+% *** PDF, URL AND HYPERLINK PACKAGES ***
+%
+%\usepackage{url}
+% url.sty was written by Donald Arseneau. It provides better support for
+% handling and breaking URLs. url.sty is already installed on most LaTeX
+% systems. The latest version can be obtained at:
+% http://www.ctan.org/tex-archive/macros/latex/contrib/misc/
+% Read the url.sty source comments for usage information. Basically,
+% \url{my_url_here}.
+
+% *** Do not adjust lengths that control margins, column widths, etc. ***
+% *** Do not use packages that alter fonts (such as pslatex).         ***
+% There should be no need to do such things with IEEEtran.cls V1.6 and later.
+% (Unless specifically asked to do so by the journal or conference you plan
+% to submit to, of course. )
+
+% correct bad hyphenation here
+\hyphenation{op-tical net-works semi-conduc-tor}
+
+\pagestyle{empty}
 
 \begin{document}
+%
+% paper title
+% can use linebreaks \\ within to get better formatting as desired
+\title{Leveraging Sparsity to Speed Up Polynomial Feature Expansions of CSR Matrices Using $K$-Simplex Numbers}
+%
+%
+% author names and IEEE memberships
+% note positions of commas and nonbreaking spaces ( ~ ) LaTeX will not break
+% a structure at a ~ so this keeps an author's name from being broken across
+% two lines.
+% use \thanks{} to gain access to the first footnote area
+% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
+% was not built to handle multiple paragraphs
+%
 
-\twocolumn[
-\icmltitle{Leveraging Sparsity to Speed Up Polynomial Feature Expansions of CSR Matrices Using $K$-Simplex Numbers}
+%\author{Michael~Shell,
+%        John~Doe,~\IEEEmembership{Fellow,~OSA,}
+%        and~Jane~Doe,~\IEEEmembership{Life~Fellow,~IEEE}% <-this % stops a space
+%\thanks{M. Shell is with the Department
+%of Electrical and Computer Engineering, Georgia Institute of Technology, Atlanta,
+%GA, 30332 USA e-mail: (see http://www.michaelshell.org/contact.html).}% <-this % stops a space
+%\thanks{J. Doe and J. Doe are with Anonymous University.}% <-this % stops a space
+%\thanks{Manuscript received April 19, 2005; revised January 11, 2007.}}
 
-% It is OKAY to include author information, even for blind
-% submissions: the style file will automatically remove it for you
-% unless you've provided the [accepted] option to the icml2017
-% package.
+% note the % following the last \IEEEmembership and also \thanks - 
+% these prevent an unwanted space from occurring between the last author name
+% and the end of the author line. i.e., if you had this:
+% 
+% \author{....lastname \thanks{...} \thanks{...} }
+%                     ^------------^------------^----Do not want these spaces!
+%
+% a space would be appended to the last name and could cause every name on that
+% line to be shifted left slightly. This is one of those "LaTeX things". For
+% instance, "\textbf{A} \textbf{B}" will typeset as "A B" not "AB". To get
+% "AB" then you have to do: "\textbf{A}\textbf{B}"
+% \thanks is no different in this regard, so shield the last } of each \thanks
+% that ends a line with a % and do not let a space in before the next \thanks.
+% Spaces after \IEEEmembership other than the last one are OK (and needed) as
+% you are supposed to have spaces between the names. For what it is worth,
+% this is a minor point as most people would not even notice if the said evil
+% space somehow managed to creep in.
 
-% list of affiliations. the first argument should be a (short)
-% identifier you will use later to specify author affiliations
-% Academic affiliations should list Department, University, City, Region, Country
-% Industry affiliations should list Company, City, Region, Country
+% The paper headers
+\markboth{Journal of \LaTeX\ Class Files,~Vol.~6, No.~1, January~2007}%
+{Shell \MakeLowercase{\textit{et al.}}: Bare Demo of IEEEtran.cls for Journals}
+% The only time the second header will appear is for the odd numbered pages
+% after the title page when using the twoside option.
+% 
+% *** Note that you probably will NOT want to include the author's ***
+% *** name in the headers of peer review papers.                   ***
+% You can use \ifCLASSOPTIONpeerreview for conditional compilation here if
+% you desire.
 
-% you can specify symbols, otherwise they are numbered in order
-% ideally, you should not use this facility. affiliations will be numbered
-% in order of appearance and this is the preferred way.
-\icmlsetsymbol{equal}{*}
+% If you want to put a publisher's ID mark on the page you can do it like
+% this:
+%\IEEEpubid{0000--0000/00\$00.00~\copyright~2007 IEEE}
+% Remember, if you use this you must call \IEEEpubidadjcol in the second
+% column for its text to clear the IEEEpubid mark.
 
-\begin{icmlauthorlist}
-\icmlauthor{Andrew Nystrom}{google}
-\icmlauthor{John Hughes}{brown}
-\end{icmlauthorlist}
+% use for special paper notices
+%\IEEEspecialpapernotice{(Invited Paper)}
 
-\icmlaffiliation{google}{Google Inc.}
-\icmlaffiliation{brown}{Brown University}
-
-\icmlcorrespondingauthor{Andrew Nystrom}{nystrom@google.com}
-\icmlcorrespondingauthor{John Hughes}{jfh@cs.brown.edu}
-
-% You may provide any keywords that you 
-% find helpful for describing your paper; these are used to populate 
-% the "keywords" metadata in the PDF but will not be shown in the document
-\icmlkeywords{csr, polynomial, sparse, expansion, triangle, tetrahedral, simplex}
-
-\vskip 0.3in
-]
-
-% this must go after the closing bracket ] following \twocolumn[ ...
-
-% This command actually creates the footnote in the first column
-% listing the affiliations and the copyright notice.
-% The command takes one argument, which is text to display at the start of the footnote.
-% The \icmlEqualContribution command is standard text for equal contribution.
-% Remove it (just {}) if you do not need this facility.
-
-%\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
-%\printAffiliationsAndNotice{\icmlEqualContribution} % otherwise use the standard text.
-%\footnotetext{hi}
+\maketitle
+\thispagestyle{empty}
 
 \begin{abstract}
-We provide an algorithm that speeds up polynomial and interaction feature expansion on sparse matrices.
-The algorithm operates on and produces a compressed sparse row matrix with no densification.
-It works by defining a bijective function involving simplex numbers of column indices in the original matrix to column indices in the expanded matrix.
-This function allows for only the nonzero elements to be iterated over and multiplied together, greatly improving the expansion of sparse matrices in compressed sparse row format.
+%\boldmath
+An algorithm is provided for performing polynomial feature expansions whose input and output are compressed sparse row (CSR) matrices.
+Previously, no such algorithm existed, and performing these expansions on CSR matrices required an intermediate densification step.
+%The algorithm operates on and produces a compressed sparse row matrix with no densification.
+The algorithm works by using a bijective function involving simplex numbers of column indices in the original matrix to column indices in the expanded matrix.
+Not only is space saved by working only in CSR format, but the bijective function allows for only the nonzero elements to be iterated over and multiplied together during the expansion, greatly improving average time complexity.
 For a vector of dimension $D$ and density $0 \le d \le 1$, the algorithm has time complexity $\Theta(d^KD^K)$ where $K$ is the polynomial-feature order; this  is an improvement by a factor $d^K$ over the standard method.
-[Keywords: compressed sparse row, csr, feature expansion, feature mapping, polynomial expansion, sparse matrix]
 \end{abstract}
+% IEEEtran.cls defaults to using nonbold math in the Abstract.
+% This preserves the distinction between vectors and scalars. However,
+% if the journal you are submitting to favors bold math in the abstract,
+% then you can use LaTeX's standard command \boldmath at the very start
+% of the abstract to achieve this. Many IEEE journals frown on math
+% in the abstract anyway.
+
+% Note that keywords are not normally used for peerreview papers.
+\begin{IEEEkeywords}
+compressed sparse row, csr, feature expansion, feature mapping, polynomial feature expansion, simplex numbers, sparse matrix, tetrahedral numbers, triangle numbers
+\end{IEEEkeywords}
+
+% For peer review papers, you can put extra information on the cover
+% page as needed:
+% \ifCLASSOPTIONpeerreview
+% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
+% \fi
+%
+% For peerreview papers, this IEEEtran command inserts a page break and
+% creates the second title. It will be ignored for other modes.
+\IEEEpeerreviewmaketitle
 
@@
 A well known and widely used feature mapping is the \emph{polynomial expansion}, which produces a new feature for each degree-$k$ monomial in the original features.
-(If the original features are $x,y,z$, the order-2 polynomial features are $x^2, y^2, z^2, xy, xz, xy$, and the order-3 ones are $x^3, y^3, z^3, x^2y, x^2z, xy^2, y^2 z, xz^2, yz^2,$ and $xyz$.
+If the original features are $x,y,z$, the order-2 polynomial features are $x^2, y^2, z^2, xy, xz, xy, yz$, and the order-3 polynomial features are $x^3, y^3, z^3, x^2y, x^2z, xy^2, y^2 z, xz^2, yz^2,$ and $xyz$.
 A $K$-order polynomial feature expansion of the feature space allows a linear model to learn polynomial relationships between dependent and independent variables.
@@
 \begin{algorithm}%[H]
-   \caption{Sparse Second Order Polynomial Expansion}
+   \caption{Incomplete Sparse Second Order Polynomial Expansion}
    \label{alg:Incomplete-Sparse-Second-Order-Polynomial-Expansion}
@@
 For reference, we list the first five triangle and tetrahedral numbers in the following table:
-
+\begin{table}
+  \centering
+  \caption{The first five triangle ($T_2$) and tetrahedral ($T_3$) numbers.}
   \begin{tabular}{| c | c | c |}
-  \caption{The first five triangle and tetrahedral numbers.}
     \hline
@@
   \end{tabular}
+\end{table}
 
@@
 \end{bmatrix}
-\label{eq:4x4mat}
+\label{eq:4by4mat}
 \end{align}
@@
 \end{bmatrix}
-\label{eq:preliminary4x4}
+\label{eq:preliminary4by4}
 \end{align}
 and then subtract the preliminary mapping from the total number of elements in the codomain to create the final mapping.
-Note that in equation \ref{eq:preliminary4x4} we have the following relation:
+Note that in equation \ref{eq:preliminary4by4} we have the following relation:
 
@@
 
-where $e^i$ is the value of any cell in row $i$ of equation \ref{eq:preliminary4x4}.
+where $e^i$ is the value of any cell in row $i$ of equation \ref{eq:preliminary4by4}.
 
-Therefore, the following mapping will produce equation \ref{eq:preliminary4x4}:
+Therefore, the following mapping will produce equation \ref{eq:preliminary4by4}:
 
@@
 
-We can now use this result to construct a mapping for equation \ref{4x4mat} by subtracting it from the size of the codomain:
+We can now use this result to construct a mapping for equation \ref{4by4mat} by subtracting it from the size of the codomain:
 
@@
 
+\begin{table}
+  \caption{Form of the required mapping for $K=3$, $D=5$.}
   \begin{tabular}{| c | c | c |}
@@
 
+% The very first letter is a 2 line initial drop letter followed
+% by the rest of the first word in caps.
+% 
+% form to use if the first word consists of a single letter:
+% \IEEEPARstart{A}{demo} file is ....
+% 
+% form to use if you need the single drop letter followed by
+% normal text (unknown if ever used by IEEE):
+% \IEEEPARstart{A}{}demo file is ....
+% 
+% Some journals put the first two words in caps:
+% \IEEEPARstart{T}{his demo} file is ....
+% 
+% Here we have the typical use of a "T" for an initial drop letter
+% and "HIS" in caps to complete the first word.
+
+%\IEEEPARstart{T}{his} demo file is intended to serve as a ``starter file''
+%for IEEE journal papers produced under \LaTeX\ using
+%IEEEtran.cls version 1.7 and later.
+% You must have at least 2 lines in the paragraph with the drop letter
+% (should never be an issue)
+%I wish you the best of success.
+
+%\hfill mds
+
+%\hfill January 11, 2007
+
+%\subsection{Subsection Heading Here}
+%Subsection text here.
+
+% needed in second column of first page if using \IEEEpubid
+%\IEEEpubidadjcol
+
+%\vskip 1.5in
+
+% An example of a floating figure using the graphicx package.
+% Note that \label must occur AFTER (or within) \caption.
+% For figures, \caption should occur after the \includegraphics.
+% Note that IEEEtran v1.7 and later has special internal code that
+% is designed to preserve the operation of \label within \caption
+% even when the captionsoff option is in effect. However, because
+% of issues like this, it may be the safest practice to put all your
+% \label just after \caption rather than within \caption{}.
+%
+% Reminder: the "draftcls" or "draftclsnofoot", not "draft", class
+% option should be used if it is desired that the figures are to be
+% displayed while in draft mode.
+%
+%\begin{figure}[!t]
+%\centering
+%\includegraphics[width=2.5in]{myfigure}
+% where an .eps filename suffix will be assumed under latex, 
+% and a .pdf suffix will be assumed for pdflatex; or what has been declared
+% via \DeclareGraphicsExtensions.
+%\caption{Simulation Results}
+%\label{fig_sim}
+%\end{figure}
+
+% Note that IEEE typically puts floats only at the top, even when this
+% results in a large percentage of a column being occupied by floats.
+
+% An example of a double column floating figure using two subfigures.
+% (The subfig.sty package must be loaded for this to work.)
+% The subfigure \label commands are set within each subfloat command, the
+% \label for the overall figure must come after \caption.
+% \hfil must be used as a separator to get equal spacing.
+% The subfigure.sty package works much the same way, except \subfigure is
+% used instead of \subfloat.
+%
+%\begin{figure*}[!t]
+%\centerline{\subfloat[Case I]\includegraphics[width=2.5in]{subfigcase1}%
+%\label{fig_first_case}}
+%\hfil
+%\subfloat[Case II]{\includegraphics[width=2.5in]{subfigcase2}%
+%\label{fig_second_case}}}
+%\caption{Simulation results}
+%\label{fig_sim}
+%\end{figure*}
+%
+% Note that often IEEE papers with subfigures do not employ subfigure
+% captions (using the optional argument to \subfloat), but instead will
+% reference/describe all of them (a), (b), etc., within the main caption.
+
+% An example of a floating table. Note that, for IEEE style tables, the 
+% \caption command should come BEFORE the table. Table text will default to
+% \footnotesize as IEEE normally uses this smaller font for tables.
+% The \label must come after \caption as always.
+%
+%\begin{table}[!t]
+%% increase table row spacing, adjust to taste
+%\renewcommand{\arraystretch}{1.3}
+% if using array.sty, it might be a good idea to tweak the value of
+% \extrarowheight as needed to properly center the text within the cells
+%\caption{An Example of a Table}
+%\label{table_example}
+%\centering
+%% Some packages, such as MDW tools, offer better commands for making tables
+%% than the plain LaTeX2e tabular which is used here.
+%\begin{tabular}{|c||c|}
+%\hline
+%One & Two\\
+%\hline
+%Three & Four\\
+%\hline
+%\end{tabular}
+%\end{table}
+
+% Note that IEEE does not put floats in the very first column - or typically
+% anywhere on the first page for that matter. Also, in-text middle ("here")
+% positioning is not used. Most IEEE journals use top floats exclusively.
+% Note that, LaTeX2e, unlike IEEE journals, places footnotes above bottom
+% floats. This can be corrected via the \fnbelowfloat command of the
+% stfloats package.
+
+% if have a single appendix:
+%\appendix[Proof of the Zonklar Equations]
+% or
+%\appendix  % for no appendix heading
+% do not use \section anymore after \appendix, only \section*
+% is possibly needed
+
+% use appendices with more than one appendix
+% then use \section to start each appendix
+% you must declare a \section before using any
+% \subsection or using \label (\appendices by itself
+% starts a section numbered zero.)
+%
+
+%\appendices
+%\section{Proof of the First Zonklar Equation}
+%Appendix one text goes here.
+
+% you can choose not to have a title for an appendix
+% if you want by leaving the argument blank
+%\section{}
+%Appendix two text goes here.
+
+% use section* for acknowledgement
+%\section*{Acknowledgment}
+
+%The authors would like to thank...
+
+% Can use something like this to put references on a page
+% by themselves when using endfloat and the captionsoff option.
+\ifCLASSOPTIONcaptionsoff
   \newpage
+\fi
 
-\bibliography{nystrom_hughes_csr_polynomial_expansions}
-\bibliographystyle{icml2017}
+% trigger a \newpage just before the given reference
+% number - used to balance the columns on the last page
+% adjust value as needed - may need to be readjusted if
+% the document is modified later
+%\IEEEtriggeratref{8}
+% The "triggered" command can be changed if desired:
+%\IEEEtriggercmd{\enlargethispage{-5in}}
 
+% references section
+
+% can use a bibliography generated by BibTeX as a .bbl file
+% BibTeX documentation can be easily obtained at:
+% http://www.ctan.org/tex-archive/biblio/bibtex/contrib/doc/
+% The IEEEtran BibTeX style support page is at:
+% http://www.michaelshell.org/tex/ieeetran/bibtex/
+\bibliographystyle{IEEEtran}
+% argument is your BibTeX string definitions and bibliography database(s)
+%\bibliography{IEEEabrv,../bib/paper}
+%
+% <OR> manually copy in the resultant .bbl file
+% set second argument of \begin to the number of references
+% (used to reserve space for the reference number labels box)
+\bibliography{iccsda_sparse_poly}
+%\bibliographystyle{iccsda_sparse_poly}
+
+% biography section
+% 
+% If you have an EPS/PDF photo (graphicx package needed) extra braces are
+% needed around the contents of the optional argument to biography to prevent
+% the LaTeX parser from getting confused when it sees the complicated
+% \includegraphics command within an optional argument. (You could create
+% your own custom macro containing the \includegraphics command to make things
+% simpler here.)
+%\begin{biography}[{\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{mshell}}]{Michael Shell}
+% or if you just want to reserve a space for a photo:
+
+%\begin{IEEEbiography}{Michael Shell}
+%Biography text here.
+%\end{IEEEbiography}
+
+% if you will not have a photo at all:
+%\begin{IEEEbiographynophoto}{John Doe}
+%Biography text here.
+%\end{IEEEbiographynophoto}
+
+% insert where needed to balance the two columns on the last page with
+% biographies
+%\newpage
+
+%\begin{IEEEbiographynophoto}{Andrew Nystrom}
+%Andrew is a Software Engineer in Google Inc's Research \& Machine Intelligence department.
+%\end{IEEEbiographynophoto}
+
+% You can push biographies down or up by placing
+% a \vfill before or after them. The appropriate
+% use of \vfill depends on what kind of text is
+% on the last page and whether or not the columns
+% are being equalized.
+
+%\vfill
+
+% Can be used to pull up biographies so that the bottom of the last one
+% is flush with the other column.
+%\enlargethispage{-5in}
+
+% that's all folks
 \end{document}
-
-% This document was modified from the file originally made available by
-% Pat Langley and Andrea Danyluk for ICML-2K. This version was
-% created by Lise Getoor and Tobias Scheffer, it was slightly modified  
-% from the 2010 version by Thorsten Joachims & Johannes Fuernkranz, 
-% slightly modified from the 2009 version by Kiri Wagstaff and 
-% Sam Roweis's 2008 version, which is slightly modified from 
-% Prasad Tadepalli's 2007 version which is a lightly 
-% changed version of the previous year's version by Andrew Moore, 
-% which was in turn edited from those of Kristian Kersting and 
-% Codrina Lauth. Alex Smola contributed to the algorithmic style files.

### 2018-04-11 paper/Final/iccsda_sparse_poly.tex
abstract tweaks

@@
 %\boldmath
-An algorithm is provided for performing polynomial feature expansions whose input and output are compressed sparse row (CSR) matrices.
+An algorithm is provided for performing polynomial feature expansions that both operates on and produces compressed sparse row (CSR) matrices.
 Previously, no such algorithm existed, and performing these expansions on CSR matrices required an intermediate densification step.
 %The algorithm operates on and produces a compressed sparse row matrix with no densification.
-The algorithm works by using a bijective function involving simplex numbers of column indices in the original matrix to column indices in the expanded matrix.
+The algorithm performs a $K$-degree expansion by using a bijective function involving $K$-simplex numbers of column indices in the original matrix to column indices in the expanded matrix.
 Not only is space saved by working only in CSR format, but the bijective function allows for only the nonzero elements to be iterated over and multiplied together during the expansion, greatly improving average time complexity.

### 2018-09-10 paper/Final/sparse_poly_proper_order.tex
a version not connected with any attempt to publish with using the canonical output ordering

--- 2018-04-11 paper/Final/iccsda_sparse_poly.tex
@@
 \usepackage{hyperref}
+
+\usepackage[T1]{fontenc}
+\usepackage[utf8]{inputenc}
+\usepackage{authblk}
 
@@
 
-%\author{Michael~Shell,
-%        John~Doe,~\IEEEmembership{Fellow,~OSA,}
-%        and~Jane~Doe,~\IEEEmembership{Life~Fellow,~IEEE}% <-this % stops a space
+\author[1]{Andrew Nystrom\thanks{nystrom@google.com}}
+\author[2]{John Hughes\thanks{jfh@cs.brown.edu}}
+\affil[1]{Google AI}
+\affil[2]{Brown University}
+
+%\renewcommand\Authands{ and }
+
+%\author{Andrew Nystrom, John Hughes \\
+%        nystrom@google.com, jfh@cs.brown.edu}% <-this % stops a space
 %\thanks{M. Shell is with the Department
@@
 An algorithm is provided for performing polynomial feature expansions that both operates on and produces compressed sparse row (CSR) matrices.
-Previously, no such algorithm existed, and performing these expansions on CSR matrices required an intermediate densification step.
+Previously, no such algorithm existed, and performing polynomial expansions on CSR matrices required an intermediate densification step.
 %The algorithm operates on and produces a compressed sparse row matrix with no densification.
 The algorithm performs a $K$-degree expansion by using a bijective function involving $K$-simplex numbers of column indices in the original matrix to column indices in the expanded matrix.
-Not only is space saved by working only in CSR format, but the bijective function allows for only the nonzero elements to be iterated over and multiplied together during the expansion, greatly improving average time complexity.
-For a vector of dimension $D$ and density $0 \le d \le 1$, the algorithm has time complexity $\Theta(d^KD^K)$ where $K$ is the polynomial-feature order; this  is an improvement by a factor $d^K$ over the standard method.
+Not only is space saved by operating in CSR format, but the bijective function allows for only the nonzero elements to be iterated over and multiplied together during the expansion, greatly improving average time complexity.
+For a vector of dimensionality $D$ and density $0 \le d \le 1$, the algorithm has average time complexity $\Theta(d^KD^K)$ where $K$ is the polynomial-feature order; this  is an improvement by a factor $d^K$ over the standard method.
+This work derives the required function for the cases of $K=2$ and $K=3$ and shows its use in the $K=2$ algorithm.
+%The becomes faster than the traditional dense algorithm even when the CSR matrix is not sparse ($d > \frac{1}{2}$) as the dimensionality of the matrix increases.
 \end{abstract}
@@
 A well known and widely used feature mapping is the \emph{polynomial expansion}, which produces a new feature for each degree-$k$ monomial in the original features.
-If the original features are $x,y,z$, the order-2 polynomial features are $x^2, y^2, z^2, xy, xz, xy, yz$, and the order-3 polynomial features are $x^3, y^3, z^3, x^2y, x^2z, xy^2, y^2 z, xz^2, yz^2,$ and $xyz$.
+If the original features are $x,y,z$, the order-2 polynomial features are $x^2, y^2, z^2, xy, xz, yz$, and the order-3 polynomial features are $x^3, y^3, z^3, x^2y, x^2z, xy^2, y^2z, xz^2, yz^2, xyz$.
 A $K$-order polynomial feature expansion of the feature space allows a linear model to learn polynomial relationships between dependent and independent variables.
@@
 \end{bmatrix}
-\label{eq:4by4mat}
+\label{eq:4by4mat_inter}
 \end{align}
@@
 
-We can now use this result to construct a mapping for equation \ref{4by4mat} by subtracting it from the size of the codomain:
+We can now use this result to construct the desired mapping by subtracting it from the size of the codomain:
 
@@
 \end{bmatrix}
-\label{eq:4by4mat}
+\label{eq:4by4mat_poly}
 \end{align}
@@
 % (used to reserve space for the reference number labels box)
-\bibliography{iccsda_sparse_poly}
-%\bibliographystyle{iccsda_sparse_poly}
+\bibliography{sparse_poly_proper_order}
+%\bibliographystyle{sparse_poly_proper_order}
 

### 2018-09-10 paper/Final/build.sh

@@ -1,0 +1,6 @@
+rm sparse_poly_proper_order.pdf
+pdflatex sparse_poly_proper_order
+bibtex sparse_poly_proper_order
+pdflatex sparse_poly_proper_order
+pdflatex sparse_poly_proper_order
+open sparse_poly_proper_order.pdf
